---
title: "Genetic Simulation of a Maize Factorial Cross"
subtitle: "Generating Yield Data under Normal and Low Phosphorus Conditions"
author: "Antigravity Genetics & Breeding Lab (NCSU)"
date: "2026-05-30"
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: true
    code-fold: show
    code-tools: true
---

> [!NOTE]
> This notebook is dedicated strictly to the genetic simulation of a 7 × 7 maize factorial cross (Line × Tester) evaluated in two environments. The objective is to establish a biologically realistic dataset for General Combining Ability (GCA) and Specific Combining Ability (SCA) calculations, and save it in a clean CSV format.

# Mating and Environmental Design

Our simulation is modeled exactly on Stelio's PhD experimental setup:

*   **Breeding Scheme:** Factorial Mating Design (North Carolina Design II / Line × Tester).

    *   **Females (Lines):** 7 Low-Phosphorus tolerant CIMMYT inbred lines (`CML364`, `CML366`, `CML434`, `CML435`, `CML439`, `CML440`, `CML532`).

    *   **Males (Testers):** 7 elite local Mozambican lines (`MOZ1` to `MOZ7`).

    *   **Crosses:** All $7 \times 7 = 49$ unique $F_1$ hybrid combinations. (No selfs, no reciprocals).

*   **Experimental Design:** Randomized Complete Block Design (RCBD) with **3 replications (blocks)** per environment.

*   **Environments ($E$):** 2 environments representing phosphorus treatments (`env1` = Normal Phosphorus, `env2` = Low Phosphorus Stress).

*   **Response Trait:** Grain Yield (tons/ha).



# R Setup and Simulation Parameters

We load the `tidyverse` package for data generation and configure the true biological parameters of our breeding population.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)

# Set seed for reproducibility
set.seed(88)
```

## Parameter Definitions
We define the genetic and environmental parameters that will determine the phenotypic yield:
1.  **Base Mean ($\mu$):** 8.2 tons/ha.

2.  **Environment Effects ($E$):** Normal P (`env1`) adds $+1.1$ t/ha; Low P (`env2`) reduces yield by $-1.3$ t/ha.

3.  **Female GCA Effects ($g_f$):** CIMMYT lines carry alleles for phosphorus use efficiency (PUE).

4.  **Male GCA Effects ($g_m$):** Local lines reflect general agroecological adaptation.

5.  **SCA Effects ($s_{fm}$):** Specific hybrid combinations show positive heterosis.

6.  **GCA × Environment Interaction ($G \times E$):** In the low-P environment (`env2`), the PUE tolerance of superior CIMMYT lines (`CML364` and `CML366`) is expressed even more strongly (enhancing their GCA by $+30\%$).

7.  **SCA × Environment Interaction ($S \times E$):** Hybrid-specific adaptation to environmental stress.

8.  **Residual Plot Error ($\epsilon$):** Normal variation among plots ($\sigma_{\epsilon} = 0.35$ t/ha).

```{r}
#| label: parameters

# Parental lines
females <- c("CML364", "CML366", "CML434", "CML435", "CML439", "CML440", "CML532")
males <- paste0("MOZ", 1:7)

# GCA effects
gca_female <- c(CML364=1.2, CML366=0.8, CML434=0.4, CML435=0.0, CML439=-0.3, CML440=-0.7, CML532=-1.1)
gca_male <- c(MOZ1=0.9, MOZ2=0.5, MOZ3=0.1, MOZ4=-0.2, MOZ5=-0.5, MOZ6=-0.8, MOZ7=-1.2)

# Specific Combining Ability (SCA) matrix
sca_matrix <- matrix(0, nrow=7, ncol=7, dimnames=list(females, males))
sca_matrix["CML364", "MOZ4"] <- 0.5   # Specific positive combining ability
sca_matrix["CML435", "MOZ1"] <- 0.4
sca_matrix["CML366", "MOZ6"] <- -0.4  # Specific negative incompatibility
```

***

# Dataset Generation

We combine these components to simulate grain yields for all $2 \text{ envs} \times 3 \text{ blocks} \times 49 \text{ crosses} = 294$ plot observations.

```{r}
#| label: data-generation

# 1. Create complete grid of observations
sim_data <- expand.grid(
  env = c("env1", "env2"),
  block = c("B1", "B2", "B3"),
  female = females,
  male = males,
  stringsAsFactors = FALSE
)

# 2. Add Cross ID
sim_data <- sim_data %>%
  mutate(cross = paste(female, male, sep = "_"))

# 3. Simulate Grain Yield (t/ha) based on our linear model
sim_data$yield <- apply(sim_data, 1, function(row) {
  env <- row["env"]
  blk <- row["block"]
  fem <- row["female"]
  mal <- row["male"]
  crs <- row["cross"]
  
  # Overall mean
  mu <- 8.2
  
  # Fixed environmental effect
  e_eff <- if(env == "env1") 1.1 else -1.3
  
  # Block nested in Environment (random block effect)
  set.seed(which(females == fem)*10 + which(males == mal) + which(c("env1", "env2") == env)*100 + which(c("B1", "B2", "B3") == blk)*1000)
  b_eff <- rnorm(1, mean = 0, sd = 0.18)
  
  # Parental General Combining Abilities
  g_fem <- gca_female[fem]
  g_mal <- gca_male[mal]
  
  # Specific Combining Ability
  s_val <- sca_matrix[fem, mal]
  
  # GCA x Environment Interaction
  # In env2 (low-P stress), the superior PUE donor GCAs are amplified by 30%
  ge_fem <- if(env == "env2") g_fem * 0.3 else -g_fem * 0.3
  ge_mal <- if(env == "env2") g_mal * 0.2 else -g_mal * 0.2
  
  # SCA x Environment Interaction (hybrid specific stress adaptation)
  se_val <- rnorm(1, mean = 0, sd = 0.10)
  
  # Residual Plot Error
  err <- rnorm(1, mean = 0, sd = 0.35)
  
  # Combine to get phenotypic yield
  y <- mu + e_eff + b_eff + g_fem + g_mal + s_val + ge_fem + ge_mal + se_val + err
  return(round(y, 3))
})
```

***

# Exporting Simulated Data

We save the completed, simulated dataset as a CSV file in the `/data` directory. This is the clean file that Stelio will import in his analysis notebooks.

```{r}
#| label: export-data

# Ensure data directory exists
dir.create("data", showWarnings = FALSE)

# Write to CSV
write_csv(sim_data, "data/maize_factorial_yield.csv")

message("Success! Dataset generated and saved to: data/maize_factorial_yield.csv")
```

## Data Preview
Let's print the first 10 rows of the exported CSV to verify its structure:

```{r}
#| label: data-preview
head(sim_data, 10)
```
