---
title: "Is the Male Tester Null Only for Yield, or Genome-Wide? A Trait-by-Trait Refit"
subtitle: "Refitting the line × tester model on every phenotype in `data/multilocation.csv` and inspecting `isSingular()` and the male (tester) effect per trait"
author: "Fausto Rodriguez Zapata"
date: "2026-06-22"
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: true
    code-fold: show
    code-tools: true
---

::: {.callout-tip appearance="simple"}
**Source code & data:** <https://github.com/faustovrz/gcamoz> · **Companion note:** [`male_gca_singularity_and_variance_routing.html`](male_gca_singularity_and_variance_routing.html)
:::

# The question this notebook settles

The yield model in [`multilocation_alpha_lattice.html`](multilocation_alpha_lattice.html)
gives a **zero male (tester) GCA** and `isSingular() == TRUE`. The companion note argues
this has a *germplasm* cause: the seven MOZL testers were **double-ascertained** (an
already-narrow CIMMYT/IIAM elite pool — CIMMYT ESA mean genetic distance ≈ 0.29, four
lines dominating 29–58 % of hybrids; Masuka et al. 2017 — then **truncation-selected on
yield** under phosphorus sufficiency), so they behave for combining-ability purposes as
*"essentially the same male."*

The open question (companion §3.3, §3.5) is **how far that "same male" claim reaches**:

| Mechanism | What it compresses | Male GCA zero for… |
|---|---|---|
| **A. Ascertainment / truncation on yield** | among-tester variance for **yield and yield-correlated traits** | yield (and its correlates) only |
| **B. Narrow base / near-identity-by-descent** | among-tester variance for **every trait** | **all** traits |

Your extrapolation — *"if the top seven are essentially the same line, the zero holds for
all phenotypes"* — requires **Mechanism B**. The decisive, falsifiable test is to **refit
the identical model on every measured trait** and read off, per trait:

1. `isSingular()` of the mixed model, and the REML **male GCA variance** (does it pin to 0?);
2. a boundary-corrected **REML likelihood-ratio test** of σ²_male — dropping `(1|male)`
   while keeping `(1|male:env)`, so the tester effect is tested against the correct
   male×environment stratum (do the testers differ at all for this trait?).

**Reading:** male null *only* for yield → **Mechanism A** (the zero is a yield-selection
artifact and does **not** generalize). Male null *across essentially all* traits →
**Mechanism B** (the testers really are interchangeable genome-wide, and the extrapolation
stands).

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(lme4)      # provides isSingular()
library(knitr)
library(kableExtra)
```

# Data, field layout, and derived flowering traits

We reuse the exact data-preparation pipeline of the yield notebook — same factor coding,
same recovery of field row/column from the 5 × 10 serpentine planting order — but we now
carry **all** plot-level phenotypes, not just yield. The `NAMULONGUE` check (`GEN = 0`) is
held out of the line × tester decomposition.

```{r}
#| label: read-and-prepare

raw_data <- read_csv("data/multilocation.csv", show_col_types = FALSE) %>%
  mutate(
    ENV    = factor(ENV),
    REP    = factor(REP, levels = c("I", "II", "III")),
    GEN    = factor(GEN, levels = as.character(0:49)),
    Female = factor(Female,
                    levels = c("Check",
                               "CML364", "CML366", "CML434", "CML435",
                               "CML439", "CML530", "CML532")),
    Male   = factor(Male,
                    levels = c("Check",
                               "MOZL3", "MOZL5", "MOZL6", "MOZL7",
                               "MOZL8", "MOZL9", "MOZL10"))
  )

# Field coordinates from the serpentine planting order (5 columns x 10 rows, snaking
# down along ENTRY) — identical to the yield notebook.
field <- raw_data %>%
  arrange(ENTRY) %>%
  group_by(ENV, REP) %>%
  mutate(
    plot_order = row_number(ENTRY),
    field_row  = ceiling(plot_order / 5),
    pos_in_row = ((plot_order - 1) %% 5) + 1,
    field_col  = if_else(field_row %% 2 == 1, pos_in_row, 6L - pos_in_row)
  ) %>%
  ungroup()
```

## Flowering dates → Julian day (no planting date available)

Stelio recorded **flowering dates**, not days-from-sowing, and there is **no planting
date** in the file. We therefore score flowering time as the **Julian day (day-of-year)**
of each event:

* **Days to silking** = Julian day of `F. D. Female` (female flowering / silk).
* **Days to anthesis** = Julian day of `F. D. Male` (male flowering / pollen shed).
* **ASI** (anthesis–silking interval) = silking − anthesis, a within-plot difference that
  is **independent of planting date** regardless.

Because `env` is a **fixed** effect in every model below, any site-level shift in planting
date is absorbed by the environment means — so the *among-tester* GCA/SCA estimates for the
Julian-day traits are unaffected by the missing sowing date (only the absolute intercept
would be). The raw dates come in two formats (`22/09` and `12-Oct`); we parse both.

```{r}
#| label: flowering-julian

fd_to_julian <- function(x) {
  x  <- trimws(as.character(x))
  dt <- parse_date_time(x, orders = c("d/m", "d-b", "dm", "db"),
                        locale = "C", quiet = TRUE)
  yday(dt)   # Julian day of year
}

field <- field %>%
  mutate(
    days_to_silking  = fd_to_julian(`F. D. Female`),
    days_to_anthesis = fd_to_julian(`F. D. Male`),
    ASI              = days_to_silking - days_to_anthesis
  )

# Confirm every flowering date parsed (no silent NAs introduced by format quirks).
parse_check <- tibble(
  column           = c("F. D. Female", "F. D. Male"),
  n_unparsed       = c(sum(is.na(field$days_to_silking)),
                       sum(is.na(field$days_to_anthesis)))
)
kable(parse_check, caption = "Flowering-date parse check (Julian day) — unparsed values should be 0") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Hybrids-only analysis frame and the trait dictionary

```{r}
#| label: hybrids-frame

hybrids <- field %>%
  filter(GEN != "0") %>%
  transmute(
    env    = droplevels(ENV),
    rep    = REP,
    row    = factor(field_row, levels = 1:10),
    col    = factor(field_col, levels = 1:5),
    female = droplevels(Female),
    male   = droplevels(Male),
    cross  = factor(`Female x Male`),
    # --- traits ---
    `Grain yield`        = `Yield t/ha`,
    `Yield per plant`    = `yield g/plant (g)`,
    `Plants per plot`    = `Plant Number`,
    `Ear weight`         = `Ear weight (g)`,
    `Grain weight`       = `grain weight (g)`,
    `Plant height`       = `Plant Height (m)`,
    `Culm diameter`      = `Culm Diameter (cm)`,
    `Days to silking`    = days_to_silking,
    `Days to anthesis`   = days_to_anthesis,
    `ASI`                = ASI,
    `Dry matter mass`    = `Dry matter mass (g)`,
    `Root weight`        = `Root Weight (g)`
  )

trait_units <- tribble(
  ~trait,              ~unit,
  "Grain yield",       "t/ha",
  "Yield per plant",   "g",
  "Plants per plot",   "count",
  "Ear weight",        "g/plot",
  "Grain weight",      "g/plot",
  "Plant height",      "m",
  "Culm diameter",     "cm",
  "Days to silking",   "Julian day",
  "Days to anthesis",  "Julian day",
  "ASI",               "days",
  "Dry matter mass",   "g",
  "Root weight",       "g"
)

kable(trait_units, caption = "Traits swept (all plot-level phenotypes in the file)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

traits <- trait_units$trait
```

# The per-trait engine

For each trait we fit the **same** random-effects structure used for yield, with
`lme4::lmer` (so that `isSingular()` is directly available — this is the `lme4`
counterpart of the `sommer::mmer` model in the yield notebook). The signal test for the
testers is a **boundary-corrected REML likelihood-ratio test** of the male GCA variance:
we drop `(1 | male)` while **keeping** `(1 | male:env)`, so the tester *main* effect is
tested above its environment interaction — the correct error stratum for a GCA in a
multi-environment trial. (A fixed-effects ANOVA F-test against plot error would be badly
anticonservative for high-G×E traits like phenology, manufacturing false "signals.")

We run the LRT on the full data **and** after dropping gross data errors
(|studentized residual| > 4 from a fixed-effects fit — see the data-quality screen). The
**robust** column decides the verdict. Trimming can cut both ways: it can erase a false
signal driven by a bad plot, *or* sharpen a real one that noise was masking.

```{r}
#| label: model-formulas

# Random terms listed explicitly, so one can be dropped cleanly for the LRT
# (avoids update() + its data-scoping pitfalls with lmer inside a function).
rand_terms <- c("(1 | female)", "(1 | male)", "(1 | cross)",
                "(1 | female:env)", "(1 | male:env)", "(1 | cross:env)",
                "(1 | env:rep)", "(1 | env:rep:row)", "(1 | env:rep:col)")

build_form <- function(drop = character()) {
  as.formula(paste("y ~ env +", paste(setdiff(rand_terms, drop), collapse = " + ")))
}
mix_form <- build_form()                                  # full mixed model

# Fixed-effects model — used ONLY to flag gross outliers via studentized
# residuals (NOT for the signal test).
fix_form <- y ~ env + env:rep + female + male + cross
```

```{r}
#| label: fit-one-trait

fit_mm <- function(form, dat) suppressMessages(suppressWarnings(
  lmer(form, data = dat, REML = TRUE,
       control = lmerControl(check.conv.singular = "ignore",
                             optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))))

# Boundary-corrected REML LRT dropping (1|term) while keeping every other term
# (so (1|male) is tested ABOVE (1|male:env), the right GCA stratum). Under H0 the
# statistic ~ 0.5·χ²₀ + 0.5·χ²₁, so the one-sided p halves the naive χ²₁ p.
# Reduced model is refit directly with explicit data — no update().
lrt_vc <- function(full_fit, dat, term) {
  red  <- fit_mm(build_form(drop = paste0("(1 | ", term, ")")), dat)
  stat <- max(0, 2 * (as.numeric(logLik(full_fit)) - as.numeric(logLik(red))))
  min(0.5 * pchisq(stat, df = 1, lower.tail = FALSE), 1)
}

fit_one <- function(trait_name, data) {
  d <- data %>% mutate(y = .data[[trait_name]]) %>% filter(!is.na(y))

  fit <- fit_mm(mix_form, d)
  vc  <- as.data.frame(VarCorr(fit))
  gv  <- function(g) { v <- vc$vcov[vc$grp == g]; if (length(v)) v else 0 }
  v_female <- gv("female"); v_male <- gv("male"); v_gen <- v_female + v_male + gv("cross")

  p_male_full <- lrt_vc(fit, d, "male")
  p_female    <- lrt_vc(fit, d, "female")

  # outlier-robust refit (drop |studentized resid| > 4 from the fixed-effects fit)
  rstud <- rstudent(lm(fix_form, data = d))
  keep  <- is.na(rstud) | abs(rstud) <= 4
  d_rob <- d[keep, ]
  p_male_robust <- lrt_vc(fit_mm(mix_form, d_rob), d_rob, "male")

  tibble(
    trait          = trait_name,
    singular       = isSingular(fit),
    sigma2_female  = v_female,
    sigma2_male    = v_male,
    male_share_pct = if (v_gen > 0) 100 * v_male / v_gen else NA_real_,
    n_outliers     = sum(!keep),
    p_male         = p_male_full,
    p_male_robust  = p_male_robust,
    p_female       = p_female
  )
}
```

# Run the sweep

```{r}
#| label: run-sweep
#| message: false
#| warning: false

results <- map_dfr(traits, fit_one, data = hybrids) %>%
  left_join(trait_units, by = "trait") %>%
  mutate(
    # verdict is decided on the OUTLIER-ROBUST male test, not the raw one
    male_verdict = if_else(!is.na(p_male_robust) & p_male_robust < 0.05,
                           "male SIGNAL", "male null"),
    raw_sig      = !is.na(p_male) & p_male < 0.05,
    rob_sig      = !is.na(p_male_robust) & p_male_robust < 0.05,
    flips        = raw_sig != rob_sig          # verdict changed after trimming (either way)
  ) %>%
  arrange(p_male_robust)

results_display <- results %>%
  transmute(
    Trait              = trait,
    Unit               = unit,
    `Singular?`        = if_else(singular, "yes", "no"),
    `σ²_female`        = round(sigma2_female, 4),
    `σ²_male`          = round(sigma2_male, 4),
    `# outliers`       = n_outliers,
    `p (male) raw`     = signif(p_male, 3),
    `p (male) robust`  = signif(p_male_robust, 3),
    `p (female)`       = signif(p_female, 3),
    Male               = if_else(flips, paste0(male_verdict, " *"), male_verdict)
  )

kable(results_display,
      caption = "Per-trait refit, sorted by robust male p. p(male) = boundary-corrected REML LRT for σ²_male (above male:env); robust = after dropping |studentized resid| > 4. p = 0.5 is the boundary value = no evidence. * = verdict changed after trimming.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(results$male_verdict == "male SIGNAL"), background = "#eafaf1") %>%
  row_spec(which(results$flips), background = "#fdedec") %>%
  row_spec(which(results$trait == "Grain yield"), bold = TRUE)
```

```{r}
#| label: male-signal-plot
#| fig-width: 8
#| fig-height: 5

ggplot(results, aes(x = reorder(trait, -p_male_robust), y = -log10(p_male_robust),
                    fill = male_verdict)) +
  geom_col() +
  # raw (pre-trim) male signal shown as a hollow marker, so outlier-inflated
  # traits reveal how far they drop after cleaning
  geom_point(aes(y = -log10(p_male)), shape = 21, fill = "white",
             color = "grey25", size = 2.4) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey30") +
  annotate("text", x = 1, y = -log10(0.05) + 0.15, label = "p = 0.05",
           hjust = 0, size = 3, color = "grey30") +
  coord_flip() +
  scale_fill_manual(values = c("male null" = "#7f8c8d", "male SIGNAL" = "#27ae60"),
                    name = NULL) +
  labs(
    title    = "Do the seven MOZL testers differ, trait by trait?",
    subtitle = "Bars = outlier-robust male LRT; hollow dots = raw (pre-trim). Above the dashed line = testers differ.",
    x = NULL, y = expression(-log[10](italic(p)~male~LRT))
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

# Data-quality screen: what the robust column is correcting for

The raw and robust male tests disagree for some traits because a few plots carry
**physically impossible values**. The screen below lists every plot whose studentized
residual exceeds 4 for any trait — these are the points dropped by the robust test, and
the ones Stelio should verify against the field books.

```{r}
#| label: outlier-screen

trait_cols <- set_names(trait_units$trait)

outliers <- map_dfr(trait_units$trait, function(nm) {
  d <- hybrids %>% mutate(y = .data[[nm]]) %>% filter(!is.na(y))
  r <- rstudent(lm(fix_form, data = d))
  d %>% mutate(rstud = r) %>%
    filter(abs(rstud) > 4) %>%
    transmute(trait = nm, env, rep, female, male,
              value = round(y, 2), studentized_resid = round(rstud, 1))
}) %>% arrange(desc(abs(studentized_resid)))

kable(outliers, caption = "Plots flagged as gross outliers (|studentized residual| > 4) — candidate data-entry errors") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

Two patterns stand out and should be fixed at source:

```{r}
#| label: impossible-values

# (1) A maize "ear weight" must be >= its "grain weight" (ear = grain + cob + husk).
ear_lt_grain <- hybrids %>%
  mutate(ear = `Ear weight`, grain = `Grain weight`, ratio = round(ear / grain, 2)) %>%
  filter(ear < grain) %>%
  transmute(env, rep, female, male, ear, grain, ratio) %>%
  arrange(ratio)

cat(sprintf("Plots where ear weight < grain weight (impossible): %d of %d\n",
            nrow(ear_lt_grain), nrow(hybrids)))

# (2) Implausible single values caught by the screen above, e.g. a 9 kg plot ear weight
#     (≈750 g/ear on 12 plants) and a 0.21 m plant height.
kable(head(ear_lt_grain, 8),
      caption = "Impossible ear < grain plots (head) — ear weight or grain weight mis-entered") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

::: {.callout-warning}
**These are data-cleaning flags, not biology.** The most extreme is `CKOKWE OPT / II /
CML364 × MOZL7` with **ear weight = 9004 g** (≈ 750 g per ear on 12 plants — impossible;
its own grain weight is only 697 g). It alone inflates ear weight's `σ²_male` to **4130** —
*larger* than `σ²_female` (1831), so by raw variance magnitude the testers would *look*
differentiated. The LRT against the male×env stratum is **not** fooled (ear weight stays
null, p ≈ 0.3–0.5), which is exactly why the verdict reads the LRT, not the variance size.
Ear weight is a **per-plot total** over ~25 plants, so the legitimate 3–5 kg plots are
normal (~150 g/ear); only the flagged plots are errors. Several plots also have
`ear < grain`, which cannot happen — likely ear/grain column swaps or typos.
:::

# Verdict

```{r}
#| label: verdict-logic

# All decisions use the OUTLIER-ROBUST male test.
n_total      <- nrow(results)
n_male_null  <- sum(results$male_verdict == "male null")
n_male_sig   <- sum(results$male_verdict == "male SIGNAL")
yield_null   <- results$male_verdict[results$trait == "Grain yield"] == "male null"
sig_traits   <- results$trait[results$male_verdict == "male SIGNAL"]
flip_traits  <- results$trait[results$flips]

verdict <- if (n_male_sig == 0) {
  "B-supported: testers null for essentially every trait — consistent with genome-wide near-identity; the 'same male' claim would generalize."
} else if (yield_null) {
  paste0("A-supported: the male term is null for grain yield and its components, but ",
         "robustly SIGNIFICANT for ", n_male_sig, " trait(s) — ",
         paste(sig_traits, collapse = ", "),
         ". The zero male GCA is a yield-selection artifact; it does NOT generalize to all phenotypes.")
} else {
  "Mixed/atypical: yield itself shows a male signal here — revisit the model before interpreting."
}

cat(sprintf("Traits swept                       : %d\n", n_total))
cat(sprintf("Male NULL  (robust p >= 0.05)      : %d\n", n_male_null))
cat(sprintf("Male SIGNAL (robust p < 0.05)      : %d\n", n_male_sig))
if (n_male_sig > 0)  cat(sprintf("  robust signal traits             : %s\n", paste(sig_traits, collapse = ", ")))
if (length(flip_traits) > 0)
  cat(sprintf("Outlier artifacts (raw signal lost): %s\n", paste(flip_traits, collapse = ", ")))
cat(sprintf("\nMechanism verdict:\n%s\n", verdict))
```

## How to read this

* **`isSingular()` flags *any* boundary component, not the male term specifically.** This
  model has nine random effects, and with only 4 environments / 7 testers some G×E or
  spatial term sits on the boundary for almost every trait — so `isSingular()` is `yes`
  for most rows *even when the male GCA is large and significant* (e.g. culm diameter, plant
  height). Read the singular flag **together with `σ²_male`**: it is the *combination*
  `isSingular == yes` **and** `σ²_male ≈ 0` that signals a pinned-to-zero tester variance
  (the boundary-zero of companion §1). The flag alone is not a per-male diagnostic.
* The **`p (male) robust`** column (boundary-corrected LRT, gross outliers removed) is the
  arbiter. Traits marked `*` changed verdict after trimming — in **both** directions: a few
  bad plots can *manufacture* a false signal (root weight) **or** *mask* a real one (days to
  silking and anthesis, where removing 2–3 errors uncovered a genuine tester difference).
  `p = 0.5` is the boundary value (the LRT statistic is 0) = no evidence at all.
* **Grain yield** (bold row) is the anchor: by construction it should land in the *male
  null* group. The diagnostic value is in the **other** traits.

::: {.callout-important}
**What the pattern means for your hypothesis.**

The robust result is **Mechanism A**. The testers are male-**null** for **grain yield and
every grain/yield component** (yield per plant, grain weight, ear weight, plants per plot)
— and also for ASI, plant height, culm diameter and root weight — but they **do** differ
(robust LRT p < 0.05) for **stover biomass (dry matter mass) and maturity (days to
anthesis, days to silking)**, traits selection never directly targeted.

* That the testers differ *at all* — in maturity and biomass — **rejects Mechanism B**:
  the seven are **not** near-identical genome-wide, so the zero yield GCA cannot be read as
  "the same line for everything."
* That the null is concentrated on **grain yield and its components** is the fingerprint of
  **ascertainment**: the testers were truncation-selected on yield, so they are homogenised
  exactly where they were selected — but they still carry the maturity and biomass
  differences that come free with diverse elite lines.
* A neat internal check: days to anthesis **and** days to silking each differ among
  testers, yet their difference **ASI is null** — the testers shift their whole flowering
  schedule together (different maturity, same anthesis–silking gap).

This matches the prior from the independent record — the same CIMMYT/IIAM elite pool *did*
respond to selection for ear height, ears-per-plant and ASI over seven years (Fato et al.
2025), so it is not genome-wide monomorphic.

**Bottom line for the original question:** "essentially the same male" is true **for
grain yield and its components**, but **false genome-wide** — the testers still differ in
maturity and stover. The extrapolation to all phenotypes does not hold.
:::

# Caveats

* **Why a likelihood-ratio test, not an ANOVA F-test.** A GCA main effect must be tested
  against its **male×environment** stratum, not against plot error. An ANOVA F-test against
  plot error is badly anticonservative for high-G×E traits — it manufactured spurious
  "signals" for culm diameter and plant height in an earlier draft. The boundary-corrected
  REML LRT (drop `(1|male)`, keep `(1|male:env)`) tests the right thing; `p = 0.5` is its
  no-evidence floor (statistic = 0, the 0.5·χ²₀ + 0.5·χ²₁ mixture).
* **Data quality is a real uncertainty.** Trimming changes the verdict in *both*
  directions: it removes a false signal (root weight, from impossible plots) and unmasks
  real ones (days to silking, days to anthesis). The data-quality screen lists ~38
  `ear < grain` plots and several gross single values. The robust column guards against the
  worst, but the cleaner move is for Stelio to **correct the flagged plots at source** and
  re-run.
* **The |studentized residual| > 4 trim is deliberately conservative** — it removes only
  gross errors, not legitimate biological spread. The robust signals (dry matter mass, days
  to anthesis, days to silking) hold under it; a softer threshold would not change those
  calls.
* **Power.** With only **7 testers** the male variance rests on 6 df, so the LRT has modest
  power. A non-significant result is "no detectable tester differentiation," not proof of
  identity. Treat borderline traits as ambiguous, and note the **female** GCA is the
  better-powered, more clearly significant signal throughout.
* **Trait scale / distribution.** Counts (`Plants per plot`) and the ASI difference are
  analysed on the raw scale for comparability with the yield pipeline; a Poisson or
  transformed refit could shift borderline p-values but not the qualitative null/signal
  split.
* **Correlated traits are not independent evidence.** Ear weight, grain weight and yield
  per plant are near-proxies for grain yield, and all land in the male-null group with it —
  consistent with the same ascertainment. The informative signals are the traits
  *genetically independent* of yield that survive trimming: dry matter mass and maturity
  (days to anthesis, days to silking).
* **Julian-day flowering.** Absolute day-of-year confounds planting date with flowering
  time, but `env` (fixed) absorbs the site-level offset, so among-tester GCA/SCA — the only
  thing tested here — is unaffected; ASI is planting-date-free by construction.
* **`lme4` vs `sommer`.** This sweep uses `lme4` to expose `isSingular()`; point estimates
  of the variance components can differ slightly from `sommer::mmer` but the boundary
  behaviour (which is what we test) is the same.

# Sources

* **Companion note:** *Zero Male GCA, Singular Fits, and Where the Genetic Variance Goes* —
  [`male_gca_singularity_and_variance_routing.html`](male_gca_singularity_and_variance_routing.html).
* **Fato, P. et al. (2025).** *Genetic Trends in Seven Years of Maize Breeding at
  Mozambique's Institute of Agricultural Research.* Agronomy 15(2):449. — non-yield traits
  respond to selection in this germplasm (prior favouring Mechanism A).
* **Masuka, B.P. et al. (2017).** *Genetic Diversity among Selected Elite CIMMYT Maize
  Hybrids in East and Southern Africa.* Crop Science 57:2395–2404. — low diversity of the
  upstream elite pool.
