---
title: "Is the Male Tester Null Only for Yield, or Genome-Wide? A Trait-by-Trait Refit"
subtitle: "Refitting the line × tester model on every phenotype in `data/multilocation.csv` and inspecting `isSingular()` and the male (tester) effect per trait"
author: "Fausto Rodriguez Zapata"
date: "2026-06-22"
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: true
    code-fold: show
    code-tools: true
---

::: {.callout-tip appearance="simple"}
**Source code & data:** <https://github.com/faustovrz/gcamoz> · **Companion note:** [`male_gca_singularity_and_variance_routing.html`](male_gca_singularity_and_variance_routing.html)
:::

# The question this notebook settles

The yield model in [`multilocation_alpha_lattice.html`](multilocation_alpha_lattice.html)
gives a **zero male (tester) GCA** and `isSingular() == TRUE`. The
[companion note](https://faustovrz.github.io/gcamoz/male_gca_singularity_and_variance_routing.html)
argues
this has a *germplasm* cause: the seven MOZL testers were **double-ascertained** (an
already-narrow CIMMYT/IIAM elite pool — CIMMYT ESA mean genetic distance ≈ 0.29, four
lines dominating 29–58 % of hybrids; Masuka et al. 2017 — then **truncation-selected on
yield** under phosphorus sufficiency), so they behave for combining-ability purposes as
*"essentially the same male."*

The open question (companion §3.3, §3.5) is **how far that "same male" claim reaches**:

| Mechanism | What it compresses | Male GCA zero for… |
|---|---|---|
| **A. Ascertainment / truncation on yield** | among-tester variance for **yield and yield-correlated traits** | yield (and its correlates) only |
| **B. Narrow base / near-identity-by-descent** | among-tester variance for **every trait** | **all** traits |

Your extrapolation — *"if the top seven are essentially the same line, the zero holds for
all phenotypes"* — requires **Mechanism B**. The decisive, falsifiable test is to **refit
the identical model on every measured trait** and read off, per trait:

1. `isSingular()` of the mixed model, and the REML **male GCA variance** (does it pin to 0?);
2. a boundary-corrected **REML likelihood-ratio test** of σ²_male — dropping `(1|male)`
   while keeping `(1|male:env)`, so the tester effect is tested against the correct
   male×environment stratum (do the testers differ at all for this trait?).

**Reading:** male null *only* for yield → **Mechanism A** (the zero is a yield-selection
artifact and does **not** generalize). Male null *across essentially all* traits →
**Mechanism B** (the testers really are interchangeable genome-wide, and the extrapolation
stands).

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(lme4)      # provides isSingular()
library(knitr)
library(kableExtra)
```

# Data, field layout, and derived flowering traits

We reuse the exact data-preparation pipeline of the yield notebook — same factor coding,
same recovery of field row/column from the 5 × 10 serpentine planting order — but we now
carry **all** plot-level phenotypes, not just yield. The `NAMULONGUE` check (`GEN = 0`) is
held out of the line × tester decomposition.

```{r}
#| label: read-and-prepare

raw_data <- read_csv("data/multilocation.csv", show_col_types = FALSE) %>%
  mutate(
    ENV    = factor(ENV),
    REP    = factor(REP, levels = c("I", "II", "III")),
    GEN    = factor(GEN, levels = as.character(0:49)),
    Female = factor(Female,
                    levels = c("Check",
                               "CML364", "CML366", "CML434", "CML435",
                               "CML439", "CML530", "CML532")),
    Male   = factor(Male,
                    levels = c("Check",
                               "MOZL3", "MOZL5", "MOZL6", "MOZL7",
                               "MOZL8", "MOZL9", "MOZL10"))
  )

# Field coordinates from the serpentine planting order (5 columns x 10 rows, snaking
# down along ENTRY) — identical to the yield notebook.
field <- raw_data %>%
  arrange(ENTRY) %>%
  group_by(ENV, REP) %>%
  mutate(
    plot_order = row_number(ENTRY),
    field_row  = ceiling(plot_order / 5),
    pos_in_row = ((plot_order - 1) %% 5) + 1,
    field_col  = if_else(field_row %% 2 == 1, pos_in_row, 6L - pos_in_row)
  ) %>%
  ungroup()
```

## Flowering dates → Julian day (no planting date available)

Stelio recorded **flowering dates**, not days-from-sowing, and there is **no planting
date** in the file. We therefore score flowering time as the **Julian day (day-of-year)**
of each event:

* **Days to silking** = Julian day of `F. D. Female` (female flowering / silk).
* **Days to anthesis** = Julian day of `F. D. Male` (male flowering / pollen shed).
* **ASI** (anthesis–silking interval) = silking − anthesis, a within-plot difference that
  is **independent of planting date** regardless.

Because `env` is a **fixed** effect in every model below, any site-level shift in planting
date is absorbed by the environment means — so the *among-tester* GCA/SCA estimates for the
Julian-day traits are unaffected by the missing sowing date (only the absolute intercept
would be). The raw dates come in two formats (`22/09` and `12-Oct`); we parse both.

```{r}
#| label: flowering-julian

fd_to_julian <- function(x) {
  x  <- trimws(as.character(x))
  dt <- parse_date_time(x, orders = c("d/m", "d-b", "dm", "db"),
                        locale = "C", quiet = TRUE)
  yday(dt)   # Julian day of year
}

field <- field %>%
  mutate(
    days_to_silking  = fd_to_julian(`F. D. Female`),
    days_to_anthesis = fd_to_julian(`F. D. Male`),
    ASI              = days_to_silking - days_to_anthesis
  )

# Confirm every flowering date parsed (no silent NAs introduced by format quirks).
parse_check <- tibble(
  column           = c("F. D. Female", "F. D. Male"),
  n_unparsed       = c(sum(is.na(field$days_to_silking)),
                       sum(is.na(field$days_to_anthesis)))
)
kable(parse_check, caption = "Flowering-date parse check (Julian day) — unparsed values should be 0") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Hybrids-only analysis frame and the trait dictionary

```{r}
#| label: hybrids-frame

hybrids <- field %>%
  filter(GEN != "0") %>%
  transmute(
    env    = droplevels(ENV),
    rep    = REP,
    row    = factor(field_row, levels = 1:10),
    col    = factor(field_col, levels = 1:5),
    female = droplevels(Female),
    male   = droplevels(Male),
    cross  = factor(`Female x Male`),
    # --- traits ---
    `Grain yield`        = `Yield t/ha`,
    `Yield per plant`    = `yield g/plant (g)`,
    `Plants per plot`    = `Plant Number`,
    `Ear weight`         = `Ear weight (g)`,
    `Grain weight`       = `grain weight (g)`,
    `Plant height`       = `Plant Height (m)`,
    `Culm diameter`      = `Culm Diameter (cm)`,
    `Days to silking`    = days_to_silking,
    `Days to anthesis`   = days_to_anthesis,
    `ASI`                = ASI,
    `Dry matter mass`    = `Dry matter mass (g)`,
    `Root weight`        = `Root Weight (g)`
  )

trait_units <- tribble(
  ~trait,              ~unit,
  "Grain yield",       "t/ha",
  "Yield per plant",   "g",
  "Plants per plot",   "count",
  "Ear weight",        "g/plot",
  "Grain weight",      "g/plot",
  "Plant height",      "m",
  "Culm diameter",     "cm",
  "Days to silking",   "Julian day",
  "Days to anthesis",  "Julian day",
  "ASI",               "days",
  "Dry matter mass",   "g",
  "Root weight",       "g"
)

kable(trait_units, caption = "Traits swept (all plot-level phenotypes in the file)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

traits <- trait_units$trait
```

# Data-quality screen

We screen every trait for gross data-entry errors **before** fitting anything, with a rule
fixed in advance: for each trait, fit a saturated fixed-effects model and flag any plot
whose **studentized residual exceeds 4** as a candidate error. Flagged plots are
**excluded from that trait's analysis** and listed here for Stelio to verify against the
field books. Every result in this notebook is computed on the **screened** data.

```{r}
#| label: screen-setup

# Saturated fixed-effects model, used ONLY to flag gross errors (never for the
# signal test). |studentized residual| > 4  =>  candidate data-entry error.
screen_form <- y ~ env + env:rep + female + male + cross

flag_keep <- function(d) {                 # TRUE = keep, FALSE = flagged outlier
  r <- rstudent(lm(screen_form, data = d))
  is.na(r) | abs(r) <= 4
}
```

```{r}
#| label: outlier-screen

outliers <- map_dfr(trait_units$trait, function(nm) {
  d <- hybrids %>% mutate(y = .data[[nm]]) %>% filter(!is.na(y))
  r <- rstudent(lm(screen_form, data = d))
  d %>% mutate(rstud = r) %>%
    filter(abs(rstud) > 4) %>%
    transmute(trait = nm, env, rep, female, male,
              value = round(y, 2), studentized_resid = round(rstud, 1))
}) %>% arrange(desc(abs(studentized_resid)))

kable(outliers, caption = "Flagged plots (|studentized residual| > 4), excluded per trait — candidate data-entry errors to verify against the field books.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

A second, physical check: a maize **ear weight** must be at least its **grain weight**
(ear = grain + cob + husk). Plots that violate this are mis-entered.

```{r}
#| label: impossible-values

ear_lt_grain <- hybrids %>%
  mutate(ear = `Ear weight`, grain = `Grain weight`, ratio = round(ear / grain, 2)) %>%
  filter(ear < grain) %>%
  transmute(env, rep, female, male, ear, grain, ratio) %>%
  arrange(ratio)

cat(sprintf("Plots where ear weight < grain weight (impossible): %d of %d\n",
            nrow(ear_lt_grain), nrow(hybrids)))

kable(head(ear_lt_grain, 8),
      caption = "Impossible ear < grain plots (head) — ear or grain weight mis-entered") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

::: {.callout-warning}
**Why we screen before interpreting any variance.** The worst single plot —
`CKOKWE OPT / II / CML364 × MOZL7`, **ear weight = 9004 g** (about 750 g per ear on 12
plants; its grain weight is only 697 g) — would, *if left in*, push ear weight's `σ²_male`
above its `σ²_female`, so by variance magnitude alone the testers would *look*
differentiated. Removing it (screening) is exactly what prevents that. Ear weight is a
**per-plot total** over ~25 plants, so legitimate 3–5 kg plots are normal (~150 g/ear);
only the flagged values are errors.
:::

# The model and the test

On the **screened** data we fit the same mixed model as the yield notebook, with
`lme4::lmer` (so `isSingular()` is available). The tester signal is judged by a
**boundary-corrected REML likelihood-ratio test** of σ²_male: drop `(1 | male)` while
**keeping** `(1 | male:env)`, so the tester *main* effect (GCA) is tested against its
**male×environment** stratum — the correct yardstick for a GCA in a multi-environment
trial. Testing a GCA against plot error is anticonservative, because each tester mean
averages over hundreds of plots, so almost any spread clears that bar. Under the null
σ²_male sits on the parameter boundary, so the LRT statistic follows a half-half mixture of
chi-square(0) and chi-square(1) and the naive p is halved; **`p = 0.5` is the no-evidence
floor.**

```{r}
#| label: model-formulas

# Random terms listed explicitly, so one can be dropped cleanly for the LRT.
rand_terms <- c("(1 | female)", "(1 | male)", "(1 | cross)",
                "(1 | female:env)", "(1 | male:env)", "(1 | cross:env)",
                "(1 | env:rep)", "(1 | env:rep:row)", "(1 | env:rep:col)")

build_form <- function(drop = character()) {
  as.formula(paste("y ~ env +", paste(setdiff(rand_terms, drop), collapse = " + ")))
}
mix_form <- build_form()
```

```{r}
#| label: fit-one-trait

fit_mm <- function(form, dat) suppressMessages(suppressWarnings(
  lmer(form, data = dat, REML = TRUE,
       control = lmerControl(check.conv.singular = "ignore",
                             optimizer = "bobyqa", optCtrl = list(maxfun = 2e5)))))

# Boundary-corrected REML LRT dropping (1|term) while keeping every other term
# (so (1|male) is tested ABOVE (1|male:env), the right GCA stratum). Under H0 the
# statistic ~ 0.5·χ²₀ + 0.5·χ²₁, so the one-sided p halves the naive χ²₁ p.
# Reduced model is refit directly with explicit data — no update().
lrt_vc <- function(full_fit, dat, term) {
  red  <- fit_mm(build_form(drop = paste0("(1 | ", term, ")")), dat)
  stat <- max(0, 2 * (as.numeric(logLik(full_fit)) - as.numeric(logLik(red))))
  min(0.5 * pchisq(stat, df = 1, lower.tail = FALSE), 1)
}

fit_one <- function(trait_name, data) {
  d_all <- data %>% mutate(y = .data[[trait_name]]) %>% filter(!is.na(y))
  keep  <- flag_keep(d_all)
  d     <- d_all[keep, ]                       # screened data = the analysis set

  fit <- fit_mm(mix_form, d)                   # PRIMARY model, on screened data
  vc  <- as.data.frame(VarCorr(fit))
  gv  <- function(g) { v <- vc$vcov[vc$grp == g]; if (length(v)) v else 0 }
  v_female <- gv("female"); v_male <- gv("male"); v_cross <- gv("cross")
  v_gen    <- v_female + v_male + v_cross

  p_male   <- lrt_vc(fit, d, "male")
  p_female <- lrt_vc(fit, d, "female")

  tibble(
    trait          = trait_name,
    singular       = isSingular(fit),
    sigma2_female  = v_female,
    sigma2_male    = v_male,
    sigma2_cross   = v_cross,
    # G×E strata and residual, carried for the entry-mean heritability section
    sigma2_fxe     = gv("female:env"),
    sigma2_mxe     = gv("male:env"),
    sigma2_cxe     = gv("cross:env"),
    sigma2_resid   = gv("Residual"),
    n_env          = n_distinct(d$env),                 # environments with data
    n_rep          = nrow(d) / n_distinct(d$env, d$cross), # mean reps per cross·env
    male_share_pct = if (v_gen > 0) 100 * v_male / v_gen else NA_real_,
    n_flagged      = sum(!keep),
    p_male         = p_male,
    p_female       = p_female
  )
}
```

# Run the sweep

```{r}
#| label: run-sweep
#| message: false
#| warning: false

results <- map_dfr(traits, fit_one, data = hybrids) %>%
  left_join(trait_units, by = "trait") %>%
  mutate(verdict = if_else(!is.na(p_male) & p_male < 0.05, "tester signal", "tester null")) %>%
  arrange(p_male)

results_display <- results %>%
  transmute(
    Trait        = trait,
    Unit         = unit,
    `Singular?`  = if_else(singular, "yes", "no"),
    `σ²_female`  = round(sigma2_female, 4),
    `σ²_male`    = round(sigma2_male, 4),
    `# flagged`  = n_flagged,
    `p(male)`    = signif(p_male, 3),
    `p(female)`  = signif(p_female, 3),
    Verdict      = verdict
  )

kable(results_display,
      caption = "Per-trait result (screened data), sorted by p(male). p(male) = boundary-corrected REML LRT for σ²_male, tested above male:env; p = 0.5 is the no-evidence floor. Green = tester signal; grain yield (bold) is the anchor.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(results$verdict == "tester signal"), background = "#eafaf1") %>%
  row_spec(which(results$trait == "Grain yield"), bold = TRUE)
```

```{r}
#| label: male-signal-plot
#| fig-width: 8
#| fig-height: 5

ggplot(results, aes(x = reorder(trait, -p_male), y = -log10(p_male),
                    fill = verdict)) +
  geom_col() +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey30") +
  annotate("text", x = 1, y = -log10(0.05) + 0.15, label = "p = 0.05",
           hjust = 0, size = 3, color = "grey30") +
  coord_flip() +
  scale_fill_manual(values = c("tester null" = "#7f8c8d", "tester signal" = "#27ae60"),
                    name = NULL) +
  labs(
    title    = "Do the seven MOZL testers differ, trait by trait?",
    subtitle = "Boundary-corrected male LRT on screened data. Bars above the dashed line = testers differ.",
    x = NULL, y = expression(-log[10](italic(p)~male~LRT))
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

# Heritability per trait

The same line × tester model already carries every variance component needed for
**entry-mean (cross-mean) heritability** — the accuracy with which cross rankings can
be selected on their mean across environments and replicates, and the most useful
heritability for multi-environment breeding decisions. Because the female (CIMMYT
lines) and male (MOZL testers) sets come from **distinct gene pools**, the among-cross
genetic variance partitions into an additive (GCA) and a dominance (SCA) part (Isik et
al. 2017; the full derivation is worked in [`gca_calculation.html`](gca_calculation.html)).
With $e$ environments and $r$ reps per environment:

* **Genetic variance of cross means:** $\sigma^2_G = \sigma^2_{\text{female}} + \sigma^2_{\text{male}} + \sigma^2_{\text{cross}}$
* **Additive component (entry-mean):** $\sigma^2_{A,\bar{x}} = \sigma^2_{\text{female}} + \sigma^2_{\text{male}}$
* **Dominance component (entry-mean):** $\sigma^2_{D,\bar{x}} = \sigma^2_{\text{cross}}$
* **Total G × E:** $\sigma^2_{GE} = \sigma^2_{\text{female:env}} + \sigma^2_{\text{male:env}} + \sigma^2_{\text{cross:env}}$
* **Entry-mean phenotypic variance:** $\sigma^2_{P,\bar{x}} = \sigma^2_G + \dfrac{\sigma^2_{GE}}{e} + \dfrac{\sigma^2_{\text{res}}}{e\,r}$
* **Narrow-sense heritability:** $h^2_{\bar{x}} = \sigma^2_{A,\bar{x}} / \sigma^2_{P,\bar{x}}$
* **Broad-sense heritability:** $H^2_{\bar{x}} = \sigma^2_G / \sigma^2_{P,\bar{x}}$

Both are bounded in $[0, 1]$, the correct behaviour for a properly specified entry-mean
form. The spatial terms (`env:rep`, `env:rep:row`, `env:rep:col`) are nuisance design
effects, not part of the genetic or phenotypic partition, so they do not enter these
formulas. Variance components are taken from the **same screened fits** as the sweep
above, so each trait's heritability is on its own analysis set.

## Why the dominance component is $\sigma^2_{\text{cross}}$

Each hybrid's genetic value is decomposed by the model into the two parents' general
combining abilities plus a specific-combination deviation:

$$g_{\text{cross}} = \underbrace{\text{GCA}_{\text{female}}}_{(1\,\mid\,\text{female})}
                   + \underbrace{\text{GCA}_{\text{male}}}_{(1\,\mid\,\text{male})}
                   + \underbrace{\text{SCA}_{\text{female}\times\text{male}}}_{(1\,\mid\,\text{cross})}.$$

The `cross` term (the `female × male` combination) is the **specific combining ability
(SCA)** — by construction the deviation of a hybrid from what its parents' GCAs predict.
GCA carries the **additive**, transmissible part of the genetic value; SCA carries what is
left over, the **non-additive** part (dominance, plus any epistasis). Taking the variance
of the three independent terms gives the among-cross genetic variance and its partition:

$$\sigma^2_G = \operatorname{Var}(g_{\text{cross}})
            = \underbrace{\sigma^2_{\text{female}} + \sigma^2_{\text{male}}}_{\sigma^2_{A,\bar{x}}\;(\text{additive, GCA})}
            + \underbrace{\sigma^2_{\text{cross}}}_{\sigma^2_{D,\bar{x}}\;(\text{dominance, SCA})}.$$

So the dominance entry-mean component is **literally the variance of the SCA effects**,
$\sigma^2_{D,\bar{x}} = \sigma^2_{\text{cross}}$.

**Why no factor of 4 here.** The relation $\sigma^2_D = 4\,\sigma^2_{\text{SCA}}$ (and
$\sigma^2_A = 2(\sigma^2_{\text{female}} + \sigma^2_{\text{male}})$) reconstructs the
*individual-organism* variances on the underlying plot scale, because SCA variance among
full-sib families equals only $\tfrac14\sigma^2_D$:

$$\sigma^2_{A,\text{ind}} = 2(\sigma^2_{\text{female}} + \sigma^2_{\text{male}}),
  \qquad \sigma^2_{D,\text{ind}} = 4\,\sigma^2_{\text{cross}}.$$

The **entry-mean** heritability here is not after individual $V_A$/$V_D$ — it partitions the
variance of the **cross means** that selection actually acts on. On that scale the
components enter *directly, unscaled*; applying the factor-4 would mix the two scales and
can drive $H^2 > 1$.

**A label caveat.** $\sigma^2_{\text{cross}}$ absorbs **all** non-additive genetic variance
— dominance *and* epistasis — so "dominance" is conventional shorthand; more precisely it is
the non-additive (SCA) variance. That is exactly why $H^2$, which includes
$\sigma^2_{\text{cross}}$, is *broad*-sense, while $h^2$, which uses only the GCA terms, is
*narrow*-sense.

```{r}
#| label: heritability-table

herit <- results %>%
  mutate(
    sigma2_G  = sigma2_female + sigma2_male + sigma2_cross,  # genetic var of cross means
    sigma2_A  = sigma2_female + sigma2_male,                 # additive (GCA) part
    sigma2_GE = sigma2_fxe + sigma2_mxe + sigma2_cxe,        # total G×E
    Vp_mean   = sigma2_G + sigma2_GE / n_env + sigma2_resid / (n_env * n_rep),
    h2_narrow = if_else(Vp_mean > 0, sigma2_A / Vp_mean, NA_real_),
    H2_broad  = if_else(Vp_mean > 0, sigma2_G / Vp_mean, NA_real_)
  ) %>%
  arrange(desc(H2_broad))

herit_display <- herit %>%
  transmute(
    Trait         = trait,
    Unit          = unit,
    `σ²_G`        = round(sigma2_G, 4),
    `σ²_GxE`      = round(sigma2_GE, 4),
    `σ²_resid`    = round(sigma2_resid, 4),
    e             = n_env,
    r             = round(n_rep, 1),
    `h² (narrow)` = round(h2_narrow, 3),
    `H² (broad)`  = round(H2_broad, 3)
  )

kable(herit_display,
      caption = "Entry-mean heritability per trait from the line × tester model (screened data), sorted by broad-sense H². σ²_G is the among-cross genetic variance (female + male + cross); h² counts only the additive GCA part (female + male). e = environments, r = mean reps per cross per environment. Grain yield (bold) is the anchor.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(herit$trait == "Grain yield"), bold = TRUE)
```

```{r}
#| label: heritability-plot
#| fig-width: 8
#| fig-height: 5

herit %>%
  transmute(trait, `h² (narrow)` = h2_narrow, `H² (broad)` = H2_broad) %>%
  pivot_longer(-trait, names_to = "kind", values_to = "value") %>%
  ggplot(aes(x = reorder(trait, value), y = value, fill = kind)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.65) +
  coord_flip(ylim = c(0, 1)) +
  scale_fill_manual(values = c("h² (narrow)" = "#2980b9", "H² (broad)" = "#8e44ad"),
                    name = NULL) +
  labs(
    title    = "Entry-mean heritability, trait by trait",
    subtitle = "Same line × tester model, screened data. Narrow-sense uses GCA only; broad-sense adds SCA.",
    x = NULL, y = "Entry-mean heritability"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "top")
```

**Reading.** The gap between $H^2$ (broad) and $h^2$ (narrow) is the SCA (dominance)
contribution; where the two bars are close, the cross-mean signal is almost all
additive GCA and so transmits reliably to the next generation. A trait can be highly
heritable here yet still male-null in the sweep above — $H^2$ pools the **female** GCA,
SCA and the male GCA, so it stays high on female differences alone even when the seven
testers are interchangeable ($\sigma^2_{\text{male}} \approx 0$). The two tables answer
different questions: the sweep asks *do the testers differ?*, while this one asks *how
repeatable is a cross mean?*

# Verdict

```{r}
#| label: verdict-logic

# All decisions use the screened male LRT (results$p_male).
n_total      <- nrow(results)
n_sig        <- sum(results$verdict == "tester signal")
n_null       <- n_total - n_sig
yield_null   <- results$verdict[results$trait == "Grain yield"] == "tester null"
sig_traits   <- results$trait[results$verdict == "tester signal"]

verdict <- if (n_sig == 0) {
  "B-supported: testers null for essentially every trait — consistent with genome-wide near-identity; the 'same male' claim would generalize."
} else if (yield_null) {
  paste0("A-supported: the male term is null for grain yield and its components, but ",
         "significant for ", n_sig, " trait(s) — ", paste(sig_traits, collapse = ", "),
         ". The zero male GCA is a yield-selection artifact; it does NOT generalize to all phenotypes.")
} else {
  "Mixed/atypical: yield itself shows a male signal here — revisit the model before interpreting."
}

cat(sprintf("Traits swept                 : %d\n", n_total))
cat(sprintf("Tester NULL    (p >= 0.05)   : %d\n", n_null))
cat(sprintf("Tester SIGNAL  (p <  0.05)   : %d\n", n_sig))
if (n_sig > 0)  cat(sprintf("  signal traits              : %s\n", paste(sig_traits, collapse = ", ")))
cat(sprintf("\nMechanism verdict:\n%s\n", verdict))
```

## How to read this

* **`isSingular()` flags *any* boundary component, not the male term specifically.** This
  model has nine random effects, and with only 4 environments / 7 testers some G×E or
  spatial term sits on the boundary for almost every trait — so `isSingular()` is `yes`
  for most rows *even when the male GCA is large and significant* (e.g. culm diameter, plant
  height). Read the singular flag **together with `σ²_male`**: it is the *combination*
  `isSingular == yes` **and** `σ²_male ≈ 0` that signals a pinned-to-zero tester variance
  (the boundary-zero of companion §1). The flag alone is not a per-male diagnostic.
* **`p(male)`** is the boundary-corrected LRT for the tester variance, computed on the
  screened data — the single arbiter. `p = 0.5` is the no-evidence floor (the LRT statistic
  is 0).
* **Grain yield** (bold row) is the anchor: by construction it should land in the *tester
  null* group. The diagnostic value is in the **other** traits.

::: {.callout-important}
**What the pattern means for your hypothesis.**

The robust result is **Mechanism A**. The testers are male-**null** for **grain yield and
every grain/yield component** (yield per plant, grain weight, ear weight, plants per plot)
— and also for ASI, plant height, culm diameter and root weight — but they **do** differ
(robust LRT p < 0.05) for **stover biomass (dry matter mass) and maturity (days to
anthesis, days to silking)**, traits selection never directly targeted.

* That the testers differ *at all* — in maturity and biomass — **rejects Mechanism B**:
  the seven are **not** near-identical genome-wide, so the zero yield GCA cannot be read as
  "the same line for everything."
* That the null is concentrated on **grain yield and its components** is the fingerprint of
  **ascertainment**: the testers were truncation-selected on yield, so they are homogenised
  exactly where they were selected — but they still carry the maturity and biomass
  differences that come free with diverse elite lines.
* A neat internal check: days to anthesis **and** days to silking each differ among
  testers, yet their difference **ASI is null** — the testers shift their whole flowering
  schedule together (different maturity, same anthesis–silking gap).

This matches the prior from the independent record — the same CIMMYT/IIAM elite pool *did*
respond to selection for ear height, ears-per-plant and ASI over seven years (Fato et al.
2025), so it is not genome-wide monomorphic.

**Bottom line for the original question:** "essentially the same male" is true **for
grain yield and its components**, but **false genome-wide** — the testers still differ in
maturity and stover. The extrapolation to all phenotypes does not hold.
:::

# Caveats

* **Why a likelihood-ratio test, not an ANOVA F-test.** A GCA main effect must be tested
  against its **male×environment** stratum, not against plot error. An ANOVA F-test against
  plot error is anticonservative for high-G×E traits — each tester mean averages over
  hundreds of plots, so almost any spread clears that yardstick. The boundary-corrected REML
  LRT (drop `(1|male)`, keep `(1|male:env)`) tests the right thing; `p = 0.5` is its
  no-evidence floor (statistic = 0, the half-half χ²₀/χ²₁ mixture).
* **Data quality.** The analysis runs on screened data (gross errors removed by the
  |studentized residual| > 4 rule); the screen also lists ~38 `ear < grain` plots that point
  to ear/grain mix-ups. Among the three signals, **dry matter mass** is the firmest; the two
  **maturity** signals (days to silking, days to anthesis, p ≈ 0.006–0.015) are the most
  modest and rest on only a handful of plots, so they are the calls to confirm once Stelio
  **corrects the flagged plots at source** and re-runs.
* **Power.** With only **7 testers** the male variance rests on 6 df, so the LRT has modest
  power. A non-significant result is "no detectable tester differentiation," not proof of
  identity. Treat borderline traits as ambiguous, and note the **female** GCA is the
  better-powered, more clearly significant signal throughout.
* **Trait scale / distribution.** Counts (`Plants per plot`) and the ASI difference are
  analysed on the raw scale for comparability with the yield pipeline; a Poisson or
  transformed refit could shift borderline p-values but not the qualitative null/signal
  split.
* **Correlated traits are not independent evidence.** Ear weight, grain weight and yield
  per plant are near-proxies for grain yield, and all land in the male-null group with it —
  consistent with the same ascertainment. The informative signals are the traits
  *genetically independent* of yield that survive screening: dry matter mass and maturity
  (days to anthesis, days to silking).
* **Julian-day flowering.** Absolute day-of-year confounds planting date with flowering
  time, but `env` (fixed) absorbs the site-level offset, so among-tester GCA/SCA — the only
  thing tested here — is unaffected; ASI is planting-date-free by construction.
* **`lme4` vs `sommer`.** This sweep uses `lme4` to expose `isSingular()`; point estimates
  of the variance components can differ slightly from `sommer::mmer` but the boundary
  behaviour (which is what we test) is the same.

# Sources

* **Companion note:** *Zero Male GCA, Singular Fits, and Where the Genetic Variance Goes* —
  [`male_gca_singularity_and_variance_routing.html`](male_gca_singularity_and_variance_routing.html).
* **Fato, P. et al. (2025).** *Genetic Trends in Seven Years of Maize Breeding at
  Mozambique's Institute of Agricultural Research.* Agronomy 15(2):449. — non-yield traits
  respond to selection in this germplasm (prior favouring Mechanism A).
* **Masuka, B.P. et al. (2017).** *Genetic Diversity among Selected Elite CIMMYT Maize
  Hybrids in East and Southern Africa.* Crop Science 57:2395–2404. — low diversity of the
  upstream elite pool.
