---
title: "Is the Male Tester Null Only for Yield, or Genome-Wide? A Trait-by-Trait Refit"
subtitle: "Refitting the line × tester model on every phenotype in `data/multilocation.csv` and inspecting `isSingular()` and the male (tester) effect per trait"
author: "Fausto Rodriguez Zapata"
date: "2026-06-22"
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: true
    code-fold: show
    code-tools: true
---

::: {.callout-tip appearance="simple"}
**Source code & data:** <https://github.com/faustovrz/gcamoz> · **Companion note:** [`male_gca_singularity_and_variance_routing.html`](male_gca_singularity_and_variance_routing.html)
:::

# The question this notebook settles

The yield model in [`multilocation_alpha_lattice.html`](multilocation_alpha_lattice.html)
gives a **zero male (tester) GCA** and `isSingular() == TRUE`. The companion note argues
this has a *germplasm* cause: the seven MOZL testers were **double-ascertained** (an
already-narrow CIMMYT/IIAM elite pool — CIMMYT ESA mean genetic distance ≈ 0.29, four
lines dominating 29–58 % of hybrids; Masuka et al. 2017 — then **truncation-selected on
yield** under phosphorus sufficiency), so they behave for combining-ability purposes as
*"essentially the same male."*

The open question (companion §3.3, §3.5) is **how far that "same male" claim reaches**:

| Mechanism | What it compresses | Male GCA zero for… |
|---|---|---|
| **A. Ascertainment / truncation on yield** | among-tester variance for **yield and yield-correlated traits** | yield (and its correlates) only |
| **B. Narrow base / near-identity-by-descent** | among-tester variance for **every trait** | **all** traits |

Your extrapolation — *"if the top seven are essentially the same line, the zero holds for
all phenotypes"* — requires **Mechanism B**. The decisive, falsifiable test is to **refit
the identical model on every measured trait** and read off, per trait:

1. `isSingular()` of the mixed model, and the REML **male GCA variance** (does it pin to 0?);
2. a boundary-free **fixed-effects ANOVA F-test** for the male term (do the testers differ
   at all for this trait?).

**Reading:** male null *only* for yield → **Mechanism A** (the zero is a yield-selection
artifact and does **not** generalize). Male null *across essentially all* traits →
**Mechanism B** (the testers really are interchangeable genome-wide, and the extrapolation
stands).

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(lme4)      # provides isSingular()
library(knitr)
library(kableExtra)
```

# Data, field layout, and derived flowering traits

We reuse the exact data-preparation pipeline of the yield notebook — same factor coding,
same recovery of field row/column from the 5 × 10 serpentine planting order — but we now
carry **all** plot-level phenotypes, not just yield. The `NAMULONGUE` check (`GEN = 0`) is
held out of the line × tester decomposition.

```{r}
#| label: read-and-prepare

raw_data <- read_csv("data/multilocation.csv", show_col_types = FALSE) %>%
  mutate(
    ENV    = factor(ENV),
    REP    = factor(REP, levels = c("I", "II", "III")),
    GEN    = factor(GEN, levels = as.character(0:49)),
    Female = factor(Female,
                    levels = c("Check",
                               "CML364", "CML366", "CML434", "CML435",
                               "CML439", "CML530", "CML532")),
    Male   = factor(Male,
                    levels = c("Check",
                               "MOZL3", "MOZL5", "MOZL6", "MOZL7",
                               "MOZL8", "MOZL9", "MOZL10"))
  )

# Field coordinates from the serpentine planting order (5 columns x 10 rows, snaking
# down along ENTRY) — identical to the yield notebook.
field <- raw_data %>%
  arrange(ENTRY) %>%
  group_by(ENV, REP) %>%
  mutate(
    plot_order = row_number(ENTRY),
    field_row  = ceiling(plot_order / 5),
    pos_in_row = ((plot_order - 1) %% 5) + 1,
    field_col  = if_else(field_row %% 2 == 1, pos_in_row, 6L - pos_in_row)
  ) %>%
  ungroup()
```

## Flowering dates → Julian day (no planting date available)

Stelio recorded **flowering dates**, not days-from-sowing, and there is **no planting
date** in the file. We therefore score flowering time as the **Julian day (day-of-year)**
of each event:

* **Days to silking** = Julian day of `F. D. Female` (female flowering / silk).
* **Days to anthesis** = Julian day of `F. D. Male` (male flowering / pollen shed).
* **ASI** (anthesis–silking interval) = silking − anthesis, a within-plot difference that
  is **independent of planting date** regardless.

Because `env` is a **fixed** effect in every model below, any site-level shift in planting
date is absorbed by the environment means — so the *among-tester* GCA/SCA estimates for the
Julian-day traits are unaffected by the missing sowing date (only the absolute intercept
would be). The raw dates come in two formats (`22/09` and `12-Oct`); we parse both.

```{r}
#| label: flowering-julian

fd_to_julian <- function(x) {
  x  <- trimws(as.character(x))
  dt <- parse_date_time(x, orders = c("d/m", "d-b", "dm", "db"),
                        locale = "C", quiet = TRUE)
  yday(dt)   # Julian day of year
}

field <- field %>%
  mutate(
    days_to_silking  = fd_to_julian(`F. D. Female`),
    days_to_anthesis = fd_to_julian(`F. D. Male`),
    ASI              = days_to_silking - days_to_anthesis
  )

# Confirm every flowering date parsed (no silent NAs introduced by format quirks).
parse_check <- tibble(
  column           = c("F. D. Female", "F. D. Male"),
  n_unparsed       = c(sum(is.na(field$days_to_silking)),
                       sum(is.na(field$days_to_anthesis)))
)
kable(parse_check, caption = "Flowering-date parse check (Julian day) — unparsed values should be 0") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## Hybrids-only analysis frame and the trait dictionary

```{r}
#| label: hybrids-frame

hybrids <- field %>%
  filter(GEN != "0") %>%
  transmute(
    env    = droplevels(ENV),
    rep    = REP,
    row    = factor(field_row, levels = 1:10),
    col    = factor(field_col, levels = 1:5),
    female = droplevels(Female),
    male   = droplevels(Male),
    cross  = factor(`Female x Male`),
    # --- traits ---
    `Grain yield`        = `Yield t/ha`,
    `Yield per plant`    = `yield g/plant (g)`,
    `Plants per plot`    = `Plant Number`,
    `Ear weight`         = `Ear weight (g)`,
    `Grain weight`       = `grain weight (g)`,
    `Plant height`       = `Plant Height (m)`,
    `Culm diameter`      = `Culm Diameter (cm)`,
    `Days to silking`    = days_to_silking,
    `Days to anthesis`   = days_to_anthesis,
    `ASI`                = ASI,
    `Dry matter mass`    = `Dry matter mass (g)`,
    `Root weight`        = `Root Weight (g)`
  )

trait_units <- tribble(
  ~trait,              ~unit,
  "Grain yield",       "t/ha",
  "Yield per plant",   "g",
  "Plants per plot",   "count",
  "Ear weight",        "g/plot",
  "Grain weight",      "g/plot",
  "Plant height",      "m",
  "Culm diameter",     "cm",
  "Days to silking",   "Julian day",
  "Days to anthesis",  "Julian day",
  "ASI",               "days",
  "Dry matter mass",   "g",
  "Root weight",       "g"
)

kable(trait_units, caption = "Traits swept (all plot-level phenotypes in the file)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)

traits <- trait_units$trait
```

# The per-trait engine

For each trait we fit the **same** random-effects structure used for yield, but with
`lme4::lmer` (so that `isSingular()` is directly available — this is the `lme4`
counterpart of the `sommer::mmer` model in the yield notebook). We then run a **fixed-effects
ANOVA** as a boundary-free signal test of the male term.

```{r}
#| label: model-formulas

# Mixed model (REML) — random female/male/cross GCA-SCA + their G×E + design terms.
mix_form <- y ~ env +
  (1 | female) + (1 | male) + (1 | cross) +
  (1 | female:env) + (1 | male:env) + (1 | cross:env) +
  (1 | env:rep) + (1 | env:rep:row) + (1 | env:rep:col)

# Fixed-effects ANOVA — boundary-free F-tests for female / male / cross.
# Balanced factorial ⇒ Type-I sequential SS for female→male→cross are order-stable.
fix_form <- y ~ env + env:rep + female + male + cross
```

```{r}
#| label: fit-one-trait

fit_one <- function(trait_name, data) {
  d <- data %>% mutate(y = .data[[trait_name]]) %>% filter(!is.na(y))

  ## --- REML mixed model + isSingular() ------------------------------------
  fit <- suppressMessages(suppressWarnings(
    lmer(mix_form, data = d, REML = TRUE,
         control = lmerControl(check.conv.singular = "ignore",
                               optimizer = "bobyqa",
                               optCtrl = list(maxfun = 2e5)))
  ))
  vc <- as.data.frame(VarCorr(fit))
  gv <- function(g) { v <- vc$vcov[vc$grp == g]; if (length(v)) v else 0 }
  v_female <- gv("female"); v_male <- gv("male"); v_sca <- gv("cross")
  v_gen    <- v_female + v_male + v_sca

  ## --- fixed-effects ANOVA (boundary-free signal tests) -------------------
  at <- anova(lm(fix_form, data = d))
  pf <- function(term) if (term %in% rownames(at)) at[term, "Pr(>F)"] else NA_real_

  tibble(
    trait          = trait_name,
    singular       = isSingular(fit),
    sigma2_female  = v_female,
    sigma2_male    = v_male,
    sigma2_sca     = v_sca,
    male_share_pct = if (v_gen > 0) 100 * v_male / v_gen else NA_real_,
    p_female       = pf("female"),
    p_male         = pf("male"),
    p_cross        = pf("cross")
  )
}
```

# Run the sweep

```{r}
#| label: run-sweep
#| message: false
#| warning: false

results <- map_dfr(traits, fit_one, data = hybrids) %>%
  left_join(trait_units, by = "trait") %>%
  mutate(
    male_verdict = if_else(!is.na(p_male) & p_male < 0.05, "male SIGNAL", "male null"),
    mechanism    = if_else(male_verdict == "male null", "A-consistent", "rejects A-as-universal")
  ) %>%
  arrange(p_male)

results_display <- results %>%
  transmute(
    Trait          = trait,
    Unit           = unit,
    `Singular?`    = if_else(singular, "yes", "no"),
    `σ²_female`    = round(sigma2_female, 4),
    `σ²_male`      = round(sigma2_male, 4),
    `male % of gen`= round(male_share_pct, 1),
    `p (male)`     = signif(p_male, 3),
    `p (female)`   = signif(p_female, 3),
    `p (cross/SCA)`= signif(p_cross, 3),
    Male           = male_verdict
  )

kable(results_display,
      caption = "Per-trait refit: isSingular(), REML male GCA variance, and fixed-effects ANOVA tests (sorted by male p-value)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(results$male_verdict == "male SIGNAL"), background = "#fdf2e9") %>%
  row_spec(which(results$trait == "Grain yield"), bold = TRUE)
```

```{r}
#| label: male-signal-plot
#| fig-width: 8
#| fig-height: 5

ggplot(results, aes(x = reorder(trait, -p_male), y = -log10(p_male),
                    fill = male_verdict)) +
  geom_col() +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey30") +
  annotate("text", x = 1, y = -log10(0.05) + 0.15, label = "p = 0.05",
           hjust = 0, size = 3, color = "grey30") +
  coord_flip() +
  scale_fill_manual(values = c("male null" = "#7f8c8d", "male SIGNAL" = "#e67e22"),
                    name = NULL) +
  labs(
    title    = "Do the seven MOZL testers differ, trait by trait?",
    subtitle = "Fixed-effects ANOVA male F-test. Bars above the dashed line = testers DO differ for that trait.",
    x = NULL, y = expression(-log[10](italic(p)~male))
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

# Verdict

```{r}
#| label: verdict-logic

n_total      <- nrow(results)
n_male_null  <- sum(results$male_verdict == "male null")
n_male_sig   <- sum(results$male_verdict == "male SIGNAL")
yield_null   <- results$male_verdict[results$trait == "Grain yield"] == "male null"
sig_traits   <- results$trait[results$male_verdict == "male SIGNAL"]

verdict <- if (n_male_sig == 0) {
  "B-supported: the testers are null for essentially every trait — consistent with genome-wide near-identity. The 'same male' claim generalizes."
} else if (yield_null) {
  paste0("A-supported: the male term is null for yield but SHOWS SIGNAL for ",
         n_male_sig, " trait(s) — ", paste(sig_traits, collapse = ", "),
         ". The zero male GCA is a yield-selection artifact and does NOT generalize to all phenotypes.")
} else {
  "Mixed/atypical: yield itself shows a male signal here — revisit the model before interpreting."
}

cat(sprintf("Traits swept            : %d\n", n_total))
cat(sprintf("Male NULL (p >= 0.05)   : %d\n", n_male_null))
cat(sprintf("Male SIGNAL (p < 0.05)  : %d\n", n_male_sig))
if (n_male_sig > 0) cat(sprintf("  signal traits         : %s\n", paste(sig_traits, collapse = ", ")))
cat(sprintf("\nMechanism verdict:\n%s\n", verdict))
```

## How to read this

* **`isSingular()` flags *any* boundary component, not the male term specifically.** This
  model has nine random effects, and with only 4 environments / 7 testers some G×E or
  spatial term sits on the boundary for almost every trait — so `isSingular()` is `yes`
  for most rows *even when the male GCA is large and significant* (e.g. culm diameter, plant
  height). Read the singular flag **together with `σ²_male`**: it is the *combination*
  `isSingular == yes` **and** `σ²_male ≈ 0` that signals a pinned-to-zero tester variance
  (the boundary-zero of companion §1). The flag alone is not a per-male diagnostic.
* The **fixed-effects ANOVA `p (male)`** is the boundary-free arbiter. A trait can show a
  near-zero REML `σ²_male` yet a small ANOVA `p (male)` if the signal is weak but real;
  conversely a clearly non-significant `p (male)` confirms the testers truly do not differ.
* **Grain yield** (bold row) is the anchor: by construction it should land in the *male
  null* group. The diagnostic value is in the **other** traits.

::: {.callout-important}
**What the pattern means for your hypothesis.**

* If **only yield** (± its tight correlates such as ear weight / grain weight / yield per
  plant) is male-null while traits selection never targeted — flowering Julian day, ASI,
  plant height, culm diameter, root weight — show a **male signal**, that is **Mechanism A**:
  the zero male GCA is an *ascertainment artifact of selecting the testers on yield*, and it
  **does not** license "the testers are the same line for everything."
* If the testers are **male-null across essentially all** traits, that is **Mechanism B**:
  evidence the seven really are near-identical-by-descent, and the genome-wide
  extrapolation holds.

The independent record makes **A** the prior expectation: the same CIMMYT/IIAM elite pool
*did* respond to selection for plant height, ear height, ears-per-plant and ASI over seven
years (Fato et al. 2025), so it is not genome-wide monomorphic. This sweep tests that
expectation directly in *these* seven testers.
:::

# Caveats

* **Power.** With only **7 testers** the male variance rests on 6 df; the fixed-effects
  ANOVA male F-test is the most reliable signal test, but it still has modest power. A
  non-significant `p (male)` is "no detectable tester differentiation," not proof of
  identity. Treat borderline traits (0.01 < p < 0.10) as ambiguous.
* **Trait scale / distribution.** Counts (`Plants per plot`) and the ASI difference are
  analysed on the raw scale for comparability with the yield pipeline; a Poisson or
  transformed refit could shift borderline p-values but not the qualitative null/signal
  split.
* **Correlated traits are not independent evidence.** Ear weight, grain weight and yield
  per plant are near-proxies for grain yield; if they pattern with yield that reflects the
  same ascertainment, not extra confirmation. The informative traits are those
  *genetically independent* of yield (phenology, height, root, culm).
* **Julian-day flowering.** Absolute day-of-year confounds planting date with flowering
  time, but `env` (fixed) absorbs the site-level offset, so among-tester GCA/SCA — the only
  thing tested here — is unaffected; ASI is planting-date-free by construction.
* **`lme4` vs `sommer`.** This sweep uses `lme4` to expose `isSingular()`; point estimates
  of the variance components can differ slightly from `sommer::mmer` but the boundary
  behaviour (which is what we test) is the same.

# Sources

* **Companion note:** *Zero Male GCA, Singular Fits, and Where the Genetic Variance Goes* —
  [`male_gca_singularity_and_variance_routing.html`](male_gca_singularity_and_variance_routing.html).
* **Fato, P. et al. (2025).** *Genetic Trends in Seven Years of Maize Breeding at
  Mozambique's Institute of Agricultural Research.* Agronomy 15(2):449. — non-yield traits
  respond to selection in this germplasm (prior favouring Mechanism A).
* **Masuka, B.P. et al. (2017).** *Genetic Diversity among Selected Elite CIMMYT Maize
  Hybrids in East and Southern Africa.* Crop Science 57:2395–2404. — low diversity of the
  upstream elite pool.
