---
title: "Stelio's Multilocation Maize Trial: RCBD Field Layout and Line × Tester GCA/SCA"
subtitle: "Serpentine 5 × 10 field layout, row/column spatial control, and line x tester GCA/SCA from `data/multilocation.csv`"
author: "Fausto Rodriguez Zapata"
date: "2026-06-04"
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: true
    code-fold: show
    code-tools: true
bibliography: references.bib
csl: https://www.zotero.org/styles/apa
---

::: {.callout-tip appearance="simple"}
**Source code & data:** <https://github.com/faustovrz/gcamoz>
:::

::: {.callout-note}
This notebook documents the experimental design of Stelio's real file, `data/multilocation.csv`, **and** runs the full combining-ability analysis directly on it. The real file carries the **line x tester pedigree** (`Female`, `Male`, `Female x Male`) and a **replicated commercial check**, in addition to the plot-level phenotypes. It describes a **multilocation field trial** — an RCBD with a 5 × 10 serpentine plot layout per replication: 49 line x tester hybrids plus one repeated check, evaluated in 3 replications per environment across 4 environments. The earlier `gca_calculation.qmd` ran the same pipeline on *simulated* RCBD data; here the EDA, arithmetic GCA/SCA, mixed-model variance components, breeding values, and heritabilities are all computed on Stelio's actual `Yield t/ha` across all four environments.
:::

# Theoretical Foundation & Experimental Setup

Before fitting a genetic model, we must separate three ideas that are easy to mix together:

1. The **genetic material** being evaluated.
2. The **experimental allocation / field layout** used to control local variation.
3. The **phenotypic model** used to analyze the traits.

The file contains plot-level phenotypes from a multilocation maize trial **and** the pedigree key that maps each genotype to its female and male parent. This means the field-design structure and a line x tester GCA/SCA analysis can both be set up directly from a single file — no external metadata table is needed anymore.

## Breeding Scheme (Line x Tester)

Stelio's research focuses on the genetic architecture of low-phosphorus use efficiency in maize. The 49 hybrid genotypes represent a **Factorial Mating Design**, often referred to as a **Line x Tester** or **North Carolina Design II**.

Two distinct parental sets are crossed:

* **Female parents / lines:** 7 CIMMYT CML tropical donor founders.
* **Male parents / testers:** 7 Mozambique tester lines (`MOZL` series).
* **Hybrid families:** all 7 x 7 = **49 F1 hybrid combinations**.

In addition to the 49 hybrids, the trial includes a **repeated commercial check** (`NAMULONGUE`) so that environments and replications can be calibrated against a known standard.

## Genetic Material Planted in the Field

The data now identify each plot with both a numeric genotype code (`GEN`) and the explicit pedigree (`Female`, `Male`, `Female x Male`):

* `GEN` has **50 levels**: `0` for the check and `1`–`49` for the hybrids.
* `GEN = 0` is the **`NAMULONGUE` check** (`Female = Check`, `Male = Check`).
* `GEN = 1`–`49` are the line x tester hybrids.

The real founder identifiers (which replace the earlier `MOZ1`–`MOZ7` placeholders) are:

| Parent set | Role in factorial design | Real IDs in the file |
| :--- | :--- | :--- |
| CIMMYT CML donors | Female / tropical donor lines | `CML364`, `CML366`, `CML434`, `CML435`, `CML439`, `CML530`, `CML532` |
| Mozambique testers | Male local tester lines | `MOZL3`, `MOZL5`, `MOZL6`, `MOZL7`, `MOZL8`, `MOZL9`, `MOZL10` |
| Commercial check | Standard reference entry | `NAMULONGUE` (`GEN = 0`) |

::: {.callout-note}
The Mozambique testers are numbered `MOZL3`, `MOZL5`, `MOZL6`, `MOZL7`, `MOZL8`, `MOZL9`, `MOZL10` (seven testers, not a simple 1–7 sequence). The CIMMYT line `CML530` is part of the female set; the earlier draft had incorrectly listed `CML440`.
:::

# R Setup and Data Import

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(sommer)
library(knitr)
library(kableExtra)

raw_data <- read_csv("data/multilocation.csv", show_col_types = FALSE) %>%
  mutate(
    ENV    = factor(ENV),
    REP    = factor(REP, levels = c("I", "II", "III")),
    GEN    = factor(GEN, levels = as.character(0:49)),
    Female = factor(Female,
                    levels = c("Check",
                               "CML364", "CML366", "CML434", "CML435",
                               "CML439", "CML530", "CML532")),
    Male   = factor(Male,
                    levels = c("Check",
                               "MOZL3", "MOZL5", "MOZL6", "MOZL7",
                               "MOZL8", "MOZL9", "MOZL10")),
    yield  = `Yield t/ha`
  )

# Field coordinates from the serpentine planting order (Stelio's procedure:
# first plot top-left, 5 columns x 10 rows, snaking down along ENTRY).
field <- raw_data %>%
  arrange(ENTRY) %>%
  group_by(ENV, REP) %>%
  mutate(
    plot_order = row_number(ENTRY),                 # 1 .. 50 within each env x rep
    field_row  = ceiling(plot_order / 5),           # 1 (top) .. 10 (bottom)
    pos_in_row = ((plot_order - 1) %% 5) + 1,
    field_col  = if_else(field_row %% 2 == 1, pos_in_row, 6L - pos_in_row)  # serpentine
  ) %>%
  ungroup()

# Hybrids-only analysis frame (the check, GEN = 0, is held out of the line x
# tester decomposition). `row`/`col` are the real field positions of each plot.
hybrids <- field %>%
  filter(GEN != "0") %>%
  transmute(
    env    = droplevels(ENV),
    rep    = REP,
    row    = factor(field_row, levels = 1:10),
    col    = factor(field_col, levels = 1:5),
    female = droplevels(Female),
    male   = droplevels(Male),
    cross  = factor(`Female x Male`),
    yield  = yield
  )
```

## Entries Planted in the Field (including checks)

The table below is the complete entry key actually planted, generated directly from the file. The check is entry `GEN = 0`; the 49 hybrids follow.

```{r}
#| label: entry-key

entry_key <- raw_data %>%
  distinct(GEN, Female, Male, `Female x Male`) %>%
  arrange(GEN) %>%
  rename(cross = `Female x Male`)

kable(entry_key, caption = "All 50 entries planted: 1 check (GEN = 0) + 49 line x tester hybrids") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "420px")
```

## Real Factorial Mating Grid (7 x 7)

Each cell is the genotype code `GEN` assigned to that female x male cross. The full grid is present, with no missing or duplicated cells.

```{r}
#| label: mating-grid

mating_grid <- raw_data %>%
  filter(GEN != "0") %>%
  distinct(Female, Male, GEN) %>%
  droplevels() %>%
  arrange(Female, Male) %>%
  pivot_wider(names_from = Male, values_from = GEN)

kable(mating_grid, caption = "Real 7 x 7 line x tester grid; cells show the GEN code for each cross") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Experimental Design

> The field design is the actual plot allocation used to evaluate the genotypes produced by the 7 x 7 line-by-tester crossing design, together with the repeated check.

Stelio's trial is a **field trial**: each plot is a full row of maize. Within every replication the 50 entries (49 hybrids + 1 check) are laid out on a **5-column × 10-row grid** and sown in **serpentine order** — first plot top-left, snaking down (see the field-layout map below).

The **49 hybrids are randomized within each replication**, and the `NAMULONGUE` check travels along the planting sequence. Because the hybrids are completely randomized within a replication — with **no separately randomized incomplete blocks** — the trial is analyzed as a **randomized complete block design (RCBD)**, with the replication as the complete block. To guard against field gradients across the 5 × 10 grid, the plot **row** and **column** (recovered from the serpentine planting order) are added as nested spatial terms — a light spatial correction layered on top of the RCBD.

## Locations / Environments

The trial contains four environments/locations, recorded in the `ENV` column:

| Location code in `ENV` | Interpretation in this notebook |
| :--- | :--- |
| `CHOKWE STS` | Chokwe stress environment |
| `CKOKWE OPT` | Chokwe optimum environment, as spelled in the CSV |
| `NHACOONGO` | Nhacoongo environment |
| `SUSSUNDENGA` | Sussundenga environment |

## Field Layout Within Each Location

Within each location and replication:

* all 49 hybrids appear **once**, plus **one** check plot — 50 plots total;
* the 50 plots occupy a **5 × 10 field grid** (5 columns × 10 rows);
* each replication is therefore a complete replicate of the 50 entries (one complete block).

Across the trial:

* **Environments (`ENV`)**: 4.
* **Replications (`REP`)**: `I`, `II`, and `III` within each environment.
* **Entries (`GEN`)**: 49 hybrids + 1 check = 50 per replication.
* **Traits**: grain yield, yield per plant, plant number, ear weight, grain weight, plant height, culm diameter, flowering dates, dry matter mass, and root weight.

This is a standard RCBD: each replication is one complete block of 50 plots. The only refinement we add for analysis is the plot **row** and **column** within each replication (from the 5 × 10 serpentine), fitted as random spatial terms so that systematic field trends are not left in the residual.

# Design Diagnostics

## Basic Dimensions

```{r}
#| label: design-summary

design_summary <- raw_data %>%
  summarize(
    rows          = n(),
    environments  = n_distinct(ENV),
    replications  = n_distinct(REP),
    total_entries = n_distinct(GEN),
    hybrids       = n_distinct(GEN[GEN != "0"]),
    checks        = n_distinct(GEN[GEN == "0"])
  )

kable(design_summary, caption = "Basic dimensions of Stelio's multilocation dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Replication Completeness

Every environment-by-replication now holds exactly 50 plots (49 hybrids + 1 check), with no duplicated entries.

```{r}
#| label: replication-completeness

rep_summary <- raw_data %>%
  count(ENV, REP, name = "observed_rows") %>%
  left_join(
    raw_data %>%
      summarize(unique_entries = n_distinct(GEN), .by = c(ENV, REP)),
    by = c("ENV", "REP")
  ) %>%
  mutate(
    expected_rows = 50,
    difference    = observed_rows - expected_rows
  )

kable(rep_summary, caption = "Observed rows and unique entries per environment-by-replication") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Cross and Check Replication Balance

```{r}
#| label: balance-check

hybrid_balance <- raw_data %>%
  filter(GEN != "0") %>%
  count(`Female x Male`, name = "plots") %>%
  count(plots, name = "n_crosses")

check_balance <- raw_data %>%
  filter(GEN == "0") %>%
  count(ENV, REP, name = "check_plots")

kable(hybrid_balance, caption = "Every hybrid cross is observed in the same number of plots") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

kable(check_balance, caption = "The NAMULONGUE check appears once per environment-by-replication") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

::: {.callout-note}
**What changed from the earlier draft.** An earlier version of this notebook reported 597 rows and a suspicious "duplicated `GEN = 49`" in three environments, plus a `NHACOONGO` that looked one replicate short. That was an artifact of the check being mislabeled as a second `GEN = 49` and missing from `NHACOONGO`. In the corrected file the check is its own entry (`GEN = 0`, `NAMULONGUE`), present once per environment-by-replication, giving a balanced **600 rows** (49 hybrids x 12 + 12 checks).
:::

# Field Coordinates from the Planting Order

The CSV has no explicit row/column columns, but the `ENTRY` column records Stelio's planting order, and the field was sown in a fixed serpentine: first plot top-left, then across a **5-column × 10-row** grid, snaking down. From that order we recover each plot's field **row** (1–10) and **column** (1–5) — computed once in the `field` frame in the setup chunk. The grid is complete in every environment-by-replication:

```{r}
#| label: field-grid-check

grid_check <- field %>%
  summarize(
    rows  = n_distinct(field_row),
    cols  = n_distinct(field_col),
    plots = n(),
    .by   = c(ENV, REP)
  ) %>%
  arrange(ENV, REP)

kable(grid_check, caption = "Every environment-by-replication fills a complete 5-column × 10-row field grid (50 plots)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  scroll_box(height = "300px")
```

These row and column coordinates are used as nested spatial terms (`env:rep:row`, `env:rep:col`) in the mixed model, so that field gradients across the 5 × 10 grid are absorbed rather than left in the residual.

# Field Layout: Serpentine Planting Map

Each replication is sown as a **5-column × 10-row** field block (5 × 10 = 50 plots) in **serpentine (boustrophedon) order** — Stelio's planting procedure: the first entry is placed in the **top-left corner**, the first row is filled left-to-right, the next row right-to-left, and so on down the block. Consecutively-planted plots are therefore physical neighbours.

The map below uses the **`ENTRY` column — the recorded planting/allocation order — within each replication** as the planting sequence and lays it onto the 5 × 10 serpentine. (`ENTRY` runs 1–600 in contiguous blocks of 50, one block per environment-by-replication; NHACOONGO occupies 301–450.) Each tile is labelled with its `GEN` code and **coloured by entry type** (line × tester hybrid vs. the `NAMULONGUE` check), so the position of the check within each replication is immediately visible.

```{r}
#| label: serpentine-layout
#| fig.width: 11
#| fig.height: 7
#| warning: false

field_layout <- field %>%
  filter(ENV == "NHACOONGO") %>%   # field_row / field_col come from the setup chunk
  mutate(entry_type = if_else(GEN == "0", "Check (NAMULONGUE)", "Line × tester hybrid"))

entry_palette <- c(
  "Line × tester hybrid" = "#2C7FB8",
  "Check (NAMULONGUE)"        = "#D7301F"
)

ggplot(field_layout, aes(x = field_col, y = field_row, fill = entry_type)) +
  geom_tile(color = "white", linewidth = 0.7) +
  geom_text(aes(label = GEN), size = 3, fontface = "bold", color = "white") +
  facet_wrap(~ REP, nrow = 1) +
  scale_fill_manual(values = entry_palette, name = "Entry type") +
  scale_x_continuous(breaks = 1:5, position = "top", expand = c(0, 0)) +
  scale_y_reverse(breaks = 1:10, expand = c(0, 0)) +
  coord_equal() +
  labs(
    title    = "Serpentine field layout for NHACOONGO (5 columns × 10 rows per replication)",
    subtitle = "First entry at the top-left corner; planting order snakes left–right then right–left. Tile = GEN code, coloured by entry type.",
    x = "Field column",
    y = "Field row (planting progresses top → bottom)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid      = element_blank(),
    strip.text      = element_text(face = "bold"),
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

::: {.callout-note}
This map is the field's planting layout: `ENTRY` is the sowing order and the 5 × 10 serpentine is Stelio's planting procedure (first plot top-left, snaking down). The field **row** and **column** read off this map are exactly the coordinates fed to the `env:rep:row` and `env:rep:col` spatial terms in the mixed model below.
:::

# Phenotypic Model

For grain yield, the fitted multilocation **RCBD** model — written out in full, exactly as it is fit below — separates each plot into the experimental-design terms, the line × tester combining-ability terms, their environment interactions, and error:

$$
\begin{aligned}
y_{hirc(pq)} = \;& \mu + E_h + R_{i(h)} + \text{Row}_{r(hi)} + \text{Col}_{c(hi)} \\
              & + g_{F_p} + g_{M_q} + s_{pq} \\
              & + (Eg_F)_{hp} + (Eg_M)_{hq} + (Es)_{hpq} \\
              & + \epsilon_{hirc(pq)}
\end{aligned}
$$

for the plot holding hybrid $(p, q)$ — female $p$ × male $q$ — in field row $r$, column $c$, replication $i$, and environment $h$:

**Experimental design (RCBD + spatial)**

* $\mu$ — overall mean.
* $E_h$ — environment (fixed; the four sites differ in mean).
* $R_{i(h)}$ — **replication (complete block) nested within environment: the RCBD blocking factor.**
* $\text{Row}_{r(hi)}$, $\text{Col}_{c(hi)}$ — field row and column nested within replication, the spatial correction across the 5 × 10 grid.

**Genetic — line × tester decomposition of the genotype effect** (North Carolina Design II partition of @comstock1948 and @isik2017)

* $g_{F_p}$ — female (line) GCA; $g_{M_q}$ — male (tester) GCA; $s_{pq}$ — SCA of the cross. Together these give the genotype effect $G_k = g_{F_p} + g_{M_q} + s_{pq}$.

**Genotype × environment** (the combined $(EG)_{hk}$ split into its three sources)

* $(Eg_F)_{hp}$ — female-GCA × environment; $(Eg_M)_{hq}$ — male-GCA × environment; $(Es)_{hpq}$ — SCA × environment.

**Error**

* $\epsilon_{hirc(pq)}$ — plot residual.

Because the female (CIMMYT) and male (Mozambique) sets are mutually exclusive, `Female` and `Male` are fit as **separate** random effects — no design-matrix overlay is needed, unlike a diallel [@isik2017]. The check (`GEN = 0`) is held outside this decomposition as a reference entry. Each random term maps one-to-one onto the `sommer` call fitted below:

| Model term | `sommer` random term |
| :--- | :--- |
| $R_{i(h)}$ | `env:rep` |
| $\text{Row}_{r(hi)}$ / $\text{Col}_{c(hi)}$ | `env:rep:row` / `env:rep:col` |
| $g_{F_p}$ / $g_{M_q}$ / $s_{pq}$ | `female` / `male` / `cross` |
| $(Eg_F)_{hp}$ / $(Eg_M)_{hq}$ / $(Es)_{hpq}$ | `female:env` / `male:env` / `cross:env` |

::: {.callout-note}
The field **row** and **column** terms come from the 5 × 10 serpentine planting layout above — Stelio's actual field coordinates — so they absorb genuine field gradients rather than leaving them in the residual. The replication (`env:rep`) is the RCBD block; row and column refine it within each replication.
:::

# Exploratory Data Analysis (EDA)

We summarise the real `Yield t/ha` of the 49 hybrid crosses before fitting any model.

## Mean Yield Grid

Mean hybrid yield in each female × male cell, averaged across all 4 environments × 3 replications.

```{r}
#| label: eda-grid

mean_yields <- hybrids %>%
  group_by(female, male) %>%
  summarize(mean_yield = round(mean(yield), 2), .groups = "drop") %>%
  pivot_wider(names_from = male, values_from = mean_yield)

kable(mean_yields, caption = "7 × 7 Factorial F1 Yield Grid (mean Yield t/ha across 4 environments × 3 reps)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) %>%
  row_spec(0, bold = TRUE, background = "#1a5276", color = "white")
```

## Performance across Environments

Yield distribution of the 49 hybrids in each of the four environments, with the `NAMULONGUE` check mean overlaid as a reference point.

```{r}
#| label: eda-plot
#| fig.width: 10
#| fig.height: 6

check_means <- raw_data %>%
  filter(GEN == "0") %>%
  group_by(ENV) %>%
  summarize(check_mean = mean(yield), .groups = "drop")

ggplot(hybrids, aes(x = env, y = yield, fill = env)) +
  geom_boxplot(alpha = 0.85, outlier.color = "grey40") +
  geom_point(
    data = check_means,
    aes(x = ENV, y = check_mean),
    inherit.aes = FALSE,
    shape = 23, size = 3.5, fill = "#D7301F", color = "black"
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title    = "Hybrid grain yield across the four environments",
    subtitle = "Boxplots: 49 line × tester hybrids (3 reps each). Red diamond: NAMULONGUE check mean.",
    x = "Environment",
    y = "Yield (t/ha)",
    fill = "Environment"
  ) +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", color = "#1a202c"))
```

# Fixed-Effects Calculations: Factorial Combining Ability

In a balanced factorial cross, GCA and SCA can be calculated arithmetically from family means. This provides a transparent foundation before using mixed models.

## Arithmetic Formulas

Let $\bar{Y}_{p.}$ be the mean yield of all crosses involving female parent $p$,
$\bar{Y}_{.q}$ the mean yield of all crosses involving male parent $q$,
$\bar{Y}_{pq}$ the mean of the hybrid cross between female $p$ and male $q$,
and $\bar{Y}_{..}$ the overall grand mean of all crosses in the factorial grid.

*   **Grand Mean:** $\bar{Y}_{..} = \frac{1}{f \cdot m} \sum_{p} \sum_{q} \bar{Y}_{pq}$
*   **GCA of Female Parent $p$:** $g_{F_p} = \bar{Y}_{p.} - \bar{Y}_{..}$
*   **GCA of Male Parent $q$:** $g_{M_q} = \bar{Y}_{.q} - \bar{Y}_{..}$
*   **SCA of Cross $pq$:** $s_{pq} = \bar{Y}_{pq} - \bar{Y}_{p.} - \bar{Y}_{.q} + \bar{Y}_{..}$

## Direct R Arithmetic

We calculate these effects arithmetically from the entire hybrid dataset (all 4 environments pooled). The check is excluded.

```{r}
#| label: arithmetic-calculations

# 1. Family means over all environments and reps
family_means <- hybrids %>%
  group_by(female, male) %>%
  summarize(mean_yield = mean(yield), .groups = "drop")

# 2. Reshape to a 7x7 matrix (7 CIMMYT females in rows, 7 MOZL males in columns)
Y_matrix <- matrix(
  0, nrow = 7, ncol = 7,
  dimnames = list(levels(hybrids$female), levels(hybrids$male))
)
for (i in seq_len(nrow(family_means))) {
  Y_matrix[as.character(family_means$female[i]),
           as.character(family_means$male[i])] <- family_means$mean_yield[i]
}

# 3. Grand mean, marginal means
mu_hat       <- mean(Y_matrix)
female_means <- rowMeans(Y_matrix)
male_means   <- colMeans(Y_matrix)

# 4. Fixed GCA effects
gca_female_fixed <- female_means - mu_hat
gca_male_fixed   <- male_means - mu_hat

# 5. Fixed SCA effects
sca_fixed <- matrix(
  0, nrow = 7, ncol = 7,
  dimnames = list(levels(hybrids$female), levels(hybrids$male))
)
for (p in rownames(sca_fixed)) {
  for (q in colnames(sca_fixed)) {
    sca_fixed[p, q] <- Y_matrix[p, q] - gca_female_fixed[p] - gca_male_fixed[q] - mu_hat
  }
}

gca_female_df <- tibble(
  Female_Parent = names(gca_female_fixed),
  Mean_Yield    = round(female_means, 3),
  Estimated_GCA = round(gca_female_fixed, 4)
)
kable(gca_female_df, caption = "Estimated GCA Effects for Female Parents (CIMMYT CML Lines), Yield t/ha") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

gca_male_df <- tibble(
  Male_Parent   = names(gca_male_fixed),
  Mean_Yield    = round(male_means, 3),
  Estimated_GCA = round(gca_male_fixed, 4)
)
kable(gca_male_df, caption = "Estimated GCA Effects for Male Parents (MOZL Testers), Yield t/ha") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

kable(round(sca_fixed, 4), caption = "Arithmetic 7 × 7 Specific Combining Ability (SCA) grid, Yield t/ha") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) %>%
  row_spec(0, bold = TRUE, background = "#4a5568", color = "white")
```

# Mixed-Effects Calculations: Factorial Variance Components across Environments

In modern breeding, combining ability is analyzed with **Linear Mixed Models (LMMs)**: parents are treated as random samples from a reference population, and each combining-ability term gets a paired environment interaction so its G×E variance can be estimated separately [@isik2017]. We fit the model with `sommer::mmer` on all four environments. Because female and male lines are distinct, **no matrix overlays are needed**.

## Fitting the Mixed Model

*   `env` enters as a **fixed effect** (different overall means across the four sites).
*   `env:rep` captures replications (complete blocks) nested within environment.
*   `env:rep:row` and `env:rep:col` capture the **field row and column** nested within each replication — the spatial correction across the 5 × 10 grid.
*   `female`, `male`, `cross` are the random Female GCA, Male GCA, and SCA terms.
*   `female:env`, `male:env`, `cross:env` are their G×E interactions.

```{r}
#| label: mixed-model-sommer

fit_gxe <- mmer(
  fixed  = yield ~ env,
  random = ~ female + male + cross +
             female:env + male:env + cross:env +
             env:rep + env:rep:row + env:rep:col,
  data    = hybrids,
  verbose = FALSE
)

var_comps <- summary(fit_gxe)$varcomp
kable(var_comps, caption = "Variance components — joint line × tester model across all 4 environments") %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

## Extracting GCA BLUPs

We extract the General Combining Ability predictions (BLUPs) for both parent sets.

```{r}
#| label: gca-blups

female_blups <- fit_gxe$U$female$yield
male_blups   <- fit_gxe$U$male$yield

female_blups_df <- tibble(
  CIMMYT_Female = gsub("female", "", names(female_blups)),
  GCA_BLUP      = round(as.numeric(female_blups), 4)
) %>%
  arrange(desc(GCA_BLUP))
kable(female_blups_df, caption = "Female GCA BLUPs (CIMMYT CML Lines), across-environment Yield t/ha") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

male_blups_df <- tibble(
  MOZL_Male = gsub("male", "", names(male_blups)),
  GCA_BLUP  = round(as.numeric(male_blups), 4)
) %>%
  arrange(desc(GCA_BLUP))
kable(male_blups_df, caption = "Male GCA BLUPs (MOZL Testers), across-environment Yield t/ha") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Extracting SCA BLUPs

We extract the Specific Combining Ability predictions (dominance deviations) and format them into a $7 \times 7$ grid. Each cross BLUP is mapped back to its female/male cell via the pedigree key.

```{r}
#| label: sca-blups

cross_blups <- fit_gxe$U$cross$yield

# Pedigree key: cross string -> (female, male)
cross_key <- hybrids %>%
  distinct(cross, female, male) %>%
  mutate(cross = as.character(cross))

sca_df <- tibble(
  cross    = gsub("cross", "", names(cross_blups)),
  SCA_BLUP = as.numeric(cross_blups)
) %>%
  left_join(cross_key, by = "cross")

sca_matrix_blup <- matrix(
  0, nrow = 7, ncol = 7,
  dimnames = list(levels(hybrids$female), levels(hybrids$male))
)
for (i in seq_len(nrow(sca_df))) {
  sca_matrix_blup[as.character(sca_df$female[i]),
                  as.character(sca_df$male[i])] <- sca_df$SCA_BLUP[i]
}

kable(round(sca_matrix_blup, 4), caption = "7 × 7 Specific Combining Ability (SCA) BLUP grid, across-environment Yield t/ha") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered")) %>%
  row_spec(0, bold = TRUE, background = "#4a5568", color = "white")
```

# Breeding Values (BV) and Genotypic Values (GV)

For hybrid selection we combine the BLUPs into two quantities:

1.  **Breeding Value ($BV$):** the additive value of a cross, the sum of parent GCAs.
    $$BV_{pq} = g_{F_p} + g_{M_q}$$
2.  **Genotypic Value ($GV$):** the total genetic value, $BV$ plus the SCA (dominance) deviation.
    $$GV_{pq} = g_{F_p} + g_{M_q} + s_{pq}$$

```{r}
#| label: bv-gv-calc

genetic_values <- expand.grid(
  Female = levels(hybrids$female),
  Male   = levels(hybrids$male),
  stringsAsFactors = FALSE
) %>%
  mutate(
    GCA_Female         = map_dbl(Female, ~ female_blups_df$GCA_BLUP[female_blups_df$CIMMYT_Female == .x]),
    GCA_Male           = map_dbl(Male,   ~ male_blups_df$GCA_BLUP[male_blups_df$MOZL_Male == .x]),
    SCA                = map2_dbl(Female, Male, ~ sca_matrix_blup[.x, .y]),
    Breeding_Value_BV  = GCA_Female + GCA_Male,
    Genotypic_Value_GV = Breeding_Value_BV + SCA
  ) %>%
  arrange(desc(Genotypic_Value_GV))

kable(head(genetic_values, 15), caption = "Top 15 line × tester hybrids ranked by Genotypic Value (across-environment Yield t/ha)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Across-Environment Heritability with G × E

We report **entry-mean (full-sib family-mean) heritability** — the accuracy with which cross rankings can be selected based on their mean across replicates and environments. With $e$ environments and $r$ reps per environment:

*   **Genetic variance of cross means:** $\sigma^2_G = \sigma^2_{GCA_F} + \sigma^2_{GCA_M} + \sigma^2_{SCA}$
*   **Additive component (entry mean):** $\sigma^2_{A,\bar{x}} = \sigma^2_{GCA_F} + \sigma^2_{GCA_M}$
*   **Total G × E variance:** $\sigma^2_{GE} = \sigma^2_{GCA_F \times E} + \sigma^2_{GCA_M \times E} + \sigma^2_{SCA \times E}$
*   **Entry-mean phenotypic variance:** $\sigma^2_{P,\bar{x}} = \sigma^2_G + \dfrac{\sigma^2_{GE}}{e} + \dfrac{\sigma^2_{\text{res}}}{e \cdot r}$
*   **Narrow-sense entry-mean heritability:** $h^2_{\bar{x}} = \sigma^2_{A,\bar{x}} / \sigma^2_{P,\bar{x}}$
*   **Broad-sense entry-mean heritability:** $H^2_{\bar{x}} = \sigma^2_G / \sigma^2_{P,\bar{x}}$

```{r}
#| label: gxe-heritability

# Pick variance components by regex on rowname. Interaction terms can appear as
# "female:env" or "env:female" depending on sommer's term ordering; match both.
# NOTE: every pattern is ANCHORED with ^. This is essential — an unanchored
# "male:env" would also match "female:env" (the word "female" ends in "male"),
# double-counting female GCA×E into the male term and inflating total G×E.
pick_vc <- function(vc, pat) {
  idx <- grep(pat, rownames(vc))
  if (length(idx) == 0) 0 else sum(vc[idx, "VarComp"])
}

var_gca_f_g <- pick_vc(var_comps, "^female\\.")
var_gca_m_g <- pick_vc(var_comps, "^male\\.")
var_sca_g   <- pick_vc(var_comps, "^cross\\.")
var_gca_f_e <- pick_vc(var_comps, "^female:env|^env:female")
var_gca_m_e <- pick_vc(var_comps, "^male:env|^env:male")
var_sca_e   <- pick_vc(var_comps, "^cross:env|^env:cross")
var_rep_g   <- pick_vc(var_comps, "^env:rep\\.")        # replication-in-env (literal dot excludes row/col)
var_row_g   <- pick_vc(var_comps, "^env:rep:row")       # field rows within replication
var_col_g   <- pick_vc(var_comps, "^env:rep:col")       # field columns within replication
var_res_g   <- pick_vc(var_comps, "^units\\.")

e_env <- 4   # number of environments
r_rep <- 3   # reps per environment

var_G_mean <- var_gca_f_g + var_gca_m_g + var_sca_g
var_A_mean <- var_gca_f_g + var_gca_m_g
var_GE     <- var_gca_f_e + var_gca_m_e + var_sca_e

# Replication, row, and column effects are design (nuisance) terms: they are
# removed by the experimental layout and do not enter the entry-mean phenotypic
# variance, exactly as `rep` does not. Modelling them only shrinks the residual.
var_phen_mean <- var_G_mean + var_GE / e_env + var_res_g / (e_env * r_rep)

h2_mean <- var_A_mean / var_phen_mean
H2_mean <- var_G_mean / var_phen_mean

gxe_results <- tibble(
  Parameter = c(
    "GCA Female variance",
    "GCA Male variance",
    "SCA variance",
    "GCA_F × Env variance",
    "GCA_M × Env variance",
    "SCA × Env variance",
    "Total G × E variance",
    "Replication variance (env:rep)",
    "Field-row variance (env:rep:row)",
    "Field-column variance (env:rep:col)",
    "Residual variance",
    "Genetic variance of cross means (Vg)",
    "Entry-mean phenotypic variance (Vp_mean)",
    "Narrow-sense h² (entry mean across envs)",
    "Broad-sense H² (entry mean across envs)"
  ),
  Value = round(c(
    var_gca_f_g, var_gca_m_g, var_sca_g,
    var_gca_f_e, var_gca_m_e, var_sca_e,
    var_GE, var_rep_g, var_row_g, var_col_g, var_res_g,
    var_G_mean, var_phen_mean,
    h2_mean, H2_mean
  ), 4)
)

kable(gxe_results, caption = "Genetic variances, G × E, and entry-mean heritabilities — joint four-environment model") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

> [!TIP]
> *   If $\sigma^2_{GE}$ is a substantial fraction of $\sigma^2_{GCA_F} + \sigma^2_{GCA_M} + \sigma^2_{SCA}$, the ranking of parents and crosses is **environment-dependent** — selection should be done within target environments (e.g., a Chokwe stress site) rather than on across-environment means.
> *   Both heritabilities are bounded in $[0, 1]$, as expected from a properly specified entry-mean form.

::: {.callout-important}
**Why these estimates look the way they do — and what is real vs. fixable.**

*   **Male (tester) GCA and SCA×E estimate to ~0, and this is correct.** A fixed-effects ANOVA gives male `MS = 1.55`, *below* the residual `MS = 2.23` (`p = 0.65`), so the unconstrained REML variance is negative and is pinned at 0. The arithmetic SCA "spread" from family means (≈ 0.18) is fully explained by the sampling error of a 12-plot mean (residual/12 ≈ 0.19). The real, significant signal is **female (line) GCA** (`p = 0.0002`) and **female GCA × environment** (`p = 0.004`): the seven CIMMYT donor lines differ and re-rank across sites; the seven local testers do not.

*   **Narrow-sense $h^2_{\bar{x}}$ is genuinely low (~0.18).** Additive variance is small and one-sided (males add nothing), and $\sigma^2_{GCA_F \times E}$ is *larger* than $\sigma^2_{GCA_F}$ — the lines re-rank strongly across four climatically distinct sites (NHACOONGO mean 5.25 vs. SUSSUNDENGA 7.41 t/ha). Across-environment narrow-sense heritability *should* be low when G×E dominates the additive signal.

*   **The field row/column spatial terms are modelled (`env:rep:row`, `env:rep:col`), which is why $H^2_{\bar{x}}$ is realistic.** The field **row** carries a **highly significant** variance (≈ 0.56, `Z ≈ 4.4`) and the column a smaller one (≈ 0.11, `Z ≈ 1.9`); together they absorb genuine field gradient across the 5 × 10 grid that would otherwise inflate the residual. Compared with a plain RCBD (rep only) the residual drops ~32% (2.24 → 1.53) and broad-sense $H^2_{\bar{x}}$ rises from ≈ 0.30 to ≈ **0.45**. These coordinates are Stelio's real planting positions (5 × 10 serpentine), so this is a legitimate spatial correction, not an arbitrary split.
:::

# Step-by-Step Checklist for Stelio {#sec-checklist}

To complete the GCA and SCA analysis for his PhD from `data/multilocation.csv`, Stelio should follow these structured steps:

1.  **Prepare the data:** keep the columns `ENV`, `REP`, `GEN`, `Female` (CIMMYT CML lines), `Male` (MOZL testers), `Female x Male` (cross ID), and the trait (`Yield t/ha`). Hold the `NAMULONGUE` check (`GEN = 0`) as a separate reference entry.

2.  **Import to R:** load `tidyverse` and `sommer`, read the CSV, set `ENV`, `REP`, `Female`, `Male`, and the cross as factors, and build a hybrids-only frame (drop `GEN = 0`).

3.  **Run the field-layout / design diagnostics:** confirm 600 balanced rows (49 hybrids × 12 + 12 checks) and 50 plots per environment-by-replication, recover each plot's field row (1–10) and column (1–5) from the `ENTRY` serpentine order, and visualise the 5 × 10 planting map.

4.  **Run line × tester arithmetic:** hand-calculate GCA and SCA from family means to understand the dataset before modelling.

5.  **Fit the mixed model:** use `sommer::mmer` with `female`, `male`, and `cross` as random effects plus their `:env` interactions; `env` fixed; and the design terms `env:rep` (RCBD block) plus `env:rep:row` and `env:rep:col` (spatial control) random. Do **not** use overlay matrix functions — the parental sets are distinct.

6.  **Calculate Breeding and Genotypic Values:**
    *   Use parental GCA BLUPs for $BV = g_F + g_M$ (parental recombination selection).
    *   Use $GV = BV + SCA$ to select the best commercial hybrid entries.

7.  **Estimate heritability with G × E:** extract variance components, separate the G × E terms, and apply the entry-mean line × tester formulas to check selection accuracy across environments.

8.  **Compare against the check:** benchmark top hybrids against the replicated `NAMULONGUE` check in each environment.

9.  **Confirm the design — alpha-lattice vs. RCBD:** Stelio described the trial as an **alpha (incomplete-block) design**, so he should clarify and provide the **incomplete-block (sub-block) assignment** for each plot. However, the data give **no sign of an alpha lattice**: within each 5 × 10 replication the hybrids appear fully randomized, with no incomplete, repeated sub-blocks of hybrids — which is why the trial is analyzed here as an **RCBD with row/column spatial control**. If Stelio supplies real sub-block labels, add an `env:rep:block` term and compare fits (log-likelihood / AIC) against the current row/column model; if he cannot, the RCBD + spatial model stands. For a finer field-trend correction, consider a continuous row/column covariate or a 2-D AR1×AR1 spatial structure on row × column.

# References {.unnumbered}

::: {#refs}
:::
