---
title: "Stelio's Multilocation Maize Trial as an Alpha-Lattice Design"
subtitle: "Experimental-design notes and field-layout reconstruction from `data/multilocation.csv`"
author: "Fausto Rodriguez Zapata"
date: "2026-06-02"
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: true
    code-fold: show
    code-tools: true
---

::: {.callout-tip appearance="simple"}
**Source code & data:** <https://github.com/faustovrz/gcamoz>
:::

::: {.callout-note}
This notebook updates the experimental-design description for Stelio's real file, `data/multilocation.csv`. The older `gca_calculation.qmd` was built from simulated data under an RCBD assumption. The real file is more consistent with a **multilocation alpha-lattice trial**: 49 genotypes evaluated in 3 replications per environment, with each replication arranged as 7 incomplete blocks of 7 plots.
:::

# Theoretical Foundation & Experimental Setup

Before fitting a genetic model, we must separate three ideas that are easy to mix together:

1. The **genetic material** being evaluated.
2. The **field layout** used to control local field variation.
3. The **phenotypic model** used to analyze the traits.

The file Stelio sent contains plot-level phenotypes from a multilocation maize trial. It does **not** contain female-parent and male-parent labels, so this notebook focuses on the field-design structure. A GCA/SCA analysis still requires a genotype key that maps each `GEN` code to its female and male parents.

## Breeding Scheme (Line x Tester)

Stelio's research focuses on the genetic architecture of low-phosphorus use efficiency in maize. My current understanding follows the same genetic interpretation used in `docs/gca_calculation.html`, section **1.1.1 Stelio's Factorial Mating Grid**: the 49 genotypes are intended to represent a **Factorial Mating Design**, often referred to as a **Line x Tester** or **North Carolina Design II**.

In this mating scheme, two distinct parental sets are crossed:

* **Female parents / lines:** the CIMMYT CML tropical donor founders.
* **Male parents / testers:** the Mozambique founder/tester lines.
* **Hybrid families:** all possible crosses between the 7 female parents and 7 male parents, giving **49 F1 hybrid combinations**.

This is the genetic design assumption. The field design for Stelio's real file is a separate issue: those 49 genotypes were evaluated in a multilocation alpha-lattice layout, not in the simulated RCBD layout used by the earlier teaching file.

## Genetic Material and Pedigree Metadata

The real data identify entries with the column `GEN`. In this file:

* `GEN` has **49 unique levels**: `1` through `49`.
* These 49 entries are the treatments/genotypes evaluated in the field.
* The file does not include columns such as `female`, `male`, or `cross`, so `GEN` must be joined to an external pedigree/crossing key before line-by-tester GCA and SCA can be estimated.

The **known tropical founder parents** are the CIMMYT CML donor lines already described in the project. The Mozambique tester/founder identifiers are not yet available, so this notebook temporarily keeps the placeholder labels `MOZ1` to `MOZ7`.

::: {.callout-warning}
`MOZ1` to `MOZ7` are provisional identifiers. Stelio should replace these labels with the real Mozambique founder/tester IDs before final GCA/SCA analysis.
:::

| Parent set | Role in factorial design | Current IDs | Status |
| :--- | :--- | :--- | :--- |
| CIMMYT CML donors | Female/tropical donor founders | `CML364`, `CML366`, `CML434`, `CML435`, `CML439`, `CML440`, `CML532` | Known from the project description |
| Mozambique founders/testers | Male local tester founders | `MOZ1`, `MOZ2`, `MOZ3`, `MOZ4`, `MOZ5`, `MOZ6`, `MOZ7` | Placeholder labels until Stelio supplies the real IDs |

## Placeholder Factorial Mating Grid

The table below repeats the factorial mating-grid assumption from the simulated RCBD notebook, but here it is explicitly a **placeholder pedigree map** for the real alpha-lattice dataset. It shows the expected `7 x 7 = 49` hybrid structure. Stelio needs to verify which `GEN` code in `multilocation.csv` corresponds to each cross.

| CIMMYT donor / Mozambique tester | MOZ1 | MOZ2 | MOZ3 | MOZ4 | MOZ5 | MOZ6 | MOZ7 |
| :--- | :---: | :---: | :---: | :---: | :---: | :---: | :---: |
| **CML364** | CML364 x MOZ1 | CML364 x MOZ2 | CML364 x MOZ3 | CML364 x MOZ4 | CML364 x MOZ5 | CML364 x MOZ6 | CML364 x MOZ7 |
| **CML366** | CML366 x MOZ1 | CML366 x MOZ2 | CML366 x MOZ3 | CML366 x MOZ4 | CML366 x MOZ5 | CML366 x MOZ6 | CML366 x MOZ7 |
| **CML434** | CML434 x MOZ1 | CML434 x MOZ2 | CML434 x MOZ3 | CML434 x MOZ4 | CML434 x MOZ5 | CML434 x MOZ6 | CML434 x MOZ7 |
| **CML435** | CML435 x MOZ1 | CML435 x MOZ2 | CML435 x MOZ3 | CML435 x MOZ4 | CML435 x MOZ5 | CML435 x MOZ6 | CML435 x MOZ7 |
| **CML439** | CML439 x MOZ1 | CML439 x MOZ2 | CML439 x MOZ3 | CML439 x MOZ4 | CML439 x MOZ5 | CML439 x MOZ6 | CML439 x MOZ7 |
| **CML440** | CML440 x MOZ1 | CML440 x MOZ2 | CML440 x MOZ3 | CML440 x MOZ4 | CML440 x MOZ5 | CML440 x MOZ6 | CML440 x MOZ7 |
| **CML532** | CML532 x MOZ1 | CML532 x MOZ2 | CML532 x MOZ3 | CML532 x MOZ4 | CML532 x MOZ5 | CML532 x MOZ6 | CML532 x MOZ7 |

For analysis, Stelio should add either a metadata file or columns to the field dataset with this minimum structure:

| Required column | Meaning | Example value |
| :--- | :--- | :--- |
| `GEN` | Genotype code used in `multilocation.csv` | `1` |
| `female` | CIMMYT CML donor parent | `CML364` |
| `male` | Mozambique tester parent | `MOZ1` or real tester ID |
| `cross` | Hybrid/cross identifier | `CML364 x MOZ1` |

This would make the real dataset parallel to the simulated RCBD file, which already had `female`, `male`, and `cross` columns. Until this pedigree metadata is available, the real data can support alpha-lattice field-design diagnostics and genotype-level analysis, but not a defensible final GCA/SCA decomposition.

## Experimental Design

> The alpha-lattice design is the actual field-plot arrangement used to evaluate the 49 genotypes produced by the 7 x 7 line-by-tester crossing design.

Stelio's real field trial is a **multilocation alpha-lattice design**. This replaces the RCBD field-layout assumption used in the simulated teaching notebook. Stelio's statement that this is an alpha lattice matches the dimensions in the file. With 49 genotypes, the natural resolvable incomplete-block layout is:

$$49 = 7 \times 7$$

### Locations / Environments

The trial contains four environments/locations, recorded in the `ENV` column:

| Location code in `ENV` | Interpretation in this notebook |
| :--- | :--- |
| `CHOKWE STS` | Chokwe stress environment |
| `CKOKWE OPT` | Chokwe optimum environment, as spelled in the CSV |
| `NHACOONGO` | Nhacoongo environment |
| `SUSSUNDENGA` | Sussundenga environment |

### Alpha-Lattice Layout Within Each Location

Within each location and replication:

* there are **7 incomplete blocks**;
* each incomplete block contains **7 plots/genotypes**;
* all 49 genotypes appear once per replication;
* the replication is therefore a complete replicate, but it is subdivided into smaller incomplete blocks.

Across the trial:

* **Replications (`REP`)**: `I`, `II`, and `III` within each environment.
* **Genotypes (`GEN`)**: 49 unique entries.
* **Traits**: grain yield, yield per plant, plant number, ear weight, grain weight, plant height, culm diameter, flowering dates, dry matter mass, and root weight.

This is different from an RCBD. In an RCBD, each replication is treated as one large complete block of 49 plots. In an alpha lattice, each replication is split into incomplete blocks, so the field error model can account for local variation among smaller neighboring plot groups.

## How the Data Fit the Alpha Lattice

The clean alpha-lattice expectation is:

$$4\ \text{environments} \times 3\ \text{replications} \times 7\ \text{incomplete blocks} \times 7\ \text{plots} = 588\ \text{plots}$$

The file has 597 rows, not 588. The reason is specific and diagnosable: in three environments (`CHOKWE STS`, `CKOKWE OPT`, and `SUSSUNDENGA`), each replication has one duplicated record for `GEN = 49`. `NHACOONGO` has the expected 49 rows per replication.

Therefore, the design evidence is:

* The genotype set is exactly 49 entries.
* Every `ENV x REP` combination has 49 unique genotypes.
* `NHACOONGO` is exactly `3 x 7 x 7`.
* The other environments contain one extra plot record per replication, always a duplicated `GEN = 49`.
* Row order reveals a repeated 7-by-7 lattice order, so the incomplete-block labels can be reconstructed as `ceiling(plot_order / 7)` after the duplicate issue is resolved.

::: {.callout-important}
The original fieldbook should be checked for the duplicate `GEN = 49` rows before final analysis. For the design illustration below, `NHACOONGO` is used because it is the clean location with exactly 49 plots in each replication.
:::

## Phenotypic Model

For a single trait such as grain yield, a practical multilocation alpha-lattice model is:

$$y_{hijk} = \mu + E_h + R_{i(h)} + B_{j(hi)} + G_k + (EG)_{hk} + \epsilon_{hijk}$$

where:

* $y_{hijk}$ is the observed phenotype of genotype $k$ in incomplete block $j$, replication $i$, and environment $h$.
* $\mu$ is the overall mean.
* $E_h$ is the environment effect.
* $R_{i(h)}$ is the replication effect nested within environment.
* $B_{j(hi)}$ is the incomplete-block effect nested within replication and environment.
* $G_k$ is the genotype effect.
* $(EG)_{hk}$ is the genotype-by-environment interaction.
* $\epsilon_{hijk}$ is plot error.

For cultivar evaluation, `GEN` and `ENV` may be fitted as fixed effects. For breeding-value prediction, `GEN`, `GEN x ENV`, and incomplete-block terms are often fitted as random effects. For GCA/SCA, the model can be extended once a pedigree key is available:

$$G_k = g_{F_p} + g_{M_q} + s_{pq}$$

for genotype/cross $k = (p,q)$, where $g_{F_p}$ is female GCA, $g_{M_q}$ is male GCA, and $s_{pq}$ is SCA.

# R Setup and Data Import

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(knitr)
library(kableExtra)
library(aplot)

raw_data <- read_csv("data/multilocation.csv", show_col_types = FALSE) %>%
  mutate(
    ENV = factor(ENV),
    REP = factor(REP, levels = c("I", "II", "III")),
    GEN = factor(GEN)
  )
```

# Design Diagnostics

## Basic Dimensions

```{r}
#| label: design-summary

design_summary <- raw_data %>%
  summarize(
    rows = n(),
    environments = n_distinct(ENV),
    replications = n_distinct(REP),
    genotypes = n_distinct(GEN)
  )

kable(design_summary, caption = "Basic dimensions of Stelio's multilocation dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Replication Completeness

```{r}
#| label: replication-completeness

rep_summary <- raw_data %>%
  count(ENV, REP, name = "observed_rows") %>%
  left_join(
    raw_data %>%
      summarize(unique_genotypes = n_distinct(GEN), .by = c(ENV, REP)),
    by = c("ENV", "REP")
  ) %>%
  mutate(
    expected_alpha_lattice_rows = 49,
    extra_rows = observed_rows - expected_alpha_lattice_rows
  )

kable(rep_summary, caption = "Observed rows and unique genotypes per environment-by-replication") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Duplicate Records to Verify

```{r}
#| label: duplicate-records

duplicate_records <- raw_data %>%
  mutate(plot_order = row_number(), .by = c(ENV, REP)) %>%
  group_by(ENV, REP, GEN) %>%
  filter(n() > 1) %>%
  ungroup() %>%
  select(ENV, REP, plot_order, GEN, `Yield t/ha`)

kable(duplicate_records, caption = "Duplicated GEN records that should be checked against the fieldbook") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Reconstructing the Alpha-Lattice Blocks

The CSV does not contain a column named `incomplete_block`. However, for the clean location `NHACOONGO`, the row order forms exactly 49 plots per replication. We can reconstruct the candidate alpha-lattice block as:

$$\text{incomplete block} = \left\lceil \frac{\text{plot order within replication}}{7} \right\rceil$$

```{r}
#| label: reconstruct-nhacoongo

nhacoongo_lattice <- raw_data %>%
  filter(ENV == "NHACOONGO") %>%
  arrange(REP) %>%
  mutate(
    plot_order = row_number(),
    rep_plot = row_number(),
    .by = REP
  ) %>%
  mutate(
    incomplete_block = ceiling(rep_plot / 7),
    plot_in_block = ((rep_plot - 1) %% 7) + 1
  )

nhacoongo_check <- nhacoongo_lattice %>%
  count(REP, incomplete_block, name = "plots") %>%
  summarize(
    incomplete_blocks = n_distinct(incomplete_block),
    min_block_size = min(plots),
    max_block_size = max(plots),
    .by = REP
  )

kable(nhacoongo_check, caption = "NHACOONGO has 7 incomplete blocks of 7 plots in every replication") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
#| label: nhacoongo-layout-table

nhacoongo_table <- nhacoongo_lattice %>%
  transmute(
    REP,
    incomplete_block = paste0("IB", incomplete_block),
    plot_in_block = paste0("Plot ", plot_in_block),
    GEN = as.character(GEN)
  ) %>%
  pivot_wider(names_from = plot_in_block, values_from = GEN)

kable(nhacoongo_table, caption = "Reconstructed alpha-lattice layout for NHACOONGO") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Alpha-Lattice Illustration with `aplot`

The figure below illustrates one location (`NHACOONGO`). Each panel is a replication. Within a panel, rows are incomplete blocks (`IB1` to `IB7`) and columns are the seven plot positions within an incomplete block. The number inside each tile is the genotype code.

```{r}
#| label: alpha-lattice-aplot
#| fig.width: 11
#| fig.height: 7
#| warning: false

layout_plot <- ggplot(
  nhacoongo_lattice,
  aes(x = plot_in_block, y = factor(incomplete_block), fill = `Yield t/ha`)
) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = GEN), size = 3.5, fontface = "bold") +
  facet_wrap(~ REP, nrow = 1) +
  scale_y_discrete(limits = rev) +
  scale_x_continuous(breaks = 1:7, expand = c(0, 0)) +
  scale_fill_viridis_c(option = "C", name = "Yield\nt/ha") +
  labs(
    title = "Alpha-lattice layout for NHACOONGO",
    subtitle = "3 replications; each replication has 7 incomplete blocks of 7 plots",
    x = "Plot position inside incomplete block",
    y = "Incomplete block"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

rep_counts <- nhacoongo_lattice %>%
  count(REP, name = "plots") %>%
  mutate(label = paste0("Rep ", REP, ": ", plots, " plots")) %>%
  ggplot(aes(x = REP, y = plots)) +
  geom_col(fill = "#2b6cb0", width = 0.65) +
  geom_text(aes(label = label), vjust = -0.4, size = 3.4) +
  coord_cartesian(ylim = c(0, 56)) +
  labs(x = NULL, y = "Plots per replication") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

layout_plot %>%
  insert_bottom(rep_counts, height = 0.24)
```

# Analysis Implications

For the real dataset, the experimental-design terms should reflect the alpha lattice:

```r
# Conceptual fixed/random structure
trait ~ ENV + GEN + ENV:GEN + REP:ENV + incomplete_block:REP:ENV
```

In practice:

* Use `ENV`, `REP`, `GEN`, and reconstructed or original incomplete-block labels.
* Prefer the original fieldbook's incomplete-block column if Stelio can provide it.
* Resolve the duplicated `GEN = 49` records before estimating final means, genotype effects, heritability, or GCA/SCA.
* Join a pedigree/cross key to `GEN` before estimating female GCA, male GCA, and SCA.

The corrected theoretical statement is therefore:

> Stelio's real multilocation dataset contains 49 maize genotypes evaluated across four environments with three replications per environment. The field design is best represented as a resolvable alpha-lattice design with seven incomplete blocks of seven plots within each replication. This layout should replace the earlier RCBD assumption used for the simulated dataset.
