---
title: "Stelio's Multilocation Maize Trial as an Alpha-Lattice Design"
subtitle: "Experimental-design notes and alpha-lattice allocation from `data/multilocation.csv`"
author: "Fausto Rodriguez Zapata"
date: "2026-06-02"
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-float: true
    number-sections: true
    code-fold: show
    code-tools: true
---

::: {.callout-tip appearance="simple"}
**Source code & data:** <https://github.com/faustovrz/gcamoz>
:::

::: {.callout-note}
This notebook updates the experimental-design description for Stelio's real file, `data/multilocation.csv`. The older `gca_calculation.qmd` was built from simulated data under an RCBD assumption. The real file is more consistent with a **multilocation alpha-lattice trial**: 49 genotypes evaluated in 3 replications per environment, with each replication allocated to 7 incomplete blocks of 7 experimental units.
:::

# Theoretical Foundation & Experimental Setup

Before fitting a genetic model, we must separate three ideas that are easy to mix together:

1. The **genetic material** being evaluated.
2. The **experimental allocation / field layout** used to control local variation.
3. The **phenotypic model** used to analyze the traits.

The file Stelio sent contains plot-level phenotypes from a multilocation maize trial. It does **not** contain female-parent and male-parent labels, so this notebook focuses on the field-design structure. A GCA/SCA analysis still requires a genotype key that maps each `GEN` code to its female and male parents.

## Breeding Scheme (Line x Tester)

Stelio's research focuses on the genetic architecture of low-phosphorus use efficiency in maize. My current understanding follows the same genetic interpretation used in `docs/gca_calculation.html`, section **1.1.1 Stelio's Factorial Mating Grid**: the 49 genotypes are intended to represent a **Factorial Mating Design**, often referred to as a **Line x Tester** or **North Carolina Design II**.

In this mating scheme, two distinct parental sets are crossed:

* **Female parents / lines:** the CIMMYT CML tropical donor founders.
* **Male parents / testers:** the Mozambique founder/tester lines.
* **Hybrid families:** all possible crosses between the 7 female parents and 7 male parents, giving **49 F1 hybrid combinations**.

This is the genetic design assumption. The field design for Stelio's real file is a separate issue: those 49 genotypes were evaluated using a multilocation alpha-lattice allocation, not the simulated RCBD layout used by the earlier teaching file.

## Genetic Material and Pedigree Metadata

The real data identify entries with the column `GEN`. In this file:

* `GEN` has **49 unique levels**: `1` through `49`.
* These 49 entries are the treatments/genotypes evaluated in the field.
* The file does not include columns such as `female`, `male`, or `cross`, so `GEN` must be joined to an external pedigree/crossing key before line-by-tester GCA and SCA can be estimated.

The **known tropical founder parents** are the CIMMYT CML donor lines already described in the project. The Mozambique tester/founder identifiers are not yet available, so this notebook temporarily keeps the placeholder labels `MOZ1` to `MOZ7`.

::: {.callout-warning}
`MOZ1` to `MOZ7` are provisional identifiers. Stelio should replace these labels with the real Mozambique founder/tester IDs before final GCA/SCA analysis.
:::

| Parent set | Role in factorial design | Current IDs | Status |
| :--- | :--- | :--- | :--- |
| CIMMYT CML donors | Female/tropical donor founders | `CML364`, `CML366`, `CML434`, `CML435`, `CML439`, `CML440`, `CML532` | Known from the project description |
| Mozambique founders/testers | Male local tester founders | `MOZ1`, `MOZ2`, `MOZ3`, `MOZ4`, `MOZ5`, `MOZ6`, `MOZ7` | Placeholder labels until Stelio supplies the real IDs |

## Placeholder Factorial Mating Grid

The table below repeats the factorial mating-grid assumption from the simulated RCBD notebook, but here it is explicitly a **placeholder pedigree map** for the real alpha-lattice dataset. It shows the expected `7 x 7 = 49` hybrid structure. Stelio needs to verify which `GEN` code in `multilocation.csv` corresponds to each cross.

| CIMMYT donor / Mozambique tester | MOZ1 | MOZ2 | MOZ3 | MOZ4 | MOZ5 | MOZ6 | MOZ7 |
| :--- | :---: | :---: | :---: | :---: | :---: | :---: | :---: |
| **CML364** | CML364 x MOZ1 | CML364 x MOZ2 | CML364 x MOZ3 | CML364 x MOZ4 | CML364 x MOZ5 | CML364 x MOZ6 | CML364 x MOZ7 |
| **CML366** | CML366 x MOZ1 | CML366 x MOZ2 | CML366 x MOZ3 | CML366 x MOZ4 | CML366 x MOZ5 | CML366 x MOZ6 | CML366 x MOZ7 |
| **CML434** | CML434 x MOZ1 | CML434 x MOZ2 | CML434 x MOZ3 | CML434 x MOZ4 | CML434 x MOZ5 | CML434 x MOZ6 | CML434 x MOZ7 |
| **CML435** | CML435 x MOZ1 | CML435 x MOZ2 | CML435 x MOZ3 | CML435 x MOZ4 | CML435 x MOZ5 | CML435 x MOZ6 | CML435 x MOZ7 |
| **CML439** | CML439 x MOZ1 | CML439 x MOZ2 | CML439 x MOZ3 | CML439 x MOZ4 | CML439 x MOZ5 | CML439 x MOZ6 | CML439 x MOZ7 |
| **CML440** | CML440 x MOZ1 | CML440 x MOZ2 | CML440 x MOZ3 | CML440 x MOZ4 | CML440 x MOZ5 | CML440 x MOZ6 | CML440 x MOZ7 |
| **CML532** | CML532 x MOZ1 | CML532 x MOZ2 | CML532 x MOZ3 | CML532 x MOZ4 | CML532 x MOZ5 | CML532 x MOZ6 | CML532 x MOZ7 |

For analysis, Stelio should add either a metadata file or columns to the field dataset with this minimum structure:

| Required column | Meaning | Example value |
| :--- | :--- | :--- |
| `GEN` | Genotype code used in `multilocation.csv` | `1` |
| `female` | CIMMYT CML donor parent | `CML364` |
| `male` | Mozambique tester parent | `MOZ1` or real tester ID |
| `cross` | Hybrid/cross identifier | `CML364 x MOZ1` |

This would make the real dataset parallel to the simulated RCBD file, which already had `female`, `male`, and `cross` columns. Until this pedigree metadata is available, the real data can support alpha-lattice field-design diagnostics and genotype-level analysis, but not a defensible final GCA/SCA decomposition.

## Experimental Design

> The alpha-lattice design is the actual pot/plot allocation used to evaluate the 49 genotypes produced by the 7 x 7 line-by-tester crossing design.

Stelio's real experiment appears to use **pots placed under field conditions**, which is common in phosphorus studies because pots make it easier to control phosphorus dose, soil/substrate, root-zone conditions, and biomass/root harvest. In this context, the alpha-lattice design should be interpreted as the **experimental allocation of genotype entries into incomplete blocks and replications**, not necessarily as a conventional row-column field layout.

Stelio's real trial is a **multilocation alpha-lattice design**. This replaces the RCBD field-layout assumption used in the simulated teaching notebook. Stelio's statement that this is an alpha lattice matches the dimensions in the file. With 49 genotypes, the natural resolvable incomplete-block allocation is:

$$49 = 7 \times 7$$

### Locations / Environments

The trial contains four environments/locations, recorded in the `ENV` column:

| Location code in `ENV` | Interpretation in this notebook |
| :--- | :--- |
| `CHOKWE STS` | Chokwe stress environment |
| `CKOKWE OPT` | Chokwe optimum environment, as spelled in the CSV |
| `NHACOONGO` | Nhacoongo environment |
| `SUSSUNDENGA` | Sussundenga environment |

### Alpha-Lattice Allocation Within Each Location

Within each location and replication:

* there are **7 incomplete blocks**;
* each incomplete block contains **7 plots/genotypes**;
* all 49 genotypes appear once per replication;
* the replication is therefore a complete replicate, but it is subdivided into smaller incomplete blocks.

Across the trial:

* **Replications (`REP`)**: `I`, `II`, and `III` within each environment.
* **Genotypes (`GEN`)**: 49 unique entries.
* **Traits**: grain yield, yield per plant, plant number, ear weight, grain weight, plant height, culm diameter, flowering dates, dry matter mass, and root weight.

This is different from an RCBD. In an RCBD, each replication is treated as one large complete block of 49 experimental units. In an alpha lattice, each replication is split into incomplete blocks, so the error model can account for local variation among smaller groups of pots or plots.

## How the Data Fit the Alpha Lattice

The clean alpha-lattice expectation is:

$$4\ \text{environments} \times 3\ \text{replications} \times 7\ \text{incomplete blocks} \times 7\ \text{plots} = 588\ \text{plots}$$

The file has 597 rows, not 588. The reason is specific and diagnosable: in three environments (`CHOKWE STS`, `CKOKWE OPT`, and `SUSSUNDENGA`), each replication has one duplicated record for `GEN = 49`. `NHACOONGO` has the expected 49 rows per replication.

Therefore, the design evidence is:

* The genotype set is exactly 49 entries.
* Every `ENV x REP` combination has 49 unique genotypes.
* `NHACOONGO` is exactly `3 x 7 x 7`.
* The other environments contain one extra plot record per replication, always a duplicated `GEN = 49`.
* Row order is compatible with a repeated 7-by-7 lattice allocation, so candidate incomplete-block labels can be derived as `ceiling(plot_order / 7)` after the duplicate issue is resolved.

::: {.callout-important}
The CSV does **not** contain physical coordinates such as pot row, pot column, range, tier, x/y position, plot ID, or an explicit incomplete-block column. Therefore, the notebook can diagnose the alpha-lattice dimensions and illustrate a candidate allocation, but it cannot reconstruct the true physical placement of pots in the field. The original fieldbook should be checked for the duplicate `GEN = 49` rows and for the actual incomplete-block or pot-position metadata before final analysis.
:::

## Phenotypic Model

For a single trait such as grain yield, a practical multilocation alpha-lattice model is:

$$y_{hijk} = \mu + E_h + R_{i(h)} + B_{j(hi)} + G_k + (EG)_{hk} + \epsilon_{hijk}$$

where:

* $y_{hijk}$ is the observed phenotype of genotype $k$ in incomplete block $j$, replication $i$, and environment $h$.
* $\mu$ is the overall mean.
* $E_h$ is the environment effect.
* $R_{i(h)}$ is the replication effect nested within environment.
* $B_{j(hi)}$ is the incomplete-block effect nested within replication and environment.
* $G_k$ is the genotype effect.
* $(EG)_{hk}$ is the genotype-by-environment interaction.
* $\epsilon_{hijk}$ is plot error.

For cultivar evaluation, `GEN` and `ENV` may be fitted as fixed effects. For breeding-value prediction, `GEN`, `GEN x ENV`, and incomplete-block terms are often fitted as random effects. For GCA/SCA, the model can be extended once a pedigree key is available:

$$G_k = g_{F_p} + g_{M_q} + s_{pq}$$

for genotype/cross $k = (p,q)$, where $g_{F_p}$ is female GCA, $g_{M_q}$ is male GCA, and $s_{pq}$ is SCA.

# R Setup and Data Import

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(knitr)
library(kableExtra)
library(aplot)

raw_data <- read_csv("data/multilocation.csv", show_col_types = FALSE) %>%
  mutate(
    ENV = factor(ENV),
    REP = factor(REP, levels = c("I", "II", "III")),
    GEN = factor(GEN)
  )
```

# Design Diagnostics

## Basic Dimensions

```{r}
#| label: design-summary

design_summary <- raw_data %>%
  summarize(
    rows = n(),
    environments = n_distinct(ENV),
    replications = n_distinct(REP),
    genotypes = n_distinct(GEN)
  )

kable(design_summary, caption = "Basic dimensions of Stelio's multilocation dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Replication Completeness

```{r}
#| label: replication-completeness

rep_summary <- raw_data %>%
  count(ENV, REP, name = "observed_rows") %>%
  left_join(
    raw_data %>%
      summarize(unique_genotypes = n_distinct(GEN), .by = c(ENV, REP)),
    by = c("ENV", "REP")
  ) %>%
  mutate(
    expected_alpha_lattice_rows = 49,
    extra_rows = observed_rows - expected_alpha_lattice_rows
  )

kable(rep_summary, caption = "Observed rows and unique genotypes per environment-by-replication") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

## Duplicate Records to Verify

```{r}
#| label: duplicate-records

duplicate_records <- raw_data %>%
  mutate(plot_order = row_number(), .by = c(ENV, REP)) %>%
  group_by(ENV, REP, GEN) %>%
  filter(n() > 1) %>%
  ungroup() %>%
  select(ENV, REP, plot_order, GEN, `Yield t/ha`)

kable(duplicate_records, caption = "Duplicated GEN records that should be checked against the fieldbook") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Inferring Candidate Alpha-Lattice Blocks

The CSV does not contain a column named `incomplete_block`, nor does it contain physical pot coordinates. However, for the clean location `NHACOONGO`, the row order forms exactly 49 records per replication. If the CSV row order follows the original fieldbook planting/allocation order, a candidate alpha-lattice block can be derived as:

$$\text{incomplete block} = \left\lceil \frac{\text{plot order within replication}}{7} \right\rceil$$

```{r}
#| label: reconstruct-nhacoongo

nhacoongo_lattice <- raw_data %>%
  filter(ENV == "NHACOONGO") %>%
  arrange(REP) %>%
  mutate(
    plot_order = row_number(),
    rep_plot = row_number(),
    .by = REP
  ) %>%
  mutate(
    incomplete_block = ceiling(rep_plot / 7),
    plot_in_block = ((rep_plot - 1) %% 7) + 1
  )

nhacoongo_check <- nhacoongo_lattice %>%
  count(REP, incomplete_block, name = "plots") %>%
  summarize(
    incomplete_blocks = n_distinct(incomplete_block),
    min_block_size = min(plots),
    max_block_size = max(plots),
    .by = REP
  )

kable(nhacoongo_check, caption = "NHACOONGO is compatible with 7 candidate incomplete blocks of 7 records in every replication") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r}
#| label: nhacoongo-layout-table

nhacoongo_table <- nhacoongo_lattice %>%
  transmute(
    REP,
    incomplete_block = paste0("IB", incomplete_block),
    plot_in_block = paste0("Plot ", plot_in_block),
    GEN = as.character(GEN)
  ) %>%
  pivot_wider(names_from = plot_in_block, values_from = GEN)

kable(nhacoongo_table, caption = "Candidate alpha-lattice allocation for NHACOONGO based on CSV row order") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

# Conceptual Alpha-Lattice Illustration with `aplot`

The figure below illustrates the expected alpha-lattice allocation for one clean location (`NHACOONGO`). Each panel is a replication. Within a panel, rows are candidate incomplete blocks (`IB1` to `IB7`) and columns are the seven positions within an incomplete block. The number inside each tile is the genotype code.

::: {.callout-note}
This is a conceptual allocation diagram, not a verified physical map of pot placement. A true pot map would require Stelio's original fieldbook with pot row/column coordinates, range/tier labels, plot IDs, or explicit incomplete-block identifiers.
:::

```{r}
#| label: alpha-lattice-aplot
#| fig.width: 11
#| fig.height: 7
#| warning: false

layout_plot <- ggplot(
  nhacoongo_lattice,
  aes(x = plot_in_block, y = factor(incomplete_block), fill = `Yield t/ha`)
) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = GEN), size = 3.5, fontface = "bold") +
  facet_wrap(~ REP, nrow = 1) +
  scale_y_discrete(limits = rev) +
  scale_x_continuous(breaks = 1:7, expand = c(0, 0)) +
  scale_fill_viridis_c(option = "C", name = "Yield\nt/ha") +
  labs(
    title = "Conceptual alpha-lattice allocation for NHACOONGO",
    subtitle = "3 replications; each replication is compatible with 7 incomplete blocks of 7 records",
    x = "Position within candidate incomplete block",
    y = "Candidate incomplete block"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    legend.position = "right"
  )

rep_counts <- nhacoongo_lattice %>%
  count(REP, name = "plots") %>%
  mutate(label = paste0("Rep ", REP, ": ", plots, " plots")) %>%
  ggplot(aes(x = REP, y = plots)) +
  geom_col(fill = "#2b6cb0", width = 0.65) +
  geom_text(aes(label = label), vjust = -0.4, size = 3.4) +
  coord_cartesian(ylim = c(0, 56)) +
  labs(x = NULL, y = "Plots per replication") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

layout_plot %>%
  insert_bottom(rep_counts, height = 0.24)
```

# Analysis Implications

For the real dataset, the experimental-design terms should reflect the alpha lattice:

```r
# Conceptual fixed/random structure
trait ~ ENV + GEN + ENV:GEN + REP:ENV + incomplete_block:REP:ENV
```

In practice:

* Use `ENV`, `REP`, `GEN`, and the original incomplete-block labels if Stelio can provide them.
* Treat row-order-derived incomplete-block labels as provisional unless the original fieldbook confirms that CSV row order matches the alpha-lattice allocation order.
* Request pot row/column coordinates, range/tier labels, plot IDs, or explicit incomplete-block identifiers if physical spatial analysis is needed.
* Resolve the duplicated `GEN = 49` records before estimating final means, genotype effects, heritability, or GCA/SCA.
* Join a pedigree/cross key to `GEN` before estimating female GCA, male GCA, and SCA.

The corrected theoretical statement is therefore:

> Stelio's real multilocation dataset contains 49 maize genotypes evaluated across four environments with three replications per environment. The experimental allocation is best represented as a resolvable alpha-lattice design with seven incomplete blocks of seven experimental units within each replication. This should replace the earlier RCBD assumption used for the simulated dataset, while recognizing that the CSV does not contain enough spatial metadata to reconstruct the true physical pot layout.
