---
title: "Phosphorus-Use Efficiency (PUE): a distribution check"
subtitle: "Is PUE one long-tailed/Gaussian distribution, or a bimodal mixture of the two phosphorus treatments? With PUE performance across environments in beeswarm + boxplot style."
author: "Fausto Rodriguez Zapata"
date: "2026-06-29"
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
---

# The question

`PUE` (phosphorus-use efficiency) in `data/multilocation.csv` looked suspicious — a very
flat, high-variance spread that could be a long tail or a wide Gaussian. The hypothesis
(Stelio) is that it is really **bimodal**, one mode per **phosphorus condition**: P-stress
vs. no stress.

That is exactly what the data say. PUE is computed as grain yield per unit applied
phosphorus,

$$\text{PUE} = \frac{\text{GY (kg/ha)}}{\text{P input (kg/ha)}},$$

so the applied P is recoverable as `GY * 1000 / PUE`. It collapses to **two discrete values
with zero within-location variance** — `50 kg/ha` (optimal) at four environments and
`10 kg/ha` (stress) at CHOKWE STS — which is carried in the `p_input` and `treatment`
columns. The small stress denominator inflates PUE, so the "second mode" is simply the
stress site.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidyverse)
library(ggbeeswarm)   # geom_quasirandom — beeswarm points
library(knitr)
library(kableExtra)
```

```{r}
#| label: data

hybrids <- read_csv("data/multilocation.csv", show_col_types = FALSE) %>%
  filter(gen != 0) %>%                       # drop the NAMULONGUE check
  mutate(
    env       = factor(env),
    treatment = factor(treatment, levels = c("Optimal", "Stress"))
  )

# Shared style: base_size 14, transparent boxplots, beeswarm points (the inv4m look,
# built on base ggplot2 theme_classic() — no ggpubr needed).
pue_theme <- theme_classic(base_size = 14) +
  theme(
    plot.title      = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.margin     = margin(t = 20, r = 8, b = 5, l = 5, unit = "pt"),
    axis.title.y    = element_text(size = 12),
    axis.title.x    = element_blank(),
    axis.text.x     = element_text(size = 11, face = "bold", color = "black",
                                   angle = 30, hjust = 1),
    axis.text.y     = element_text(size = 10),
    legend.position = "top"
  )

trt_pal <- c("Optimal" = "#2c7fb8", "Stress" = "#d95f0e")
```

# Performance across Environments

PUE for the 49 line × tester hybrids in each environment, drawn in the **beeswarm +
transparent-boxplot** style: the data points sit in the background (`geom_quasirandom`) with
a transparent boxplot in front (`alpha = 0`), coloured by phosphorus **treatment**. The
single stress environment (CHOKWE STS, orange) stands far above the four optimal sites — the
visual signature of the bimodality.

```{r}
#| label: pue-across-env
#| fig-width: 10
#| fig-height: 6
#| warning: false

ggplot(hybrids, aes(x = env, y = PUE, color = treatment)) +
  geom_quasirandom(size = 1.1, alpha = 0.55, width = 0.32) +          # points: background
  geom_boxplot(width = 0.25, linewidth = 1.0, alpha = 0,              # transparent box: front
               outlier.shape = NA) +
  scale_color_manual(values = trt_pal, name = NULL) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.12))) +
  labs(title = "PUE across the five environments",
       y = expression("PUE  (g grain " * g^{-1} * " P)")) +
  pue_theme
```

# Is PUE bimodal?

Pooled across environments, PUE is clearly **two humps**, not one. Splitting by `treatment`
shows each treatment is its own well-separated, roughly unimodal cluster — so the overall
bimodality is **entirely** the optimal-vs-stress contrast, nothing within a P regime.

```{r}
#| label: pue-hist
#| fig-width: 9
#| fig-height: 4.5
#| warning: false

ggplot(hybrids, aes(x = PUE, fill = treatment)) +
  geom_histogram(binwidth = 50, color = "white", alpha = 0.85,
                 position = "identity") +
  scale_fill_manual(values = trt_pal, name = NULL) +
  labs(title = "PUE is bimodal — the two modes are the two P treatments",
       x = expression("PUE  (g grain " * g^{-1} * " P)"), y = "Plots") +
  theme_classic(base_size = 13) +
  theme(legend.position = "top",
        plot.title = element_text(face = "bold", hjust = 0.5))
```

```{r}
#| label: pue-by-treatment

hybrids %>%
  group_by(Treatment = treatment, `P input (kg/ha)` = p_input) %>%
  summarise(n = n(),
            `mean PUE` = round(mean(PUE), 0),
            `sd`       = round(sd(PUE), 0),
            min        = round(min(PUE), 0),
            max        = round(max(PUE), 0),
            .groups = "drop") %>%
  kable(caption = "PUE summary by phosphorus treatment. The stress site's 10 kg/ha denominator (vs 50) is what lifts its PUE into the upper mode.") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

**Reading.** PUE is not a single skewed/Gaussian distribution — it is a **mixture of two P
treatments**. PUE (and the other efficiency ratios) are therefore only comparable *within*
a phosphorus regime; pooling optimal and stress plots is what produced the flat,
high-variance "distribution" that looked wrong. For any model, the P treatment belongs in
the design (here it is absorbed by the fixed `env` term, since `CHOKWE OPT` and `CHOKWE STS`
are distinct environments).
