---
title: "F₂ QTL population — data provenance, marker QC and ABH encoding"
subtitle: "Permanent record of the pre-mapping pipeline (DArTseq order DMz26-3123)"
date: 2026-07-30
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    df-print: kable
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(data.table)
D <- "data/qtl/derived"
have <- function(p) file.exists(p)
```

::: {.callout-important}
## Scope and status

This notebook is the **record of everything done before QTL mapping begins** — data
inventory, population verification, marker QC, the AGPv4 → NAM v5 liftover, parental
allele calling against the CML530 assembly, and the A/H/B encoding for R/qtl.

**No QTL mapping is performed here.** No genetic map has been estimated yet: the R/qtl
object carries placeholder positions and `est.map` has not been run.

The raw DArTseq reports and the phenotype workbook are **not redistributable** and are
gitignored (`data/qtl/`). This notebook and `scripts/qtl/` are the tracked record; the
data itself must be obtained from Stélio Nuvunga.
:::

# Provenance

| | |
|---|---|
Order | `DMz26-3123` |
Vendor | SEQART AFRICA (AICAD Complex, JKUAT Juja, Kenya) |
Client | Stélio Boaventura Nuvunga (Makerere University, PASET-RSIF) |
Product | maize DArTseq (GBS) 1.0 |
Order date | 2026-03-11 |
Alignment reference | B73 **AGPv4**, blast ≥90 % identity, e-value 5e-10 |
Samples | 186, two 96-well plates |

Three report files, all carrying the same 186 samples:

| File | Content | Used |
|---|---|---|
`Report_..._SNP_mapping_2.csv` | 5,208 SNPs, one-row format `0`/`1`/`2`/`-` | **yes** |
`Report_..._SNP_2.csv` | same 5,208 SNPs, two-row per-allele format | no (redundant) |
`Report_..._SilicoDArT_1.csv` | 11,518 dominant presence/absence markers | no |

The SilicoDArT markers are **dominant** — they cannot separate a heterozygote from one
homozygote, so they carry no dominance information and are unusable for an F₂. They were
excluded, not filtered.

The CML530 reference assembly (`Zm-CML530-REFERENCE-HiLo-1.0`, MaizeGDB, released June
2024) is distributed under the **Toronto Agreement**, which restricts publishing
genome-wide analyses ahead of the assembly producers. Relevant to manuscript timing.

# Population

## It is a single biparental F₂

Stélio's project description refers to F₂ populations from *all* 49 combinations
(plural), which would invalidate the R/qtl `f2` cross type if the 186 plants were pooled
across families. Three independent tests say they are not
(`scripts/qtl/qtl_structure_check.R`):

**Allele frequency.** In a single selfed F₁ every informative marker must sit at
*p* ≈ 0.5. Observed SD of *p*(ref) across 1,398 framework markers = **0.028**, against a
pure binomial expectation at *n* = 186 of **0.0259** — a ratio of 1.08, i.e. sampling
noise and nothing else. **Zero** markers fall outside *p* = 0.40–0.60.

**PCA.** PC1 explains **7.9 %**, decaying smoothly (7.9, 7.0, 6.5, 6.1, 5.7 …), with
PC1~sd~/PC2~sd~ = 1.07. No dominant axis, no clusters. *k* pooled families would produce
*k*−1 large separated components.

**Relatedness.** Median pairwise IBS **0.376** — the textbook F₂ full-sib value. No pair
above 0.90, so no duplicated samples. No plate effect (*p* = 0.75).

## No parent was genotyped in this order

The parents are inbred lines, so heterozygosity must be ≈ 0. Over the 1,398 informative
markers the **minimum** observed heterozygosity across all 186 samples is **0.267**, and
the median is **0.5016** — dead on the F₂ expectation of 0.50. Not one sample is
consistent with an inbred. Combined with the sample naming (`plant 1` … `plant 188`, no
line names) and uniform `none [Original species: Zea maize]` comments, no parental
sample was submitted.

Incidentally, that median of 0.5016 also shows **allelic dropout is not distorting the
retained marker set**, despite modest read depth (median 6.9× ref / 5.5× alt). The
low-`FreqHets` tail is uninformative markers where the parents share the allele, not a
dropout artefact.

## Parents

From `README.md` and `multilocation_alpha_lattice.qmd`, the NC Design II female set is
**CML364, CML366, CML434, CML435, CML439, CML530, CML532** and the male testers are
**EN17, EN20, EN21, EN25, EN31, EN32, EN64** (IIAM IDs; legacy `MOZL3`–`MOZL10`).

`multilocation_alpha_lattice.qmd` records explicitly that *"the earlier draft had
incorrectly listed CML440"* — so **CML530 is a female parent of this programme**, and an
earlier parent list circulated with CML440 in its place.

::: {.callout-warning}
## Assumed, not confirmed
That this specific F₂ derives from a **CML530 × ENxx** cross, and **which** tester.
`paper/maize_pue_manuscript.md` lists CML530 × EN31 among the most stable high-yielding
hybrids by WAASY, but stability ranking is not a phosphorus contrast — that is a lead,
not evidence. **Confirm with Stélio before publishing allele-effect directions.**
:::

# Phenotypes

`data/qtl/mozpue_phenotype.xlsx`, sheet `data base`. Row 1 is a **units** row, not data.
188 body rows, **168 with data**; 20 IDs entirely blank
(3, 11, 19, 21, 39, 47, 61, 74, 110, 116, 127, 137, 138, 156, 158, 160, 161, 169, 175, 177).

Trait names in the source file are free-text Excel headers. They are mapped to the
project abbreviations through **`data/phenotype_dictionary.csv`**, which is the single
authority (built by `scripts/qtl/phenotype_dictionary_unify.R`) and carries a
`raw_header` column so no downstream script duplicates the mapping.

```{r dict}
if (have("data/phenotype_dictionary.csv")) {
  dd <- fread("data/phenotype_dictionary.csv")[project == "qtl_f2" & role == "trait"]
  dd[, .(variable, name, unit, formula, raw_header)][order(variable)]
}
```

## Derived formulas, recovered numerically

None of the derived columns were documented. All were reproduced from the data to
floating-point precision (`scripts/qtl/phenotype_dictionary_unify.R`):

| Trait | Formula | max&#124;diff&#124; |
|---|---|---|
`GY` | `GYPP × 0.0533333` | 5.0e-06 |
**`PUE`** | **`GY(kg/ha) / 20`** | 1.4e-14 |
`HI` | `GYPP / SW` | 0 |
`RSR` | `RW / SW` | 0 |
`RYE` | `GYPP / RW` | 0 |
`TDM` | `SW + RW` | 0 |

::: {.callout-note}
## The phosphorus rate was recoverable from the arithmetic
`PUE = GY(kg/ha)/p_input` with the implied `p_input` having **sd = 2.8e-16** across all
168 plants ⇒ the F₂ trial received **20 kg P/ha**. This is neither of the multilocation
rates (10 stress / 50 optimal), so the F₂ trial is its own phosphorus regime and is
**not** directly comparable to the diallel's stress treatment. Worth confirming with
Stélio, but the value is unambiguous.

The `GY` conversion implies ~53,333 plants/ha (0.1875 m²/plant).
:::

::: {.callout-warning}
## Three cautions that affect interpretation

1. **`HI` here is not harvest index.** `HI = GYPP/SW`, and `TDM = SW + RW` confirms `SW`
   excludes the ear. So this is grain-per-shoot, not grain ÷ total above-ground biomass.
   Reporting it as harvest index needs a definitional footnote.
2. **`HI`, `RSR`, `RYE` are algebraically linked** — three ratios over the same three
   measurements, with `RYE/HI = 1/RSR`. They *will* share QTL by construction. Any
   co-localization claim among them is arithmetic, not pleiotropy. Decide which traits
   are primary before interpreting.
3. **`PAPP` is an ordinal 1–4 score.** `scanone` assumes a roughly normal continuous
   phenotype; use `model = "np"` for it.
:::

15 plants are flagged `Higher M.C` (moisture content): 65, 72, 86, 95, 111, 121, 125,
128, 136, 139, 140, 153, 163, 172, 178. **Grain yield is not standardised to a common
moisture basis.**

`HI` and `RYE` in the *multilocation* dataset could **not** be reproduced by any of 10
and 7 candidate formulas respectively (closest miss 0.643 and 15.6, far beyond rounding).
Both appear in the manuscript. That is a provenance gap in the diallel data, recorded
here because it shares abbreviations with this dataset.

# Effective population size

```{r join}
data.table(
  quantity = c("genotyped", "phenotyped", "**both (mapping population)**",
               "genotyped only", "phenotyped only"),
  n = c(186, 168, 166, 20, 2))
```

**n = 166.** For an F₂ at a 5 % genome-wide threshold this gives roughly 80 % power for
QTL explaining ≥ 8–10 % of phenotypic variance. Smaller effects will be missed and
detected effects will be **Beavis-inflated**. There is **no within-genotype
replication**, so no heritability can be estimated and genetic variance cannot be
separated from micro-environmental variance.

This was a deliberate design choice, not an oversight: in July 2025 Joe Gage set out both
the F₂:₃ route (row-level BLUPs, replicable) and direct F₂ phenotyping, noting the latter
*"will likely have more error in it, since the F₂ plants cannot be replicated."* The
faster route was taken.

::: {.callout-tip}
## Largest available improvement
If F₃ seed was collected from these 186 selfed F₂ plants, F₂:₃ rows can be planted and
the **same genotypes** re-mapped against replicated BLUPs — one season, no new
genotyping. Worth asking Stélio.
:::

# Marker QC

`scripts/qtl/qtl_data_audit.R`. The 1:2:1 test stands in for "polymorphic between the
parents", which is otherwise unobservable because no parent was genotyped.

```{r mkqc}
data.table(
  step = c("SNPs in report", "anchored to AGPv4 chr 1–10",
           "effectively monomorphic (MAF < 0.05)",
           "+ CallRate ≥ 0.90, RepAvg ≥ 0.95, MAF ≥ 0.15, 1:2:1 p > 0.01",
           "− DArT secondaries (>1 SNP per 69-bp tag)"),
  n = c(5208, 3884, 2041, 1410, 1398))
```

Per-sample call rate 0.64–0.97 (median 0.936); only **plant 14** below 0.80.

::: {.callout-note}
## A caveat on the 1:2:1 filter
A paralogous DArT tag collapsing two loci produces apparent heterozygotes at high
frequency and **can pass a 1:2:1 test** while not being a single locus. F₁-het
ascertainment (as used in the airmine population) is direct evidence that parents differ;
a 1:2:1 fit is only statistical. Two independent handles on this exist: the `find_quirky`
map-geometry rule (see *Next steps*), and the CML530 `het-call` class below.
:::

# Liftover AGPv4 → NAM v5

`scripts/qtl/qtl_liftover_v4_to_v5.R`, reusing `lift_unique()` from
`zealhmm/R/teonam_liftover.R` verbatim — single hop, unique 1:1 mappings only.

```{r lift}
data.table(
  quantity = c("submitted (v4-anchored, chr 1–10)", "lifted 1:1", "failed / multi-mapping",
               "lifted onto non-chr1-10 scaffold", "chromosome changers",
               "framework markers surviving"),
  n = c("3,884", "3,828 (98.6 %)", "56", "0", "0", "1,384 of 1,398 (99.0 %)"))
```

**v4 → v5 is far cleaner than v2 → v5.** Zero chromosome changers and zero off-chromosome
scaffolds, so the chr7-displaced-block class of artefact that motivated `find_quirky`'s
island rule does not appear. Per-chromosome Spearman ρ between v4 and v5 position is
≥ 0.988 everywhere (chr5, 8, 9 exactly 1.00000). 715 markers change local rank —
fine-scale reshuffling within an otherwise collinear order.

Two things to watch: **chr3 has the lowest collinearity** (ρ = 0.98839), and **chr6's
marker span shrank 170.8 → 146.3 Mb** while losing only 3 markers.

::: {.callout-caution}
## A silent-failure class encountered twice
The chain file names chromosomes `1`…`10`; both FASTAs on disk name them
`chr1`…`chr10`. Without a prefix, `bcftools mpileup -T` matches **nothing** and returns an
empty result *without erroring*. Separately, a third leading column in an R/qtl `csvsr`
genotype file is parsed as **an extra individual** whose values become unrecognised
genotype codes silently converted to missing (verified in
`scripts/qtl/qtl_rqtl_format_probe.R`). Both produce a wrong answer with a clean exit.
:::

# Parental alleles from the CML530 assembly

`scripts/qtl/qtl_cml530_genotypes.sh`. Because CML530 **is** a parent, this yields actual
parental alleles and ABH polarity is directly determined rather than proxied.

The join key is the **B73 v5 coordinate**, not a read name. Two mapping steps, because
CML530 sequence must be obtained before it can be placed in v5:

1. `minibwa map` the 3,828 tags → CML530 assembly — locates each tag's homologous locus.
2. Extract the **aligned span only** (no flanking sequence — the tag is the assay), map
   → B73 v5.
3. `bcftools mpileup -T markers_v5.bed` at the 3,828 known v5 positions.

The genotype is taken positionally from the pileup rather than by walking the SNP offset
by hand: if CML530 carries an indel near the site, the CIGAR keeps the base assignment
honest where offset arithmetic would silently shift it.

## Attrition, reconciled

`scripts/qtl/qtl_cml530_attrition.R` derives every stage from the BAMs and **asserts**
that the losses sum to the observed `no-hit` total.

```{r attrition}
data.table(
  stage = c("no CML530 locus (unmapped, flag 0x4)", "secondary / supplementary",
            "mapped to CML530 but MAPQ < 20", "span failed to map back to B73 v5",
            "mapped to v5 but no pileup record at the marker position",
            "**total losses**",
            "less markers sharing a v5 position (one VCF record serves two)",
            "**= no-hit**"),
  n = c(88, 0, 563, 10, 52, 713, -5, 708))
```

::: {.callout-important}
## The dominant loss is the MAPQ filter, not absence of a CML530 locus
563 versus 88. Only 88 of the 708 `no-hit` markers are genuine mapping failures — the
rest are ambiguous placements, which is a **mappability** statement, not presence/absence
variation. The flat per-chromosome `no-hit` distribution therefore mostly tracks repeat
content. MAPQ 20 is a conventional default, not a considered choice; the MAPQ histogram
was strongly bimodal (2,591 of 3,740 at exactly 60), so a lower threshold would recover
part of the 563.
:::

## Result

```{r cmlclass}
if (have(file.path(D, "CML530_marker_alleles.tsv"))) {
  al <- fread(file.path(D, "CML530_marker_alleles.tsv"),
              colClasses = list(character = "chr_v5"))
  merge(al[, .(all_markers = .N), by = cml_class],
        al[framework == TRUE, .(framework = .N), by = cml_class],
        by = "cml_class", all = TRUE)[order(-all_markers)]
} else {
  data.table(cml_class = c("ref", "alt", "no-hit", "het-call"),
             all_markers = c(2086, 1033, 708, 1), framework = c(698, 533, 152, 1))
}
```

**1,232 of 1,384 framework markers (89.0 %) carry real parental polarity.** The
`ref`/`alt` split confirms that CML530 carries the B73 allele at some sites and the
alternate at others with no global pattern — polarity **must** be assigned per marker.
`ref` means CML530 matches B73, because DArT defined the reference allele by its AGPv4
alignment; it says nothing about the ENxx tester.

Exactly **one `het-call`** across 3,828 tags. An inbred cannot be truly heterozygous, so
that is the paralog / collapsed-repeat signal — and finding one says the tag alignments
are behaving cleanly. It is treated as unresolved, never as a genotype.

# A/H/B encoding for R/qtl

`scripts/qtl/qtl_make_abh.R`, a reimplementation of TASSEL 5 `GenosToABHPlugin`
(algorithm read from source). The TASSEL-equivalent core operates on diploid genotype
strings — the same abstraction TASSEL uses — so the parity harness exercises the code
that actually runs.

Rules copied from TASSEL: reject a site if the parents are identical, either is unknown,
or either is heterozygous; match `H` against either phase; exact diploid equality tested
in order A, B, H, else missing.

Two deliberate divergences:

1. **Missing data and off-parent third alleles are kept distinct.** TASSEL collapses both
   into `NA` via its fall-through.
2. **Unpolarizable markers are retained, not dropped.** TASSEL deletes rejected sites.
   Recombination fractions are **orientation-invariant**, so a marker with no CML530 call
   still carries full mapping information under an arbitrary but internally consistent
   orientation — only the *sign* of its allele effect is unavailable. `polarity_source` in
   `abh_marker_info.tsv` records `CML530` or `arbitrary` per marker, so the assumption
   travels with the data.

```{r abh}
data.table(
  quantity = c("individuals", "markers", "chromosomes", "traits",
               "percent genotyped", "A", "H", "B", "missing", "off-parent calls"),
  value = c("166", "1,384", "10", "14", "99.6 %",
            "25.0 %", "50.3 %", "24.7 %", "0.4 %", "0"))
```

A/H/B at 25.0 / 50.3 / 24.7 against the F₂ expectation of 25/50/25, with **zero**
off-parent calls, is an end-to-end check on the whole chain — DArT calls → filter →
liftover → polarity → encoding. `read.cross(format = "csvsr", crosstype = "f2")`
round-trips, asserted inside the script.

# Reproduction

Software: **R 4.5.2** (`qtl`, `data.table`, `readxl`, `rtracklayer`, `GenomicRanges`,
`Biostrings`), **minibwa 0.6-r416**, **samtools/bcftools 1.23.1**, UCSC `liftOver`.

::: {.callout-warning}
## minibwa must be built from source
`brew install minibwa` is built with `gpl=0`, which omits the low-memory BWT builder:
`ERROR: option -l not compiled as it depends on GPL'd code`. The flag still appears in
`--help`, so help text is not evidence it works. Without `-l`, indexing needs 18 N ≈ 39 GB
for a 2.2 Gb genome. Use `scripts/qtl/install_minibwa_from_source.sh`.
:::

Run order:

| # | Script | Notes |
|---|---|---|
1 | `phenotype_dictionary_unify.R` | writes `data/phenotype_dictionary.csv` |
2 | `qtl_data_audit.R` | inventory, marker + sample QC, geno × pheno join |
3 | `qtl_structure_check.R` | F₂ verification, parent absence, duplicates |
4 | `qtl_liftover_v4_to_v5.R` | writes `markers_v5.tsv`, `tags_for_cml530.fa` |
5 | `install_minibwa_from_source.sh` | once |
6 | `qtl_index_genomes.sh` | **~45 min**, ~9.5 GB, peak 7.2 GB RSS, single-threaded |
7 | `qtl_cml530_genotypes.sh` | minutes; writes `CML530_marker_alleles.tsv` |
8 | `qtl_cml530_attrition.R` | reconciles the funnel, asserts the arithmetic |
9 | `qtl_make_abh.R` | writes `rqtl_gen_abh.csv`, `rqtl_phe.csv` |

Supporting: `qtl_minibwa_index_bench.sh`, `qtl_minibwa_index_threads_bench.sh` (why `-l`
and why local rather than HPC), `qtl_rqtl_format_probe.R` (csvsr layout),
`qtl_index_wait_verify.sh` (artefact-based index validation).

::: {.callout-note}
## Two implementation traps recorded
**minibwa aborts on duplicate FASTA names** — `Assertion failed: (i - j <= 2),
map-main.c:144`. It reads consecutive same-named records as mate pairs and hard-aborts on
a third. `samtools faidx -r` names records by region and several tags share a CML530 span,
so spans are renamed to their unique `AlleleID`.

**Recursive `grep` here is `ugrep` and honours `.gitignore`**, so searches from the repo
root silently skip `agent/` and `data/qtl/`. Use explicit paths for those.
:::

# Next steps

Not yet done. The map has **not** been estimated; the R/qtl object carries placeholder
positions.

1. **Build the map on the fixed B73 physical order.** With 166 individuals, de novo
   ordering of 1,384 markers is under-determined and would be worse than the physical
   order. Use `est.rf` to flag markers whose recombination pattern *contradicts* the
   physical order rather than reordering wholesale.
2. **Apply the two-round `est.map` bootstrap** from `zealhmm/scripts/teonam_qtl_permap.R`:
   segregation-distortion drop → preliminary map → `find_quirky` island drop → refined
   map. Adapt the expected genotype frequencies from BC1S4 `(0.734, 0.031, 0.234)` to F₂
   `(0.25, 0.50, 0.25)`. Use `error.prob ≈ 0.01`, not the 1e-4 default.
   Note the distortion filter flags a fixed ~2.5 % upper tail by construction, so a low
   drop rate is **not** validation — total map length is.
3. **Sanity check:** total map length ~1,400–1,800 cM for maize. Much longer means
   residual genotyping error is inflating distances.
4. **Then** scans: `scanone` + 1,000-permutation thresholds, `stepwiseqtl`/`fitqtl` for
   PVE and additive/dominance effects, support intervals for co-localization, and a v5
   candidate-gene screen.

## Outstanding questions for Stélio

1. **Which cross** produced these 186 F₂s — which CML × which ENxx tester — and whether
   one F₁ was selfed or several.
2. **Field book:** design, plot coordinates, planting/harvest dates, soil test P. (The
   applied rate is already recovered: 20 kg P/ha.) Plot coordinates would allow a spatial
   trend to be fitted, partly compensating for the absent replication.
3. **Was F₃ seed collected** from the selfed F₂ plants?
4. **Plate submission sheet** — confirm `plant N` ↔ field-book plant N. An off-by-one in
   plate layout would silently destroy the analysis and nothing currently documents the
   correspondence.
5. **`HI` and `RYE` definitions used in the diallel** — neither is reproducible from
   `multilocation.csv`, and both appear in the manuscript.
6. **`PAPP` and `EH`** — new abbreviations introduced here; his field-book names take
   precedence.
7. From SEQART: the **failed-sample list** (`targetslist` has 215 IDs, the report has 186).
