---
title: "F₂ Genetic Map"
subtitle: "QTL series notebook 2 — the linkage map: 1,152 markers, 1,858 cM, CML530 coordinates"
author: "Fausto Rodriguez Zapata"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
---

::: {.callout-tip appearance="simple"}
**QTL series — notebook 2.** Follows [marker QC and ABH encoding](qtl_marker_pipeline.html).
The full protocol with every parameter and its basis is in `scripts/qtl/PROTOCOL.md`.
:::

::: {.callout-note}
**This notebook reads only public data.** The raw DArTseq genotypes are not
redistributable and are gitignored, but the map itself — marker names, linkage group,
genetic position and both coordinate systems — is published as
[`data/f2_genetic_map.tsv`](https://github.com/faustovrz/gcamoz/blob/main/data/f2_genetic_map.tsv).
Everything below is reproduced from that file alone.
:::

# What this map is

A single biparental F₂ (186 genotyped plants, DArTseq order `DMz26-3123`) from the
7 × 7 line × tester programme. **CML530 is the female parent** — established by the
design, not assumed. The male is one of the seven `EN` testers; which one is still
unknown and matters only for the sign of allele effects.

**Positions are CML530's own**, from `Zm-CML530-REFERENCE-HiLo-1.0`. B73 is not a parent
of this cross, and the two assemblies disagree about which chromosome carries a marker
for 37 of them. Because `est.map` estimates distances along a *fixed* order and never
reorders, a wrong order cannot be recovered downstream — so the order comes from a real
parent of the cross. B73 v5 coordinates are carried alongside for cross-referencing.

```{r setup}
#| message: false
#| warning: false
library(data.table)
library(ggplot2)
library(knitr)
map <- fread("data/f2_genetic_map.tsv")
map[, lg := factor(lg, levels = 1:10)]
theme_set(theme_classic(base_size = 12))
```

# Summary

```{r summary}
#| label: tbl-summary
CML_LEN <- c(`1` = 312749524, `2` = 253527201, `3` = 248930463, `4` = 258114092,
             `5` = 226836904, `6` = 187680145, `7` = 184149065, `8` = 188635608,
             `9` = 166200635, `10` = 153755684)
s <- map[, .(markers = .N,
             cM = round(max(cM), 1),
             mean_gap_cM = round(max(cM) / (.N - 1), 3),
             first_Mb = round(min(cml_pos) / 1e6, 1),
             last_Mb  = round(max(cml_pos) / 1e6, 1)), by = lg]
s[, chr_Mb := round(CML_LEN[as.character(lg)] / 1e6, 1)]
s[, span_Mb := round(last_Mb - first_Mb, 1)]
s[, pct_covered := round(100 * span_Mb / chr_Mb, 1)]
setorder(s, lg)
kable(s, caption = "The map by linkage group. Coverage is the span of mapped markers against the CML530 chromosome length; lost ends are regions where a QTL cannot be found.")
```

```{r totals}
cat(sprintf("total: %d markers, %.1f cM, %.1f%% physical coverage (%.0f of %.0f Mb)\n",
            nrow(map), sum(s$cM), 100 * sum(s$span_Mb) / sum(s$chr_Mb),
            sum(s$span_Mb), sum(s$chr_Mb)))
```

# Ideogram

Each horizontal line is one marker. Solid black, no transparency — this map is sparse
enough that individual markers should be readable rather than blended into a density.

```{r ideogram}
#| label: fig-ideogram
#| fig-width: 9
#| fig-height: 7
#| fig-cap: "Marker positions on the ten linkage groups. Every marker is drawn as an opaque black tick."
W <- 0.26   # half-width of the bar, in x units
map[, xn := as.numeric(lg)]
s[,   xn := as.numeric(lg)]
ggplot() +
  geom_segment(data = s, aes(x = xn, xend = xn, y = 0, yend = cM),
               linewidth = 9, colour = "grey88", lineend = "butt") +
  geom_segment(data = map, aes(x = xn - W, xend = xn + W, y = cM, yend = cM),
               linewidth = 0.35, colour = "black") +
  scale_x_continuous(breaks = 1:10, labels = 1:10) +
  scale_y_reverse(expand = expansion(mult = c(0.02, 0.04))) +
  labs(x = "linkage group (CML530 chromosome)", y = "position (cM)",
       title = sprintf("F2 genetic map - %d markers, %.1f cM", nrow(map), sum(s$cM)))
```

# Marey map

Genetic against physical position. The line must only ever rise: a reversal would mean
two markers are in the wrong order. All ten are monotonic, and all show the classic
maize sigmoid — steep near the chromosome ends where recombination is concentrated, flat
through the pericentromeric middle where tens of megabases accumulate almost no cM.

```{r marey}
#| label: fig-marey
#| fig-width: 12
#| fig-height: 6
#| fig-cap: "Marey map per linkage group. Monotonic throughout, confirming the marker order; the flat middles are pericentromeric recombination suppression."
map[, Mb := cml_pos / 1e6]
ggplot(map[order(lg, Mb)], aes(Mb, cM)) +
  geom_line(colour = "grey65", linewidth = 0.4) +
  geom_point(size = 0.7, colour = "black") +
  facet_wrap(~lg, scales = "free", ncol = 5) +
  labs(x = "physical position (Mb, CML530)", y = "genetic position (cM)",
       title = "Marey map") +
  theme_bw(base_size = 11)
```

# Marker spacing

```{r gaps}
#| label: fig-gaps
#| fig-width: 9
#| fig-height: 3.4
#| fig-cap: "Distribution of inter-marker genetic distances. The single interval above 25 cM is on chr10."
g <- map[order(lg, cM), .(gap = diff(cM)), by = lg]
kable(data.table(
  statistic = c("intervals", "median cM", "mean cM", "q95 cM", "max cM",
                "intervals > 10 cM", "intervals > 25 cM"),
  value = c(nrow(g), round(median(g$gap), 3), round(mean(g$gap), 3),
            round(quantile(g$gap, .95), 3), round(max(g$gap), 2),
            sum(g$gap > 10), sum(g$gap > 25))),
  caption = "Inter-marker spacing.")

ggplot(g[gap < 10], aes(gap)) +
  geom_histogram(binwidth = 0.25, fill = "grey25", boundary = 0) +
  labs(x = "inter-marker distance (cM), truncated at 10", y = "intervals",
       title = "Marker spacing")
```

# What this map cannot do

- **It is ~20% too long** — 1,858 cM against a 1,348–1,596 cM expectation for maize. That
  inflates cM positions and the *cM width* of support intervals. It does **not** affect
  physical (Mb) intervals, which is what a candidate-gene search uses.
- **chr6 covers 78.4%**, losing 34.3 Mb off the start. A QTL there cannot be found.
- **Resolution is limited by the population, not the map.** 372 meioses set a 0.27 cM
  quantum, and the map already resolves far finer than an F₂ of this size can exploit:
  QTL localisation precision is on the order of 10–20 cM for realistic effect sizes. At
  one marker per 1.9 Mb the map is well past saturation for this design.

# Data availability

`data/f2_genetic_map.tsv` — 1,152 markers.

| column | meaning |
|---|---|
| `marker` | DArT `AlleleID` |
| `lg` | linkage group (= CML530 chromosome) |
| `cM` | genetic position, Haldane, zeroed per group |
| `cml_chr`, `cml_pos` | CML530 assembly coordinate (the mapping order) |
| `b73v5_chr`, `b73v5_pos` | B73 v5 coordinate, for cross-referencing |
| `cml_allele` | which DArT allele CML530 carries (`ref`/`alt`) — the A/H/B polarity |
| `maf`, `p121` | minor allele frequency, 1:2:1 goodness-of-fit p |

Raw genotypes are not redistributable. Pipeline scripts are in `scripts/qtl/`; the
protocol with every parameter and its justification is `scripts/qtl/PROTOCOL.md`.
