---
title: "F₂ QTL Scans — All Phenotypes"
subtitle: "QTL series notebook 3 — scanone across 14 traits with per-trait genome-wide permutation thresholds"
author: "Fausto Rodriguez Zapata"
date: today
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
---

::: {.callout-tip appearance="simple"}
**QTL series — notebook 3.** Uses the map from
[notebook 2](qtl_genetic_map.html); marker QC and encoding are in
[notebook 1](qtl_marker_pipeline.html).
:::

::: {.callout-note}
**Public data only.** Scan results — LOD by position and the per-trait permutation
thresholds — are published as `data/f2_qtl_scan_summary.tsv` and
`data/f2_qtl_scan_profiles.tsv`. No individual genotypes or phenotypes are included, and
everything below is reproduced from those two files.
:::

# Method

Every phenotype was scanned with `scanone` on the 1,152-marker map at a 1 cM pseudomarker
step, with `error.prob = 0.01` in `calc.genoprob` to match the value the map was built at
(measured immaterial at scan time: *r* = 0.998 against 2.5 × 10⁻³ across 3,005 positions).

Thresholds are **per-trait**, from 1,000 permutations each — not a shared threshold, since
the permutation distribution depends on the trait's own distribution. `PAPP` is an ordinal
1–4 score and was scanned non-parametrically (`model = "np"`); everything else used the
normal model.

```{r setup}
#| message: false
#| warning: false
library(data.table)
library(ggplot2)
library(knitr)
res <- fread("data/f2_qtl_scan_summary.tsv")
pr  <- fread("data/f2_qtl_scan_profiles.tsv")
theme_set(theme_bw(base_size = 11))
```

# Results by trait

```{r summary}
#| label: tbl-scans
setorder(res, -lod)
kable(res[, .(trait, model, n, shapiro_p,
              chr = peak_chr, `peak cM` = peak_cM, `peak Mb` = peak_Mb,
              LOD = lod, `thr 5%` = `perm_thr_0.05`, `thr 10%` = `perm_thr_0.10`,
              `sig 5%` = fifelse(sig05, "yes", ""),
              `sig 10%` = fifelse(sig10, "yes", ""))],
      caption = "Peak LOD per trait against its own 1,000-permutation genome-wide threshold. Ranked by LOD.")
```

```{r counts}
cat(sprintf("traits scanned: %d\n", nrow(res)))
cat(sprintf("clearing their own 5%% genome-wide threshold : %d (%s)\n",
            sum(res$sig05), paste(res[sig05 == TRUE, trait], collapse = ", ")))
cat(sprintf("clearing 10%% but not 5%%                     : %d (%s)\n",
            sum(res$sig10 & !res$sig05),
            paste(res[sig10 == TRUE & sig05 == FALSE, trait], collapse = ", ")))
```

# Scan profiles

Dashed line is each trait's own 5 % genome-wide threshold.

```{r profiles}
#| label: fig-profiles
#| fig-width: 12
#| fig-height: 14
#| fig-cap: "scanone profiles for all traits. Each panel carries its own permutation threshold."
pr[, chrf := factor(chr, levels = 1:10)]
pr <- merge(pr, res[, .(trait, thr = `perm_thr_0.05`, sig05)], by = "trait")
pr[, trait := factor(trait, levels = res$trait)]
ggplot(pr, aes(pos_cM, lod)) +
  geom_line(aes(colour = sig05), linewidth = 0.35) +
  geom_hline(aes(yintercept = thr), linetype = "dashed",
             colour = "firebrick", linewidth = 0.3) +
  facet_grid(trait ~ chrf, scales = "free_x", space = "free_x", switch = "y") +
  scale_colour_manual(values = c(`TRUE` = "black", `FALSE` = "grey55"), guide = "none") +
  labs(x = "position (cM)", y = "LOD") +
  theme(panel.spacing.x = unit(0.04, "lines"),
        strip.text.y.left = element_text(angle = 0), strip.placement = "outside")
```

# The traits that clear their own threshold

```{r hits}
#| label: tbl-hits
hits <- res[sig05 == TRUE][order(-lod)]
kable(hits[, .(trait, chr = peak_chr, `peak Mb` = peak_Mb, LOD = lod,
               `thr 5%` = `perm_thr_0.05`,
               margin = round(lod - `perm_thr_0.05`, 3),
               `CI Mb` = paste0(ci_lo_Mb, "–", ci_hi_Mb),
               `CI width Mb` = round(ci_hi_Mb - ci_lo_Mb, 1))],
      caption = "Traits above their own 5% genome-wide threshold, with 1.5-LOD support intervals in CML530 megabases.")
```

**EW** (ear weight) is the strongest signal and by far the best localised — a 4.9 Mb
support interval on chr1, against 13.6 Mb for PH and 7.4 Mb for HI. **PH** (plant height)
peaks on chr5 at 19.6 Mb with a dominance-driven effect (*d* = 9.76 ± 2.46 against
*a* = 3.82 ± 1.78). **HI** clears by 0.023 LOD.

## Correlated traits share peaks by construction

```{r shared}
#| label: tbl-shared
kable(res[order(peak_chr, peak_cM),
          .(trait, chr = peak_chr, `peak cM` = peak_cM, LOD = lod)][
          `peak cM` %in% res[, .N, by = peak_cM][N > 1, peak_cM]],
      caption = "Traits peaking at identical map positions. These are not independent observations.")
```

`GY`, `GYPP` and `PUE` share an identical peak and identical LOD because they are
arithmetic functions of one another; the same holds for `SW`, `TDM`, `SD` and `RW` at
chr2. The handover records the algebra: `RYE/HI = 1/RSR`, and `TDM = SW + RW`. Counting
these as separate findings would be double-counting.

# Nothing survives correction for 14 traits

```{r correction}
#| label: tbl-correction
k <- nrow(res)
cat(sprintf("traits scanned: %d\n", k))
cat(sprintf("expected traits clearing a per-trait 5%% threshold by chance: %.1f\n", 0.05 * k))
cat(sprintf("observed: %d\n", sum(res$sig05)))
cat(sprintf("P(at least one false positive) if traits were independent: %.1f%%\n",
            100 * (1 - 0.95^k)))
kable(hits[, .(trait, LOD = lod, `thr 5%` = `perm_thr_0.05`,
               margin = round(lod - `perm_thr_0.05`, 3))],
      caption = "How much room the significant traits actually have.")
```

Three traits clear their own 5 % genome-wide threshold. That is the correct threshold for
**one** trait tested in isolation — it is not the right one here, because 14 traits were
scanned.

- Under the null, **0.7 traits** would be expected to clear a per-trait 5 % threshold by
  chance. Three were observed, but the margins are **0.186, 0.059 and 0.023 LOD** — all
  three sit essentially on the line.
- The traits are **not independent**, so neither the naive 51 % family-wise error rate nor
  a plain Bonferroni correction is right. The true rate lies between 5 % and 51 %.
- Under any correction for 14 traits — Bonferroni gives a per-trait α of 0.0036 — **none
  of the three survives.**

::: {.callout-important}
## Conclusion

**No QTL in this experiment survives correction for the number of traits scanned.** Three
traits reach their individual 5 % genome-wide thresholds, but by margins of 0.02–0.19 LOD,
in a set of 14 correlated traits where ~0.7 such results are expected by chance.

Localisation is a separate question from significance, and the intervals should be read
on their own terms. Counted against the CML530 annotation:

| trait | interval | width | protein-coding genes |
|---|---|---|---|
| EW | chr1:286.38–291.24 Mb | 4.9 Mb | **124** |
| HI | chr5:215.97–223.40 Mb | 7.4 Mb | 329 |
| PH | chr5:17.06–30.61 Mb | 13.6 Mb | 283 |
| GY / GYPP / PUE | chr1:202.00–259.45 Mb | 57.5 Mb | 1,165 |

EW's 124-gene interval is a workable starting point for a candidate search — that is a
normal first-pass F₂ resolution, not a dead end. The 57 Mb suggestive intervals are not.
What limits EW is therefore **not** its localisation but its significance: a 0.186 LOD
margin in a 14-trait experiment.

This is a property of the **experiment**, not of the map. With 166 phenotyped plants the
detection floor is roughly 8–10 % of phenotypic variance, and the two most heritable
traits in the parallel diallel — days to anthesis and silking, h² ≈ 0.60 with Baker's
ratio ≈ 0.98, i.e. almost purely additive — **were never scored on these plants**. The
best F₂-mappable trait available is TDM at h² = 0.266, and PUE, the trait of interest, is
the least heritable of all at h² = 0.028.
:::

# What would change the answer

- **Phenotype flowering time on the F₂**, or on F₂:₃ rows if seed was kept. h² ≈ 0.60 and
  near-pure additivity is a completely different proposition from h² = 0.03.
- **Replicated F₂:₃ families.** One season, no new genotyping, and replication would cut
  the residual variance that currently sets the detection floor.
- **More plants.** At n = 166 the floor is ~8–10 % PVE; QTL below that are undetectable
  regardless of map quality.

Improving the map further will not help. At one marker per 1.9 Mb it already resolves far
finer than 372 meioses can exploit.

# Data availability

| file | contents |
|---|---|
| `data/f2_qtl_scan_summary.tsv` | one row per trait: peak position, LOD, permutation thresholds, support interval |
| `data/f2_qtl_scan_profiles.tsv` | LOD at every scan position for every trait |
| `data/f2_genetic_map.tsv` | the map the scans were run on |

Scripts: `scripts/qtl/scan_all_traits.R` (all traits), `scripts/qtl/scan_trait.R` (one
trait, with effect estimates). Protocol: `scripts/qtl/PROTOCOL.md`.
