---
title: "F₂ QTL population — data provenance, marker QC and ABH encoding"
subtitle: "Permanent record of the pre-mapping pipeline (DArTseq order DMz26-3123)"
date: 2026-07-30
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    df-print: kable
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(data.table)
D <- "data/qtl/derived"
have <- function(p) file.exists(p)
```

::: {.callout-important}
## Scope and status

This notebook is the **record of everything done before QTL mapping begins** — data
inventory, population verification, marker QC, the AGPv4 → NAM v5 liftover, parental
allele calling against the CML530 assembly, and the A/H/B encoding for R/qtl.

**No QTL mapping is performed here** — this notebook ends at the encoded marker matrix.
The map itself is in **[the genetic map notebook](qtl_genetic_map.html)** (1,152 markers,
1,858.4 cM, published as `data/f2_genetic_map.tsv`); the parameter-by-parameter protocol
is `scripts/qtl/PROTOCOL.md`.

The raw DArTseq reports and the phenotype workbook are **not redistributable** and are
gitignored (`data/qtl/`). This notebook and `scripts/qtl/` are the tracked record; the
data itself must be obtained from Stélio Nuvunga.
:::

# Provenance

| | |
|---|---|
Order | `DMz26-3123` |
Vendor | SEQART AFRICA (AICAD Complex, JKUAT Juja, Kenya) |
Client | Stélio Boaventura Nuvunga (Makerere University, PASET-RSIF) |
Product | maize DArTseq (GBS) 1.0 |
Order date | 2026-03-11 |
Alignment reference | B73 **AGPv4**, blast ≥90 % identity, e-value 5e-10 |
Samples | 186, two 96-well plates |

Three report files, all carrying the same 186 samples:

| File | Content | Used |
|---|---|---|
`Report_..._SNP_mapping_2.csv` | 5,208 SNPs, one-row format `0`/`1`/`2`/`-` | **yes** |
`Report_..._SNP_2.csv` | same 5,208 SNPs, two-row per-allele format | no (redundant) |
`Report_..._SilicoDArT_1.csv` | 11,518 dominant presence/absence markers | no |

The SilicoDArT markers are **dominant** — they cannot separate a heterozygote from one
homozygote, so they carry no dominance information and are unusable for an F₂. They were
excluded, not filtered.

::: {.callout-important}
## The genotype coding is undocumented in the delivery — and verified here

`metadata.json` carries a `datafiles` block with format descriptions, and it contains
exactly **two** entries: `Report_..._SilicoDArT_1.csv` and `Report_..._SNP_2.csv`.
**`Report_..._SNP_mapping_2.csv` — the one-row file that every genotype in this
analysis comes from — has no entry.** SEQART documented two formats and shipped a third
without one.

That matters because the three formats use *different* codings:

| file | coding |
|---|---|
SilicoDArT | `1` = presence, `0` = absence, `-` = low count |
two-row SNP | `0`/`1`/`-` per allele row — presence of that allele |
**one-row SNP (undocumented)** | **`0` = hom ref, `1` = hom alt, `2` = HETEROZYGOTE, `-` = missing** |

**It is NOT a minor-allele dosage.** A dosage would put the heterozygote at `1` and
hom-alt at `2`. Reading it that way — the natural guess for a `0/1/2` column, and
consistent with the habit the two documented files establish — swaps H and B on
**every** marker. Because the swap is systematic, the resulting map still looks
internally coherent while every heterozygote is mislabelled a homozygote.

Verified directly against DArT's own per-marker summary columns
(`scripts/qtl/verify_dart_codes.R`), not inferred:

The test recomputes each genotype frequency from the raw calls and compares it to the
column DArT reports. The number below is the **largest absolute disagreement across all
5,208 markers** — not a p-value, and no statistical test is involved. Agreement at
`5e-07` is DArT's printing precision, i.e. exact.

| recomputed from the calls | DArT's column | largest absolute difference (5,208 markers) |
|---|---|---|
frequency of code `0` | `FreqHomRef` | 0.0000005 |
frequency of code `1` | `FreqHomSnp` | 0.0000005 |
frequency of code `2` | `FreqHets` | **0.0000005** |
called ÷ n samples | `CallRate` | 0.0000005 |

The dosage misreading, tested the same way, disagrees by **0.995** — frequency of code
`1` against `FreqHets`, and code `2` against `FreqHomSnp`, both fail. So
`code 2 = heterozygote` is established outright, not by inference.

This was originally taken from the values themselves plus the fact that A/H/B came out
near 25/50/25 — suggestive but not proof, and it was labelled in the scripts as though
settled before it had been checked. Every genotype in the pipeline depends on it.
:::

The CML530 reference assembly (`Zm-CML530-REFERENCE-HiLo-1.0`, MaizeGDB, released June
2024) is distributed under the **Toronto Agreement**, which restricts publishing
genome-wide analyses ahead of the assembly producers. Relevant to manuscript timing.

# Population

::: {.callout-note}
The tests below were run on an **independent set of 1,398 informative markers** selected
for population verification. That set is not a stage of the mapping funnel in *Marker QC*
below, and the two counts should not be reconciled against each other.
:::

## It is a single biparental F₂

Stélio's project description refers to F₂ populations from *all* 49 combinations
(plural), which would invalidate the R/qtl `f2` cross type if the 186 plants were pooled
across families. Three independent tests say they are not
(`scripts/qtl/structure_check.R`):

**Allele frequency.** In a single selfed F₁ every informative marker must sit at
*p* ≈ 0.5. Observed SD of *p*(ref) across the 1,398 informative markers used for this
test = **0.028**, against a
pure binomial expectation at *n* = 186 of **0.0259** — a ratio of 1.08, i.e. sampling
noise and nothing else. **Zero** markers fall outside *p* = 0.40–0.60.

**PCA.** PC1 explains **7.9 %**, decaying smoothly (7.9, 7.0, 6.5, 6.1, 5.7 …), with
PC1~sd~/PC2~sd~ = 1.07. No dominant axis, no clusters. *k* pooled families would produce
*k*−1 large separated components.

**Relatedness.** Median pairwise IBS **0.376** — the textbook F₂ full-sib value. No pair
above 0.90, so no duplicated samples. No plate effect (*p* = 0.75).

## No parent was genotyped in this order

The parents are inbred lines, so heterozygosity must be ≈ 0. Over the same 1,398
informative markers the **minimum** observed heterozygosity across all 186 samples is
**0.267**, and
the median is **0.5016** — dead on the F₂ expectation of 0.50. Not one sample is
consistent with an inbred. Combined with the sample naming (`plant 1` … `plant 188`, no
line names) and uniform `none [Original species: Zea maize]` comments, no parental
sample was submitted.

Incidentally, that median of 0.5016 also shows **allelic dropout is not distorting the
retained marker set**, despite modest read depth (median 6.9× ref / 5.5× alt). The
low-`FreqHets` tail is uninformative markers where the parents share the allele, not a
dropout artefact.

## Parents

From `README.md` and `multilocation_alpha_lattice.qmd`, the NC Design II female set is
**CML364, CML366, CML434, CML435, CML439, CML530, CML532** and the male testers are
**EN17, EN20, EN21, EN25, EN31, EN32, EN64** (IIAM IDs; legacy `MOZL3`–`MOZL10`).

`multilocation_alpha_lattice.qmd` records explicitly that *"the earlier draft had
incorrectly listed CML440"* — so **CML530 is a female parent of this programme**, and an
earlier parent list circulated with CML440 in its place.

::: {.callout-warning}
## The female parent is established; the tester is not
**CML530 is the female parent** — it is one of the seven CIMMYT female donors in the
7 × 7 line × tester design, so any F₂ from this programme with CML530 in its pedigree has
it as the female. That is the design, not an assumption.

**Which of the seven `EN` testers was the male is still open.**
`paper/maize_pue_manuscript.md` lists CML530 × EN31 among the most stable high-yielding
hybrids by WAASY, but stability ranking is not a phosphorus contrast — a lead, not
evidence. **Confirm with Stélio before publishing allele-effect directions.**

Worth noting from the pedigrees (`multilocation_alpha_lattice.qmd`): every tester is
CIMMYT-derived on one side — EN17/20/21 from `CML491-1`, EN25/31/32 from `CLRCY034-1`,
EN64 from `CML539-1`. So both parents share CIMMYT ancestry, and at loci where the tester
inherited the CML haplotype the two parents carry the same allele by descent, making
those markers monomorphic in this cross.
:::

# Phenotypes

`data/qtl/mozpue_phenotype.xlsx`, sheet `data base`. Row 1 is a **units** row, not data.
188 body rows, **168 with data**; 20 IDs entirely blank
(3, 11, 19, 21, 39, 47, 61, 74, 110, 116, 127, 137, 138, 156, 158, 160, 161, 169, 175, 177).

Trait names in the source file are free-text Excel headers. They are mapped to the
project abbreviations through **`data/phenotype_dictionary.csv`**, which is the single
authority (built by `scripts/qtl/phenotype_dictionary_unify.R`) and carries a
`raw_header` column so no downstream script duplicates the mapping.

```{r dict}
if (have("data/phenotype_dictionary.csv")) {
  dd <- fread("data/phenotype_dictionary.csv")[project == "qtl_f2" & role == "trait"]
  dd[, .(variable, name, unit, formula, raw_header)][order(variable)]
}
```

## Derived formulas, recovered numerically

None of the derived columns were documented. All were reproduced from the data to
floating-point precision (`scripts/qtl/phenotype_dictionary_unify.R`):

| Trait | Formula | max&#124;diff&#124; |
|---|---|---|
`GY` | `GYPP × 0.0533333` | 5.0e-06 |
**`PUE`** | **`GY(kg/ha) / 20`** | 1.4e-14 |
`HI` | `GYPP / SW` | 0 |
`RSR` | `RW / SW` | 0 |
`RYE` | `GYPP / RW` | 0 |
`TDM` | `SW + RW` | 0 |

::: {.callout-note}
## The phosphorus rate was recoverable from the arithmetic
`PUE = GY(kg/ha)/p_input` with the implied `p_input` having **sd = 2.8e-16** across all
168 plants ⇒ the F₂ trial received **20 kg P/ha**. This is neither of the multilocation
rates (10 stress / 50 optimal), so the F₂ trial is its own phosphorus regime and is
**not** directly comparable to the diallel's stress treatment. Worth confirming with
Stélio, but the value is unambiguous.

The `GY` conversion implies ~53,333 plants/ha (0.1875 m²/plant).
:::

::: {.callout-warning}
## Three cautions that affect interpretation

1. **`HI` here is not harvest index.** `HI = GYPP/SW`, and `TDM = SW + RW` confirms `SW`
   excludes the ear. So this is grain-per-shoot, not grain ÷ total above-ground biomass.
   Reporting it as harvest index needs a definitional footnote.
2. **`HI`, `RSR`, `RYE` are algebraically linked** — three ratios over the same three
   measurements, with `RYE/HI = 1/RSR`. They *will* share QTL by construction. Any
   co-localization claim among them is arithmetic, not pleiotropy. Decide which traits
   are primary before interpreting.
3. **`PAPP` is an ordinal 1–4 score.** `scanone` assumes a roughly normal continuous
   phenotype; use `model = "np"` for it.
:::

15 plants are flagged `Higher M.C` (moisture content): 65, 72, 86, 95, 111, 121, 125,
128, 136, 139, 140, 153, 163, 172, 178. **Grain yield is not standardised to a common
moisture basis.**

`HI` and `RYE` in the *multilocation* dataset could **not** be reproduced by any of 10
and 7 candidate formulas respectively (closest miss 0.643 and 15.6, far beyond rounding).
Both appear in the manuscript. That is a provenance gap in the diallel data, recorded
here because it shares abbreviations with this dataset.

# Effective population size

```{r join}
data.table(
  quantity = c("genotyped", "phenotyped", "**both (mapping population)**",
               "genotyped only", "phenotyped only"),
  n = c(186, 168, 166, 20, 2))
```

**The map is built on all 186 genotyped plants** (`est.map` uses no phenotypes), but
**scans use n = 166** — the plants with both genotype and phenotype. For an F₂ at a 5 %
genome-wide threshold this gives roughly 80 % power for QTL explaining ≥ 8–10 % of
phenotypic variance. Smaller effects will be missed and
detected effects will be **Beavis-inflated**. There is **no within-genotype
replication**, so no heritability can be estimated and genetic variance cannot be
separated from micro-environmental variance.

This was a deliberate design choice, not an oversight: in July 2025 Joe Gage set out both
the F₂:₃ route (row-level BLUPs, replicable) and direct F₂ phenotyping, noting the latter
*"will likely have more error in it, since the F₂ plants cannot be replicated."* The
faster route was taken.

::: {.callout-tip}
## Largest available improvement
If F₃ seed was collected from these 186 selfed F₂ plants, F₂:₃ rows can be planted and
the **same genotypes** re-mapped against replicated BLUPs — one season, no new
genotyping. Worth asking Stélio.
:::

# Marker QC

The 1:2:1 test stands in for "polymorphic between the parents", which is otherwise
unobservable because no parent was genotyped.

```{r mkqc}
data.table(
  step = c("SNPs in the DArT report",
           "anchored to AGPv4 chr 1–10",
           "lifted 1:1 to NAM v5, same chromosome",
           "CML530 allele defined (ref or alt)",
           "MAPQ ≥ 30, no XA, mapped",
           "on CML530 chr 1–10",
           "MAF > 0.15",
           "1:2:1 at BH FDR 0.05",
           "one marker per 69-bp tag (CloneID)",
           "|cor| ≥ 0.30 with a neighbour  →  **encoded**"),
  n = c(5208, 3884, 3828, 3728, 2994, 2991, 2067, 1698, 1692, 1688))
```

Per-sample call rate 0.64–0.97 (median 0.936); only **plant 14** below 0.80. Missing
genotypes in the encoded matrix: **0.47 %**.

::: {.callout-important}
## `CallRate` and `RepAvg` are NOT used, and that was measured

Earlier drafts filtered on `CallRate ≥ 0.90` and `RepAvg ≥ 0.95`. Neither survived
testing against markers that actually misbehave — those that correlate with none of
their neighbours (`scripts/qtl/report_callrate_repavg.R`):

* **`RepAvg` is flat.** The offender rate is 4.1 % in the lowest band against 3.6 % in the
  highest. A 0.95 cut takes 46 of 117 offenders while sacrificing **768 sound markers**.
* **`CallRate` has a real gradient** (19.3 % offenders at 0.90–0.95 vs 3.5 % above 0.95)
  but **80 of the 117 offenders sit above 0.95**, so it cannot reach them either.

`RepAvg` is in any case DArT's *internal* technical-replicate repeatability — it cannot be
recomputed from these 186 samples, which contain no duplicate pairs (max IBS 0.687).

What replaced them is the **alignment-quality filter**, which does separate the two
classes: markers anti-phase to both neighbours multi-map at 70.4 % against a 6.2 %
baseline (median MAPQ 4 vs 60, Fisher *p* = 2.3 × 10⁻⁸).
:::

::: {.callout-note}
## A caveat on the 1:2:1 filter
A paralogous DArT tag collapsing two loci produces apparent heterozygotes at high
frequency and **can pass a 1:2:1 test** while not being a single locus. F₁-het
ascertainment (as used in the airmine population) is direct evidence that parents differ;
a 1:2:1 fit is only statistical. Two independent handles on this exist: the `find_quirky`
map-geometry rule (see *Next steps*), and the CML530 `het-call` class below.
:::

# Liftover AGPv4 → NAM v5

`scripts/qtl/liftover_v4_to_v5.R`, reusing `lift_unique()` from
`zealhmm/R/teonam_liftover.R` verbatim — single hop, unique 1:1 mappings only.

```{r lift}
data.table(
  quantity = c("submitted (v4-anchored, chr 1–10)", "lifted 1:1", "failed / multi-mapping",
               "lifted onto non-chr1-10 scaffold", "chromosome changers",
               "carried into allele calling"),
  n = c("3,884", "3,828 (98.6 %)", "56", "0", "0", "3,828"))
```

**v4 → v5 is far cleaner than v2 → v5.** Zero chromosome changers and zero off-chromosome
scaffolds, so the chr7-displaced-block class of artefact that motivated `find_quirky`'s
island rule does not appear. Per-chromosome Spearman ρ between v4 and v5 position is
≥ 0.988 everywhere (chr5, 8, 9 exactly 1.00000). 715 markers change local rank —
fine-scale reshuffling within an otherwise collinear order.

Two things to watch: **chr3 has the lowest collinearity** (ρ = 0.98839), and **chr6's
marker span shrank 170.8 → 146.3 Mb** while losing only 3 markers.

::: {.callout-caution}
## A silent-failure class encountered twice
The chain file names chromosomes `1`…`10`; both FASTAs on disk name them
`chr1`…`chr10`. Without a prefix, `bcftools mpileup -T` matches **nothing** and returns an
empty result *without erroring*. Separately, a third leading column in an R/qtl `csvsr`
genotype file is parsed as **an extra individual** whose values become unrecognised
genotype codes silently converted to missing (verified in
`scripts/qtl/rqtl_format_probe.R`). Both produce a wrong answer with a clean exit.
:::

# Parental alleles from the CML530 assembly

`scripts/qtl/cml530_alleles_direct.R`. Because CML530 **is** the female parent, this
yields actual parental alleles and ABH polarity is directly determined rather than
proxied.

**The direct route is the one in the pipeline.** The 3,828 tags are mapped to the CML530
assembly once (`tags_vs_CML530.bam`); for each tag the CIGAR is walked to the tag's own
`SnpPosition` and CML530's base read there, reverse-complemented on reverse-strand
alignments. Walking the CIGAR rather than computing an offset by hand keeps the base
assignment honest if CML530 carries an indel near the site.

A second, **superseded** route (`cml530_genotypes_pileup.sh`) extracted each aligned span,
re-mapped it to B73 v5 and called the base from a pileup at the lifted v5 coordinate.
It is retained only as an independent check — see the convergence test below — and is
listed in `scripts/qtl/DELETE_CANDIDATES.md`.

## Attrition

The direct route loses only the tags that do not align at all: of 3,828 tags, **88 are
unmapped** and the remainder yield a base. Two sites carry a third allele — CML530's base
matches neither of DArT's — leaving **3,728 markers with a defined `ref`/`alt` polarity**.

```{r attrition}
data.table(
  cml_class = c("ref", "alt", "no-hit (unmapped tag)", "third allele",
                "**A = CML530 defined**"),
  n = c("2,403", "1,325", "98", "2", "**3,728**"))
```

::: {.callout-important}
## Mappability is handled downstream, not by discarding here
The earlier pileup route discarded 563 markers at `MAPQ < 20` before any genotype was
read, and reported 708 `no-hit`. The direct route keeps them and **carries MAPQ and the
`XA` alternative-alignment tag forward into `encode_abh.R`**, where they became the filter
that actually mattered: markers whose genotypes contradict their neighbours are
overwhelmingly multi-mapping tags (70.4 % vs a 6.2 % baseline). Filtering there rather
than here means the decision is made against evidence of misbehaviour, not against a
conventional MAPQ default.
:::

## Result

```{r cmlclass}
if (have(file.path(D, "CML530_marker_alleles_direct.tsv"))) {
  al <- fread(file.path(D, "CML530_marker_alleles_direct.tsv"),
              colClasses = list(character = "chr_v5"))
  al[, .N, by = cml_class][order(-N)]
} else {
  data.table(cml_class = c("ref", "alt", "no-hit", "third-allele"),
             N = c(2403, 1325, 98, 2))
}
```

**3,728 of 3,828 tags (97.4 %) carry real parental polarity**, and 1,688 of them survive
into the encoded matrix (902 `ref`, 786 `alt`). The `ref`/`alt` split confirms that CML530
carries the DArT reference allele at some sites and the alternate at others with no global
pattern — polarity **must** be assigned per marker.

The tag-base check is exact: CML530's base read from the assembly matched the tag's own
base at **3,730 of 3,730** alignments once strand was accounted for, with 1,850 of 3,740
alignments on the reverse strand.
::: {.callout-important}
## Correction: DArT's reference allele is **not** B73's allele
An earlier version of this notebook stated that DArT defined its reference allele by the
AGPv4 alignment, so that `ref` meant "CML530 matches B73". **That is false**, and it was
measured (`scripts/qtl/test_pileup_coords.R`): the B73 v5 base equals the DArT ref
allele at only **~76%** of markers.

DArT calls SNPs within its own tag clusters across the assayed panel, so ref/alt is a
convention derived from *this germplasm*; the `Chrom_`/`ChromPosSnp_` columns are added
afterwards by blasting the tag to B73. `ref` therefore means **CML530 carries the DArT
reference allele** — a statement about the DArT panel, not about B73 and not about the
ENxx tester.

This mattered: it produced two routes to CML530's allele that silently answered different
questions and agreed on only 71% of markers. Where DArT's ref happened to be B73's base
they agreed 98.7% of the time; where it did not, 20.8%. Coordinates were never the
problem — the liftover position and the span's independent v5 alignment agree exactly at
97.6% of markers.
:::

Exactly **one `het-call`** across 3,828 tags. An inbred cannot be truly heterozygous, so
that is the paralog / collapsed-repeat signal — and finding one says the tag alignments
are behaving cleanly. It is treated as unresolved, never as a genotype.

# A/H/B encoding for R/qtl

`scripts/qtl/encode_abh.R`. Only markers with a defined CML530 allele are encoded, so
**A = CML530 on every marker** and no arbitrary-polarity markers enter the matrix:

```
cml_class "ref"  ->  CML530 carries REF  ->  A = code 0, B = code 1
cml_class "alt"  ->  CML530 carries ALT  ->  A = code 1, B = code 0
code 2 = H (verified above), "-" = missing
```

Markers are ordered by **CML530 coordinate** (`cml_chr`, `ref_pos`), and all 186 genotyped
plants are kept — `est.map` uses no phenotypes, so the map should use every genotyped
plant, not only the 166 that were also phenotyped.

```{r abh}
data.table(
  quantity = c("individuals", "markers", "chromosomes", "traits",
               "percent genotyped", "A", "H", "B", "missing", "off-parent calls"),
  value = c("186", "1,688", "10", "14", "99.5 %",
            "25.1 %", "49.6 %", "25.4 %", "0.47 %", "0"))
```

A/H/B at 25.1 / 49.6 / 25.4 against the F₂ expectation of 25/50/25, with **zero**
off-parent calls, is an end-to-end check on the whole chain — DArT calls → alignment
filter → liftover → polarity → encoding. `read.cross(format = "csvsr", crosstype = "f2")`
round-trips, asserted inside the script.

::: {.callout-warning}
## Correction: recombination fractions are **not** orientation-invariant

An earlier draft kept unpolarisable markers on the grounds that "recombination fractions
are orientation-invariant, so a marker with no CML530 call still carries full mapping
information under an arbitrary orientation." **That is false.** Flipping *one* marker of a
pair turns θ into 1 − θ, so an out-of-phase marker looks *unlinked* to its neighbours and
`est.map` saturates at rf = 0.5 (1001.5 cM). Only flipping *both* leaves θ unchanged.

The consequence is visible in `build_map.R`, which drops markers anti-correlated with
**both** neighbours: each one otherwise makes every individual read as a double
recombinant. Encoding now admits only markers with a defined CML530 polarity, so the
question does not arise for the matrix itself.
:::

# Reproduction

Software: **R 4.5.2** (`qtl`, `data.table`, `readxl`, `rtracklayer`, `GenomicRanges`,
`Biostrings`), **minibwa 0.6-r416**, **samtools/bcftools 1.23.1**, UCSC `liftOver`.

::: {.callout-warning}
## minibwa must be built from source
`brew install minibwa` is built with `gpl=0`, which omits the low-memory BWT builder:
`ERROR: option -l not compiled as it depends on GPL'd code`. The flag still appears in
`--help`, so help text is not evidence it works. Without `-l`, indexing needs 18 N ≈ 39 GB
for a 2.2 Gb genome. Use `scripts/qtl/install_minibwa_from_source.sh`.
:::

Run order:

Full parameter-by-parameter detail is in **`scripts/qtl/PROTOCOL.md`**.

| # | Script | Notes |
|---|---|---|
1 | `phenotype_dictionary_unify.R` | writes `data/phenotype_dictionary.csv` |
2 | `data_audit.R` | **inventory only** — no filter it reports is applied downstream |
3 | `structure_check.R` | F₂ verification, parent absence, duplicates |
4 | `verify_dart_codes.R` | **proves `2` = heterozygote** against `FreqHets`; run before trusting any genotype |
5 | `liftover_v4_to_v5.R` | writes `markers_v5.tsv`, `tags_for_cml530.fa` |
6 | `install_minibwa_from_source.sh` | once; must be the source build (GPL `-l`) |
7 | `index_genomes.sh` | **~45 min**, ~9.5 GB, peak 7.2 GB RSS, single-threaded |
8 | `cml530_alleles_direct.R` | A = CML530 read from the assembly at each tag's `SnpPosition`; **3,728 markers**; writes `tags_vs_CML530.bam` |
9 | `encode_abh.R` | alignment-quality filter, MAF, FDR, A/H/B with A = CML530, **CML530 marker order** → **1,688**; writes `rqtl_gen_abh_all.csv` |
10 | `encode_phenotypes.R` | phenotypes in csvsr → `rqtl_phe.csv` |
11 | `build_map.R` | drop inverted → distortion → LD prune → `est.map` → `find_quirky` → `est.map` → **1,152 markers, 1,858 cM**; writes `data/f2_genetic_map.tsv` |
12 | `scan_all_traits.R` | scanone + 1,000 permutations per trait → `data/f2_qtl_scan_*.tsv` |

::: {.callout-note}
## The map is built — see the next notebook

This notebook ends at the encoded marker matrix. The map itself, its parameters and its
limitations are in **[the genetic map notebook](qtl_genetic_map.html)**, and the full
parameter-by-parameter protocol is `scripts/qtl/PROTOCOL.md`. The map is published as
`data/f2_genetic_map.tsv`.

Two things changed after this notebook was first written, and both mattered:

**The marker order is CML530's, not B73's.** B73 is not a parent of this cross. The two
assemblies place 37 markers on *different chromosomes*, and `est.map` estimates distances
along a fixed order without ever reordering — so a wrong order cannot be recovered later.
Within a chromosome the two orders barely differ (Spearman 0.9987); the gain is the
chromosome assignment.

**An alignment-quality filter was added.** Markers whose genotypes contradicted their
neighbours turned out to be tags that align to several places in CML530, so the parental
allele was read off an arbitrary copy: 70.4 % of them multi-map against a 6.2 % baseline
(median MAPQ 4 vs 60; Fisher p = 2.3 × 10⁻⁸), and 11 of 27 carry an alternative hit on the
very chromosome B73 assigns them to. Filtering on `MAPQ ≥ 30` with no `XA` tag cut that
class from 27 markers to 5. **MAPQ and `XA` were available all along and never used** —
`qc_pass` does not encode them.

`scripts/qtl/DELETE_CANDIDATES.md` lists 13 dead-end scripts from routes that failed.
:::

Supporting: `qtl_minibwa_index_bench.sh`, `qtl_minibwa_index_threads_bench.sh` (why `-l`
and why local rather than HPC), `qtl_rqtl_format_probe.R` (csvsr layout),
`qtl_index_wait_verify.sh` (artefact-based index validation).

::: {.callout-note}
## Two implementation traps recorded
**minibwa aborts on duplicate FASTA names** — `Assertion failed: (i - j <= 2),
map-main.c:144`. It reads consecutive same-named records as mate pairs and hard-aborts on
a third. `samtools faidx -r` names records by region and several tags share a CML530 span,
so spans are renamed to their unique `AlleleID`.

**Recursive `grep` here is `ugrep` and honours `.gitignore`**, so searches from the repo
root silently skip `agent/` and `data/qtl/`. Use explicit paths for those.
:::

# Next steps

The map is built — **[genetic map notebook](qtl_genetic_map.html)**, 1,152 markers,
1,858.4 cM, 96.7 % physical coverage, published as `data/f2_genetic_map.tsv`. Decisions
made after this notebook was written, all recorded with their evidence in
`scripts/qtl/PROTOCOL.md`:

1. **De novo grouping confirmed the chromosome assignment.** Connected components on the
   rf/LOD graph give a perfectly block-diagonal linkage-group × chromosome table — 0 of
   1,236 markers land in a component whose majority chromosome is not their own — so the
   physical order was never reordered, only validated.
2. **LD pruning replaced `findDupMarkers`**, which needs identical genotype patterns and
   found only 28 markers. Pruning is at Pearson **r = 0.95** on a correlation-derived
   genetic axis (Teng & Xu 2026: *r* = 1 − 2θ, so *d* = −50·ln *r*), thinned by the exact
   1-D MDdIS greedy (Jena et al. 2018).
3. **`error.prob = 0.01` throughout** — both `est.map` rounds and `calc.genoprob` at scan
   time. Immaterial at scan time (r = 0.998 against 2.5 × 10⁻³).
4. **Scans are done** for all 14 phenotypes with 1,000-permutation thresholds.

::: {.callout-warning}
## A length target that should not have been used

Earlier drafts of this notebook, and the `build_map.R` guardrail, compare total map
length against **1,348–1,596 cM**. That figure is **Chen et al. 2019 Table 1, the
per-subpopulation range for TeoNAM** — a *teosinte introgression BC1S4* population with
~13,733 SNPs per subpop. It is not a maize consensus map, not an F₂ expectation, and was
inherited wholesale from `teonam_qtl_permap.R` without being checked.

It also points the wrong way: map length *estimates* grow with accumulated meioses
(Teng & Xu show F₂ → F₃ → F₄ expansion), so a BC1S4 should read **longer** than an F₂ of
the same genome, not equal. Any statement in this repo that 1,858 cM is "20 % above the
maize expectation" is measured against a borrowed constant from the wrong design.

Pending a proper reference (Coe composite, NAM consensus, or similar), length is best
judged against **physical coverage and marker spacing**, which need no borrowed constant.
:::

## Outstanding questions for Stélio

X   one F₁ was selfed or several.
2. **Field book:** design, plot coordinates, planting/harvest dates, soil test P. (The
   applied rate is already recovered: 20 kg P/ha.) Plot coordinates would allow a spatial
   trend to be fitted, partly compensating for the absent replication.
3. **Was F₃ seed collected** from the selfed F₂ plants?
4. **Plate submission sheet** — confirm `plant N` ↔ field-book plant N. An off-by-one in
   plate layout would silently destroy the analysis and nothing currently documents the
   correspondence.
5. **`HI` and `RYE` definitions used in the diallel** — neither is reproducible from
   `multilocation.csv`, and both appear in the manuscript.
6. **`PAPP` and `EH`** — new abbreviations introduced here; his field-book names take
   precedence.
7. From SEQART: the **failed-sample list** (`targetslist` has 215 IDs, the report has 186).
