---
title: "Retrieving CML IDs for CIMMYT Maize Accessions from Genesys"
author: "Fausto Rodríguez-Zapata"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: show
    df-print: kable
execute:
  warning: false
  message: false
---

## Why the API, not URL scraping

The DOI strings in Genesys accession URLs are persistent identifiers
(one per accession, minted in the ITPGRFA GLIS), not an addressing
scheme you are meant to construct or parse. The web UI was never the
intended retrieval path, so the lack of predictable URLs is expected
rather than an anti-scraping measure.

The supported route is the REST API at `https://api.genesys-pgr.org`,
wrapped for R by the `genesysr` package. Genesys identifies an
accession by the triplet **instituteCode + accessionNumber + genus**;
where a DOI exists it is used as the unique key. Queries are expressed
as a `filter` list (AND across filters, OR within a filter) plus the
passport fields to return — this is the "complex multi-ID search" you
remember, now as a JSON request body instead of a URL.

**On CML IDs**: `CML` = *CIMMYT Maize Line*. This designation is not a
dedicated Genesys field; it is carried in the accession name
(`ACCENAME`). So the workflow is: fetch passport data for the target
accessions, then extract the `CML###` token from that field.

## Setup

**Purpose**: Load `genesysr` and authenticate in a way that also works for
scripted / rendered runs.

**Approach**: `genesysr::user_login()` runs the OAuth2 authorization-code
(browser) flow, but it only fires inside an *interactive* R session and keeps
the token in memory — so it cannot be reused when this document is rendered or
sourced. We instead use a small token bridge (`R/genesys_auth.R`) that runs the
*same* browser/Google login once and persists the returned token, **including
its refresh token**, to `~/.genesys_token.rds`. Every later run loads it and
refreshes silently — no further browser trips. The OAuth client is genesysr's
built-in public client; Genesys accepts Google as an identity provider.

**Expected outcome**: An authenticated session against production.

```{r setup}
library(genesysr)
library(tidyverse)

# Token bridge: browser login (once) -> saved, auto-refreshing token.
source("R/genesys_auth.R")

# Production is the default; this is explicit for reproducibility.
setup_production()
```

Run the browser login **once** (interactive; opens a browser for the Google
login). It writes the refreshable token to disk, after which this chunk is not
needed again:

```{r login, eval=FALSE}
genesys_browser_login()
```

Load (and, if expired, refresh) the saved token for this session:

```{r auth}
genesys_use_saved_token()
```

## Target accessions

**Purpose**: Define the 21 CIMMYT maize accessions to resolve.

**Approach**: Store the exact accession-number strings as given. CIMMYT
prefixes its bank accession numbers (`CIMMYTMA`), so the numeric tail is
also derived for defensive matching later.

**Expected outcome**: A character vector plus a normalization helper.

```{r targets}
target_accessions <- c(
  "CIMMYTMA 29755", "CIMMYTMA 29754", "CIMMYTMA 29753",
  "CIMMYTMA 29752", "CIMMYTMA 29751", "CIMMYTMA 29750",
  "CIMMYTMA 28038", "CIMMYTMA 28037", "CIMMYTMA 28035",
  "CIMMYTMA 28034", "CIMMYTMA 28033", "CIMMYTMA 27991",
  "CIMMYTMA 27990", "CIMMYTMA 27989", "CIMMYTMA 27988",
  "CIMMYTMA 27987", "CIMMYTMA 27986", "CIMMYTMA 27985",
  "CIMMYTMA 27984", "CIMMYTMA 27983", "CIMMYTMA 27982"
)

# Collapse internal whitespace so "CIMMYTMA 29755" and
# "CIMMYTMA  29755" compare equal.
normalize_accenumb <- function(x) {
  str_squish(toupper(x))
}
```

## Targeted query by accession number

**Purpose**: Pull passport data for only the 21 accessions, avoiding a
bulk download of CIMMYT's ~28k maize collection.

**Approach**: Filter on institute (`MEX002`) and genus (`Zea`), and narrow to
each accession with the API's free-text key, `_text`. The MCPD-style
`accessionNumber` filter key returns **HTTP 500** on the current production API,
and most other candidate keys (`acceNumb`, `accessionName`, …) are silently
ignored — `_text` is the only key that actually narrows results (verified
empirically). So we run one small `_text` search per accession and exact-match
locally in the next section. Request the identifier and name fields plus the
fields likely to hold the CML token.

**Expected outcome**: A data frame with at least the 21 targets (a free-text
search may also return a few near-matches, removed by the exact join below).

```{r query-targeted}
fields <- list(
  "INSTCODE", "ACCENUMB", "DOI", "GENUS", "SPECIES", "ACCENAME"
)

base_filter <- list(
  institute = list(code = list("MEX002")),
  taxonomy  = list(genus = list("Zea"))
)

# One cheap free-text query per accession; combine and de-duplicate.
acc_raw <- map_dfr(target_accessions, function(a) {
  filt <- c(base_filter, list("_text" = a))
  get_accessions(filters = filt, fields = fields)
}) %>%
  distinct()

# Diagnostics
cat("Rows returned:", nrow(acc_raw),
    "| Targets requested:", length(target_accessions), "\n")
```

## Extract CML IDs

**Purpose**: Pull the `CML###` token from the accession name and align
it to the requested accession numbers.

**Approach**: Because the query is issued with `mcpd = true`, the returned
columns carry the MCPD field codes (`ACCENUMB`, `ACCENAME`, `DOI`), not the
JSON property names. Locate the accession-number and name columns
case-insensitively, extract `CML` followed by digits from the name, and
left-join back onto the full target list so missing hits are explicit.

**Expected outcome**: One row per target accession with `cml_id`
populated where found, `NA` otherwise.

```{r extract-cml}
#' Find a column by case-insensitive name fragment
#'
#' @param df A data frame.
#' @param fragment Substring to match against column names.
#' @return Character vector of matching column names (possibly empty).
find_cols <- function(df, fragment) {
  names(df)[str_detect(names(df), regex(fragment, ignore_case = TRUE))]
}

acn_col  <- find_cols(acc_raw, "ACCENUMB")[1]
name_col <- find_cols(acc_raw, "ACCENAME")[1]

stopifnot(
  !is.na(acn_col),
  !is.na(name_col)
)

# CML token, with optional space normalized away (CML 579 -> CML579).
cml_pattern <- "CML[[:space:]]?[0-9]+"

acc_cml <- acc_raw %>%
  mutate(
    accenumb_norm = normalize_accenumb(.data[[acn_col]]),
    cml_id = str_remove(
      str_extract(.data[[name_col]], cml_pattern), "\\s"
    )
  )

# Align to the requested set so unmatched targets are visible.
result <- tibble(accession_number = target_accessions) %>%
  mutate(accenumb_norm = normalize_accenumb(accession_number)) %>%
  left_join(
    acc_cml %>% select(accenumb_norm, cml_id, all_of(name_col)),
    by = "accenumb_norm"
  ) %>%
  select(
    accession_number, cml_id, accession_name = all_of(name_col)
  )

# Diagnostics
cat("Matched:", sum(!is.na(result$cml_id)),
    "| Unmatched:", sum(is.na(result$cml_id)), "\n")
```

```{r show-result}
result
```

## Save outputs

```{r save}
dir.create("output", showWarnings = FALSE)
write_csv(result, "output/cml_ids.csv")
saveRDS(acc_raw, "output/genesys_passport_raw.rds")
```

## Fallback: institute + genus, match locally

**Purpose**: A path that does not depend on per-accession filtering at all,
in case `_text` behavior changes.

**Approach**: Query CIMMYT maize once, fetching only identifier and name
fields, then filter locally by normalized accession number. This pulls
a large collection (~28k rows), so run it only if the targeted query failed.

**Expected outcome**: Same `result` table, derived by local matching.

```{r fallback, eval=FALSE}
acc_all <- get_accessions(
  filters = list(
    institute = list(code = list("MEX002")),
    taxonomy  = list(genus = list("Zea"))
  ),
  fields = fields
)

acn_col <- find_cols(acc_all, "ACCENUMB")[1]

acc_all %>%
  mutate(accenumb_norm = normalize_accenumb(.data[[acn_col]])) %>%
  filter(accenumb_norm %in% normalize_accenumb(target_accessions))
```

## Appendix

### On the accession-number filter key

The MCPD-style `accessionNumber` filter key is **not** honored by the current
production `/acn/query` endpoint: passing it (in any shape — list, scalar, or
alone) returns **HTTP 500**. Other plausible keys (`acceNumb`, `accessionName`,
`names.name`) are accepted but silently ignored (the filter is not applied, so
all rows come back). Empirically, `_text` — the free-text search key — is the
only one that narrows results, which is why the targeted query uses it.

To re-check this against a future API version, open
`https://www.genesys-pgr.org/a/overview`, search for one of these accessions in
the UI, and inspect the POST body your browser sends in the Network tab; adopt
whatever key it uses.

### Headless authentication

For scheduled or scripted runs, register a client with Genesys
(`helpdesk@genesys-pgr.org`) and configure client credentials instead of
`user_login()`:

```{r headless, eval=FALSE}
setup(
  server        = "https://api.genesys-pgr.org",
  client_id     = Sys.getenv("GENESYS_CLIENT_ID"),
  client_secret = Sys.getenv("GENESYS_CLIENT_SECRET")
)
```

### Rate limits

The API is paginated and rate-limited, and bulk pulls are explicitly
flagged "use sparingly." Prefer the targeted query; batch by institute
or taxon if you ever scale this beyond a short ID list.