---
title: "Stability Analysis: AMMI and GGE Biplots"
subtitle: "Analysis 6 of the PUE hybrid series — genotype × environment interaction and identification of stable, high-performing hybrids for GY and PUE"
author: "Fausto Rodriguez Zapata"
date: "2026-07-06"
format:
  html:
    theme: cosmo
    embed-resources: true
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    code-tools: true
---

::: {.callout-tip appearance="simple"}
**Paper series — notebook 6.** Shared data preparation: `R/prep_traits.R`.
:::

# Methods (for the paper)

Genotype × environment interaction and stability were analysed for the two priority traits
(**GY, PUE**) with the `metan` package across the five environments (49 hybrids, 3 reps).
The **AMMI** model (additive main effects + multiplicative interaction) decomposes the G×E
into interaction principal components (IPCA); the **AMMI2 biplot** (IPCA1 vs IPCA2) shows how
hybrids and environments interact, and stability is summarised by the **WAAS** index
(weighted average of absolute scores — smaller = more stable). Because a stable but
low-yielding hybrid is not useful, we combine stability and performance in the **WAASY**
index (equal weight to mean and stability) to rank hybrids. The **GGE biplot** (genotype +
G×E) gives the *which-won-where* view (mega-environments and their winning hybrids) and the
mean-vs-stability view.

::: {.callout-note appearance="simple"}
PUE across environments blends the optimal and P-stress sites (different applied-P
denominators); its stability here is the stability of the across-environment PUE ranking.
:::

```{r}
#| label: setup
#| message: false
#| warning: false

library(metan)
library(tidyverse)
library(knitr)
library(kableExtra)

source("R/prep_traits.R")

dat <- hybrids %>% transmute(ENV = env, GEN = cross, REP = rep, GY, PUE)

ammi_gy  <- performs_ammi(dat, ENV, GEN, REP, resp = GY,  verbose = FALSE)
ammi_pue <- performs_ammi(dat, ENV, GEN, REP, resp = PUE, verbose = FALSE)
gge_gy   <- gge(dat, ENV, GEN, resp = GY)
gge_pue  <- gge(dat, ENV, GEN, resp = PUE)
waas_gy  <- waas(dat, ENV, GEN, REP, resp = GY,  verbose = FALSE)
waas_pue <- waas(dat, ENV, GEN, REP, resp = PUE, verbose = FALSE)
```

# AMMI analysis

```{r}
#| label: ammi-anova
#| warning: false
bind_rows(
  ammi_gy$GY$ANOVA   %>% mutate(Trait = "GY",  .before = 1),
  ammi_pue$PUE$ANOVA %>% mutate(Trait = "PUE", .before = 1)
) %>%
  filter(Source %in% c("ENV","GEN","GEN:ENV","PC1","PC2")) %>%
  transmute(Trait, Source, Df,
            `Mean Sq` = round(`Mean Sq`, 3),
            `F` = round(`F value`, 2),
            p = signif(`Pr(>F)`, 2),
            `% G×E` = round(Proportion, 1)) %>%
  kable(caption = "AMMI ANOVA (key sources) for GY and PUE; PC1/PC2 = interaction principal components with their share of the G×E sum of squares.") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r}
#| label: ammi-biplot
#| fig-width: 12
#| fig-height: 5.5
#| warning: false
arrange_ggplot(
  plot(ammi_gy,  type = 2) + labs(title = "GY"),
  plot(ammi_pue, type = 2) + labs(title = "PUE"),
  ncol = 2
)
```

# GGE biplots

*Which-won-where*: the polygon vertices are the best hybrids in each sector; environments
falling in a sector are "won" by that vertex hybrid (mega-environments).

```{r}
#| label: gge-www
#| fig-width: 12
#| fig-height: 5.5
#| warning: false
arrange_ggplot(
  plot(gge_gy,  type = 3) + labs(title = "GY — which-won-where"),
  plot(gge_pue, type = 3) + labs(title = "PUE — which-won-where"),
  ncol = 2
)
```

*Mean vs stability*: the abscissa ranks hybrids by mean performance; distance along the
ordinate measures instability.

```{r}
#| label: gge-mvs
#| fig-width: 12
#| fig-height: 5.5
#| warning: false
arrange_ggplot(
  plot(gge_gy,  type = 2) + labs(title = "GY — mean vs stability"),
  plot(gge_pue, type = 2) + labs(title = "PUE — mean vs stability"),
  ncol = 2
)
```

# Stable, high-performing hybrids (WAASY)

```{r}
#| label: waasy
#| warning: false
top_waasy <- function(w, resp, n = 8) {
  w[[resp]]$model %>%
    dplyr::filter(type == "GEN") %>%
    dplyr::arrange(dplyr::desc(WAASY)) %>%
    dplyr::transmute(Trait = resp, Hybrid = Code,
                     Mean = round(Y, 3), WAAS = round(WAAS, 3), WAASY = round(WAASY, 1)) %>%
    head(n)
}

bind_rows(top_waasy(waas_gy, "GY"), top_waasy(waas_pue, "PUE")) %>%
  kable(caption = "Top 8 hybrids per priority trait by WAASY (joint mean + stability; higher = better). WAAS is the stability score (lower = more stable).") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

# Summary

* **G×E structure:** the AMMI ANOVA shows how much of the interaction the first two IPCA axes
  capture; a large PC1 share means a few environments/hybrids dominate the interaction.
* **Which-won-where:** the GGE polygon identifies whether the five environments form one or
  several mega-environments and names the winning hybrid in each — directly relevant to
  whether one hybrid can be recommended across sites or selection must be environment-specific.
* **Recommendation set:** the WAASY table lists hybrids that are simultaneously high-yielding
  and stable for GY and for PUE — the shortlist that, together with the GCA/BLUP results
  (notebook 4) and the indirect-selection traits (notebook 5), supports the breeding
  recommendation for low-phosphorus environments.
